### 2019-06-05 Latex/Figures/_compactness.tex
draft revisions

% # •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
% Compactness Example: District 6 of the Court Adopted Remedial Plan and District 7 of the 2011 Plan
% =================================================================
% -- FIGURE -- FIGURE -- FIGURE -- FIGURE -- FIGURE -- FIGURE --   % 
% ----------------------------------------------------------------- 
\begin{figure}
    \begin{center}
    \caption{Compactness Example: District 6 of the Court Adopted Remedial Plan and District 7 of the 2011 Plan}
    \label{fig:compact}
    \includegraphics[width=1\textwidth]{Figures/fig_compactness.pdf}
    \end{center}
    \tabnotes{District 7 of the Enacted plan is recognized as the ``Goofy kicking Donald Duck'', while District 6 of the Remedial map has the same geographic center.}
\end{figure}
% -----------------------------------------------------------------
% -- END FIGURE -- END FIGURE -- END FIGURE -- END FIGURE -- END FI %
% =================================================================

### 2019-06-05 Latex/Figures/_density_2016.tex
draft revisions

--- 2019-06-05 Latex/Figures/_compactness.tex
@@
 % # •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-% Compactness Example: District 6 of the Court Adopted Remedial Plan and District 7 of the 2011 Plan
+% Simulated 2016 Elections by Plan
 % =================================================================
@@
     \begin{center}
-    \caption{Compactness Example: District 6 of the Court Adopted Remedial Plan and District 7 of the 2011 Plan}
-    \label{fig:compact}
-    \includegraphics[width=1\textwidth]{Figures/fig_compactness.pdf}
+    \caption{Simulated 2016 Elections by Plan}
+    \label{fig:density2016}
+    \includegraphics[width=1\textwidth]{Figures/fig_sims_2016_maps_density.pdf}
     \end{center}
-    \tabnotes{District 7 of the Enacted plan is recognized as the ``Goofy kicking Donald Duck'', while District 6 of the Remedial map has the same geographic center.}
+    \tabnotes{Plots based on the 2016 Five Election Composite data centered at the 2016 vote share.
+All percentages in terms of Republican share of the two-party vote from the composite measure of five state-wide races in 2016.
+Shaded area contains one standard deviation on either side of the mean, representing 68\% of the simulated seat percentages.}
 \end{figure}

### 2019-06-05 Latex/Figures/_density_2018.tex
draft revisions

--- 2019-06-05 Latex/Figures/_density_2016.tex
@@
 % # •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-% Simulated 2016 Elections by Plan
+% Simulated 2018 Elections by Plan
 % =================================================================
@@
     \begin{center}
-    \caption{Simulated 2016 Elections by Plan}
-    \label{fig:density2016}
-    \includegraphics[width=1\textwidth]{Figures/fig_sims_2016_maps_density.pdf}
+    \caption{Simulated 2018 Elections by Plan}
+    \label{fig:density2018}
+    \includegraphics[width=1\textwidth]{Figures/fig_sims_2018_maps_density.pdf}
     \end{center}
-    \tabnotes{Plots based on the 2016 Five Election Composite data centered at the 2016 vote share.
+    \tabnotes{Plots based on the 2016 Five Election Composite data centered at the 2018 vote share.
 All percentages in terms of Republican share of the two-party vote from the composite measure of five state-wide races in 2016.

### 2019-06-05 Latex/Figures/_density_congress.tex
draft revisions

--- 2019-06-05 Latex/Figures/_density_2018.tex
@@
 % # •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-% Simulated 2018 Elections by Plan
+% Comparing the Simulated Elections with US Congressional Elections (PA)
 % =================================================================
@@
     \begin{center}
-    \caption{Simulated 2018 Elections by Plan}
-    \label{fig:density2018}
-    \includegraphics[width=1\textwidth]{Figures/fig_sims_2018_maps_density.pdf}
+    \caption{Comparing the Simulated Elections with US Congressional Elections (PA)}
+    \label{fig:densityplots_congressional}
+    \includegraphics[width=1\textwidth]{Figures/fig_sims_actual_density.pdf}
     \end{center}
-    \tabnotes{Plots based on the 2016 Five Election Composite data centered at the 2018 vote share.
+    \tabnotes{Plots based on the 2016 Five Election Composite data centered at the vote share for each year.
 All percentages in terms of Republican share of the two-party vote from the composite measure of five state-wide races in 2016.

### 2019-06-05 Latex/Tables/_tab_congsum.tex
draft revisions

% =================================================================
% -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE  % 
% -----------------------------------------------------------------
% US House Election Summaries (PA 2012-2016)
% ••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••• 
%
\begin{table}[!htbp] \centering 
  \caption{US House Election Summaries (PA 2012-2016)} 
  \label{tab:congsum} 
\begin{tabular}{@{\extracolsep{-5pt}} ccccc} 
 & 2012 & 2014 & 2016 & AVE \\ 
\hline \\[-1.8ex] 
Seats &  [13R-5D] &  [13R-5D] &  [13R-5D] & [13R-5D] \\ 
Seat \% & 72.2\% & 72.2\% & 72.2\% & $0.720$ \\ 
Votes & 51.1\% & 55.5\% & 54.2\% & $0.533$ \\ 
Bias & 0.13 & 0.1 & 0.11 & $0.113$ \\ 
Efficiency Gap & 0.21 & 0.11 & 0.17 & $0.163$ \\ 
Mean/Median & 0.06 & 0.06 & 0.06 & $0.060$ \\ 
Declination & 0.46 & 0.36 & 0.39 & $0.403$ \\ 
\end{tabular}
\tabnotes{Uncontested races are imputed with 0.25 and 0.75 for the respective winners. All votes are calculated from the Republican perspective of the two-party vote. We've adjusted all gerrymandering measures such that negative numbers indicate bias in favor of the Democrats.}
\end{table} 
%
% -----------------------------------------------------------------
% -- END TABLE -- END TABLE -- END TABLE -- END TABLE -- END TABL %
% =================================================================

### 2019-06-05 Latex/Tables/_tab_summaries.tex
draft revisions

% =================================================================
% -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE  % 
% -----------------------------------------------------------------
% County Splits and Compactness Scores of the Plans
% ••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••• 
%
\begin{table}[!htbp] \centering 
  \caption{County Splits and Compactness Scores of the Plans} 
  \label{tab:summaries} 
\begin{tabular}{@{\extracolsep{0pt}} cccc} 
  &  \textbf{County}  &  \textbf{Polsby}   &   \\
  \textbf{Plan}  &  \textbf{Splits} &  \textbf{Popper}  &  \textbf{Reock}  \\
\hline \\[-1.8ex] 
2011 Enacted & $41$ & $0.271$ & $0.164$ \\ 
Joint Legislative & $19$ & $0.365$ & $0.282$ \\ 
Gov. Wolf & $19$ & $0.393$ & $0.310$ \\ 
2018 Court Remedial & $17$ & $0.426$ & $0.324$ \\ 
\end{tabular}
\tabnotes{County splits include all the pieces in which a county is split, not just the total number of counties that have been split. (The latter number is the one most often reported in both court documents and in the media, but we regard the measure we report as both more precise and more informative.)}
\end{table} 
%
% -----------------------------------------------------------------
% -- END TABLE -- END TABLE -- END TABLE -- END TABLE -- END TABL %
% =================================================================

### 2019-06-05 Latex/_maintext.tex
draft revisions

% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
% -- SECTION -- SECTION -- SECTION -- SECTION -- SECTION -- SECTI %
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
\section{Introduction}
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
    \lettrine[lines=3, lraise=0.1, nindent=0em]{\color{lightgray}I}{n} \textit{League of Women Voters v. Pennsylvania}, decided January 22, 2018 (henceforth abbreviated \textit{LWV}), the Pennsylvania Supreme Court offered a new and distinctive test for unconstitutional partisan gerrymandering under which it found the state’s congressional districts to be a ``severe and durable" partisan gerrymander that violated a long-standing provision of the state’s constitution. The congressional map that was struck down was drawn in 2011 by a Republican-controlled legislature and signed into law by the then Republican governor. It was used in elections from 2012 to 2016. Prior to the \textit{LWV} decision, many academics had identified this map as a partisan gerrymander \citep[see e.g.,][]{Royden2017}, with some even identifying it as the most egregious congressional districting in the U.S. \citep[][ esp. Figure 3]{Wang2016_SLR}.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{See also \citet{Mcgann_et_al_2015_ELJ, McGann_et_al_2016_gerrymandering}}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
\par
    The Pennsylvania Supreme Court held that ``the fact that the 2011 Plan cannot, as a statistical matter, be a plan directed at complying with traditional redistricting requirements is \underline{sufficient} to establish that it violates the Free and Equal Elections Clause" [of the Pennsylvania State Constitution] (slip op. at p.128, emphasis added).
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{``Elections shall be free and equal; and no power, civil or military, shall at any time interfere to prevent the free exercise of the right of suffrage." Pa. Const. art. I, § 5.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
    By \textit{traditional} principles (also sometime called \textit{neutral} principles, or \textit{good government} principles), the Court meant drawing contiguous districts, satisfying the \textit{one person, one vote} standard, avoiding diluting the voting strength of protected racial or ethnic minority groups, minimizing unnecessary political sub-unit splits, and providing reasonably compact districts.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{Avoiding fragmentation of communities of interest is also sometimes included in this list.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
    However, the Court also considered a variety of other indicators involving the partisan aspects of the plan in coming to the conclusion that the map created a ``severe and durable" partisan gerrymander.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{The trial court considered a variety of factors offered in expert witness testimony, including evidence about the map‘s political impact. The trial court’s findings of fact reveal that the 2011 map exhibited clearly a multiplicity of indicia of partisan gerrymandering. In his review, the trial judge found, in essence, that the legislative process through which the plan had been chosen demonstrated the indicia of a partisan gerrymander, that the weird shapes of a number of the districts demonstrated the indicia of a partisan gerrymander, that the dismemberment of municipalities, townships and counties in the state demonstrated the indicia of a partisan gerrymander, that the cracking and packing of Democrat-leaning voters demonstrated the indicia of a partisan gerrymander, and that the frozen 13R-5D results over the course of three elections in a state that is highly politically competitive statewide demonstrated the indicia of a partisan gerrymander. And the trial judge also found that none of these features of the map could be explained as necessitated by the electoral geography and demography of the state, though he suggested the possibility that consideration such as incumbency protection might have mattered. However, the trial court then, rather regretfully, concluded that it did not have a legal basis to declare the plan unconstitutional under either the Pennsylvania State Constitution or under then existing federal law precedents. Instead, it expressed a pious hope that the U.S. Supreme Court, in its review of the appeals of lower court cases, such as the 2017 case of \href{https://www.brennancenter.org/sites/default/files/legal-work/16-1161_Opinion.pdf}{\textit{Gill v. Whitford 585 U. S. \_\_\_ (2018)}.}, which had overturned as a partisan gerrymander a Wisconsin legislative map, would finally adopt a (manageable) standard for identifying unconstitutional partisan gerrymandering that all courts could apply. The Pennsylvania Supreme Court, for all practical purposes, adopted in \textit{toto} the factual finding of the trial court magistrate, Judge Brobson, but came to a diametrically opposite legal conclusion about the constitutionality of the congressional plan and the legal grounding for that finding.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
\par
    Having invalidated the \textit{2011 Enacted} map, the Pennsylvania Supreme Court offered deferrence to the State to draw a remedy plan. The Pennsylvania legislature failed to enact a lawful remedy plan of its own due mostly to divided government that led to disagreement as to how to proceed between the newly elected Democratic governor and the still Republican-controlled legislature. Had such a plan been passed (and signed), the Court would have reviewed its constitutionality. In the absence of a legislatively enacted remedial plan, the Court invited submissions of plans from the parties to the lawsuit, as well the public at large, to submit remedy plans; and it provided guidance as to the features of the plan in which it was most interested. The Court Order in February 2018 inviting submissions of feasible maps mentioned only good government criteria: minimizing the number of county, township and municipality splits, and maximizing district compactness under several different measures of compactness. It made no mention of incumbent protection, nor voting rights act protections, nor of protecting communities of interests, nor did it mention political outcome projections.
\par
    While a remedy plan was offered by Republican leaders from both chambers (the \textit{Joint legislative} plan) and by the Democratic Governor (\textit{Gov. Wolf}), as well as by others, the Court rejected all such plans. Instead, the Court ordered into place for the 2018 election a map of its own drawn for it by a court-appointed consultant. In a split decision, the Court map was endorsed only by judges with Democratic affiliations, though some of the disagreement on the Court had less to do with the map than with the timing of its implementation.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{Pennsylvania Supreme Court Judges are elected or retained to ten-year terms in elections with party labels on the ballot. Judges with Democratic affiliations now constitute the majority of the Court In 2011 this was not the case.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
    The initial ruling that the 2011 plan was unconstitutional was sharply attacked by Republicans as judicial overreach, and the map the court majority adopted was rather viciously attacked by Republicans as simply a pro-Democratic gerrymander under the guise of a court-drawn map.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{On March 20, 2018, Rep. Cris Dush, of Jefferson County introduced a measure that would impeach the four Democratic members of the court that voted to replace the map. His argument is that the members of the court overstepped their rights and violated the Separation of Powers principle of the state. Others saw this as simply sour grapes: \\ \href{https://wapo.st/2LlEVvE}{``Pennsylvania Republicans lost the redistricting battle. Now, they’re declaring war on the courts", https://wapo.st/2LlEVvE (Christopher Ingraham, February 22, 2018).;} \\ \href{https://www.philly.com/philly/news/politics/will-republicans-impeach-pennsylvania-supreme-court-justices-20180222.html}{``Pa. Republicans are talking about impeaching state Supreme Court justices. Do they have an argument?", https://www.philly.com/philly/news/politics/will-republicans-impeach-pennsylvania-supreme-court-justices-20180222.html (Andrew Seidman \& Jonathan Lai, February 22, 2018).}}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
    However, the U.S. Supreme Court has declined reviewing either the state court opinion rejecting the 2011 plan or the state court’s chosen map -- presumably because these decision are based entirely on state law.
\par
    In contrast to the \textit{LWV} finding that the 2011 Pennsylvania congressional map was unconstitutional under state law, a 2017 challenge to the Pennsylvania congressional map under the Elections Clause of the U.S. Constitution was rejected by a three-judge federal court by a vote of 2-1. Because the \textit{LWV} decision mooted the federal litigation, the federal challenge was not appealed to the U.S. Supreme Court. However, a 2018 remand of a Wisconsin partisan gerrymandering case, \textit{Gill v. Whitford} has, to a limited extent, provided clues about the current [ca. May 2019] state of federal partisan gerrymandering jurisprudence than anything before it.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{The \textit{Gill} remand decision still leaves open the question that has haunted opponents of partisan gerrymandering for the last thirty plus years, namely `When, if ever, will the U.S. Supreme Court actually declare a redistricting map unconstitutional?'.  The potential for such a finding of unconstitutionality is laid out in \textit{Davis v. Bandemer, 478 U.S. 109 (1986)}, which declared partisan gerrymandering justiciable, but under which the Supreme Court has never yet (as of May 2019) struck down a plan as an unconstitutional partisan gerrymander. Despite the 7-2 decision on the Wisconsin remand, which rejected the view that the Court should simply bow out of the process of reviewing maps as possible unconstitutional partisan gerrymanders, the retirement of Justice Kennedy and his replacement by a more conservative Justice creates even greater uncertainty about whether, in the lifetime of the present authors, there will ever be a U.S. Supreme Court majority that rejects some challenged district as an unconstitutional partisan gerrymander. We will, relatively soon, have the answer to that question due to the inevitable appeals of partisan gerrymandering cases from North Carolina and Maryland that have already had oral argument on March 26, 2019 before the U.S. Supreme Court (\textit{Rucho, et. al}. v. \textit{Common Cause, et. al} [1:16-CV-1026]) and \textit{Lamone, et. al.} v. \textit{Benisek, et. al.} [1:13-cv-03233-JKB]), with decisions expected in June 2019. In addition, there is a partisan gerrymandering challenge in Michigan, \textit{League of Women Voters of Michigan} v. \textit{Johnson} that will certainly be appealed to the U S. Supreme Court, as will the next lower court decision in \textit{Gill} itself.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
    In that remand decision (\textit{Gill v. Whitford, 585 U.S. \_\_\_ (2018)}), the Supreme Court indicated that any standard for partisan gerrymandering must be a challenge to individual districts which demonstrates that the voters in the particular district have had their votes either ``packed" or ``cracked."
\par
    Districts that have been \textit{packed} are those which have had voters from one partisan leaning grouped together so as to waste votes of the group because they win the district by an overwhelming majority, thus preventing those  excess votes from being used more efficiently in other districts. Districts that have been \textit{cracked} are ones where voters of one partisan leaning have had their share of the district’s voting strength reduced so as to make it impossible (or unlikely) that members of this group will be able to carry the district. The standard for unconstitutional gerrymandering at the federal level -- if there is ever is one -- can thus be expected to be very different from that offered by the Pennsylvania Supreme Court, since \textit{LWV} relies on data about the features and effects of the plan as a whole, while the US Supreme Court appears to be insisting on a district-specific test of packing and cracking. 
\par
    We can identify five key features of the \textit{LWV} ruling: (a) though some districts might be particularly egregious, evidence of unconstitutional gerrymandering can be found in a plan as a whole; (b) evidence of purposeful discrimination was not necessary in order to find unconstitutionality; (c) evidence of discriminatory partisan effects was not necessary to show a violation; (d) evidence that a plan complied with good government criteria is almost certainly not sufficient to rule out a finding of unconstitutionality, even though a finding that a plan failed to comply with such criteria was sufficient to invalidate the plan; and (e) statistical evidence about the likelihood that  observed features of a plan could not be explained either by chance or by electoral geography played a pivotal role in the decision. 
\par
    Although only directly applicable in Pennsylvania, \textit{LWV} is, we believe, nonetheless the most important partisan gerrymandering case yet decided in a final and definitive fashion this decade. 
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{Indeed, if the Supreme Court continues to dodge any responsibility for checking egregious partisan gerrymandering, \textit{LWV} may well prove to be the most important partisan gerrymandering decision ever \citep{Grofman_Cervas_2018_ELJ}.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
    Because the US Constitution is the minimum rights provided to all citizens, states offer additional protections. Language virtually identical to the provision of the state constitution that the Pennsylvania Supreme Court relied on is found in the constitutions of another dozen states, and another thirteen have very similar language \citep{Douglas2014_RightToVote}.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{See also \citet{Elmendorf2018} and further discussion in the conclusions section of this paper.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
    That language could be readily adapted to control gerrymandering in those states in much the same way as did the Pennsylvania court. In fact, in its opinion, the Pennsylvania Supreme Court offers what we read as direct encouragement to state courts in other states to follow the lead of Pennsylvania in crafting a way to overturn partisan gerrymandering as a violation of a \underline{state constitution}.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{Footnote 71 of the Opinion (slip op. pp.116-117)}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
\par
    However, as we show later, some states in which possible state law challenges could be made will be insulated from those challenges by factors such as the nature of institutional control over the redistricting process in the state, or political gridlock in the state courts. Nonetheless, if a state court does follow the Pennsylvania Supreme Court lead, we view the expected effects on limiting partisan gerrymandering to be potentially greater than we might expect under federal law, since they would involve the overturning of whole plans, not just individual districts -- although, in practice, a finding of overall unconstitutionality is likely to lead to changes only in the most egregiously gerrymandered districts and some of the districts adjacent to them. Unlike all federal courts that have ruled on partisan gerrymandering to date, the Pennsylvania opinion does not require proof of intent to discriminate, which could make proving a violation easier under state law than under any likely federal standard. Moreover, although the Pennsylvania Supreme Court opinion does not require evidence of partisan effects, and emphasizes failure to comply with good government criteria, we believe that the \textit{LWV} opinion can be ``tweaked" to address what \citet{GrofmanCervas2018_WashPo} in their on-line essay in the \textit{Washington Post} refer to as \textit{stealth gerrymanders}, i.e., plans that satisfy good government redistricting criteria to a reasonable degree but are nonetheless carefully crafted invidious partisan gerrymanders in terms of their expected political impact. The reason for our optimism in this respect are the findings in the opinion that do address partisan effects of the Pennsylvania congressional map. 
\par
    In the remainder of this essay we will not seek to review in any more detail the various legal and expert witness issues raised in the Pennsylvania litigation itself.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{\citet{Grofman_Cervas_2018_ELJ} provide a review of the various expert witness testimony in the Pennsylvania case.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
    Rather, our primary aim is the straightforward one of comparing and contrasting the 2011 and 2018 Pennsylvania congressional maps in terms of a variety of proposed metrics for detecting partisan gerrymandering. In particular, we examine the evidence for the claim that the 2011 map was a pro-Republican gerrymander, and the claim that the 2018 map was a pro-Democratic gerrymander. The former claim is one virtually universally acknowledged \citep{Fang2014, Mcgann_et_al_2015_ELJ, McGann_et_al_2016_gerrymandering, Wang2016_SLR, Wang2016_ELJ}, while the latter claim has been offered not only as partisan griping by Republicans on the losing side of the litigation, but by at least one independent journalist.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{See NYT Upshot: \href{https://nyti.ms/2w8b19k}{``Democrats Didn’t Even Dream of This Pennsylvania Map. How Did It Happen?", https://nyti.ms/2w8b19k (Nate Cohn, Feb. 21, 2018).;} \\
        \href{https://nyti.ms/2wkpEFT}{``Hundreds of Simulated Maps Show How Well Democrats Fared in Pennsylvania", https://nyti.ms/2wkpEFT (Nate Cohn, Feb. 26, 2018).}}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
    However, the U.S. Supreme Court has declined reviewing either the state court opinion rejecting the 2011 plan or the state court’s chosen map -- presumably because these decision are based entirely on state law.
\par
    In addition to the rejected 2011 and the court-ordered 2018 map, we also address the partisan and other features of two maps offered to the Pennsylvania Court as a proposed remedy for constitutional violation found in the 2011 map: the \textit{Joint Plan} offered by Republican legislators from both chambers, and the congressional map proposed by the Democratic governor (\textit{Gov. Wolf}). Here we focus on the question of whether either of these plans was a \textit{stealth gerrymander} done to aid the party drawing the map.
\par
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
% -- SECTION -- SECTION -- SECTION -- SECTION -- SECTION -- SECTI %
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
\section{Evaluating the Various Plans -- Methods and Theory}
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
    As noted above, to contrast the Court-ordered Remedial plan with the 2011 Enacted plan, and to make further comparisons to the two most politically important remedy plans offered to the Court, the Joint Plan proposed by the Republicans leaders of the two legislative chambers, and the Governor’s proposed plan as a Democratic alternative, we will look at a number of features of each of these four plans as a whole.  We do not, however, discuss district-specific criteria for identifying gerrymandering in these maps, since, although some district-specific evidence was presented about the 2011 map in the trial testimony by an expert who was highly knowledgeable about communities of interest in the state, the Pennsylvania Supreme Court opinion itself offers a \underline{statewide} standard for a finding of unconstitutional partisan gerrymandering. 
   % ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{As noted earlier, given the language in the U.S. Supreme Court’s remand of the \textit{Gill} case, it appears that any federal standard for \underline{partisan gerrymandering} would need to be district specific.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% ================================================================= 
    All the plans create contiguous districts. Similarly, all four plans appear to satisfy voting rights requirements. Thus, we do not discuss these two ``good government” criteria further. Consequently, we limit ourselves to metrics that report results for a jurisdiction as whole. 
\par
    We begin our analyses describing map features distinct from projected (or actual) partisan outcomes or proposed measures of partisan effects. Table \ref{tab:summaries} offers a comparison of two key good government features of the map.
\begin{itemize}
    \item the number of county splits, since the Pennsylvania Supreme Court specifically sought to limit the number of counties spanning multiple districts
    \item compactness measures, since compactness is also a criterion specifically referenced by the Pennsylvania Supreme Court
\end{itemize}
\par
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
% -- SUBSECTION -- SUBSECTION -- SUBSECTION -- SUBSECTION -- SUBS %
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
\subsection*{Measurement and Data}
\subsubsection*{County Splits}
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
    Avoiding county splits is incorporated into the language of many state constitutions \citep{Grofman1985} and can be seen as a way of guarding against gerrymandering by making it more difficult to draw partisan maps. There are a total of 67 counties in PA, and the state is allocated 18 congressional districts.  Some county splits are necessitated by the size of the county population relative to the ideal population of a congressional district (705,687).  Allegheny and Montgomery must each be split at least twice, while Philadelphia must be split at least three times. If no set of counties exactly equals the ideal congressional population, every districting plan requires a minimum of $n-1$ splits (where $n$ equals the number of congressional districts). The Pennsylvania Supreme Court found that an unusually and egregiously large number of counties were split in the 2011 congressional map, and they found this fact deeply troubling.
\par
% =================================================================
% -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE  % 
% -----------------------------------------------------------------
% County Splits, Compactness Scores of the Plans
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
 \begin{center} INSERT TABLE \ref{tab:summaries} ABOUT HERE \end{center}
% -----------------------------------------------------------------
% -- END TABLE -- END TABLE -- END TABLE -- END TABLE -- END TABL %
% =================================================================
\par
    The data provided in Table \ref{tab:summaries} shows how the high number of county splits in the 2011 map was completely unnecessary. Both the Republican and Democrat proposed remedy plans cut the number of county splits from 41 to 19.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
       \footnote{There is a debate about how to count split counties, which was evident when different parties in the case counted different total county splits for the same plan. One way to count splits to count the total number of counties that are in at least two districts. The other way is to count the number of times a county is split, i.e., the number of separate pieces into which the county is divided. In the first counting method, a county can have portions of itself found in 18 different districts, but have this counted as only one county spit. Under the second method, this wholesale splitting of the county would be counted as 18 splits. While both methods provide useful information, we regard the second as more informative and it is the one reported in Table 1. However, we must also be alert for divisions of a county into two or more discontiguous pieces found within the same district. We call this “fracking,” to provide a term similar sounding to ‘packing’ and ‘cracking.’ We find “fracking” a particularly egregious problem since there is no need for counties to be fracked, and fracking may be used as a gerrymandering tool.  Instances where counties  have multiple discontiguous pieces located in a single district are not discussed in the Pennsylvania Supreme Court ruling on the unconstitutionality of the 2011 Pennsylvania congressional plan. However, our analysis of the 2011 plan reveals several cases of fracking: two of the most egregious are districts 14 and 18 in Allegheny County (Pittsburgh). This concept has also been addressed by the US District Court in North Carolina (\textit{Common Cause v. Rucho} No 1:16-CV-1026 (U.S. District Court, Middle District of North Carolina, 2018, slip op at p. 105 [p. 194])). }
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
    The Court plan goes even further and gets the number of split counties down to 17. 
\par
    Which counties get split and how many ways can impact a plan’s election outcome, which can affect both national politics and a county's relationship with the federal government \citep{Ansolabehere2002a}. That the 2011 map is an eyesore vis-a-vis unnecessary splits of county boundaries is revealed clearly in Figure \ref{fig:maps}.
%
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
    \footnote{We find counties that are split into multiple pieces particularly egregious. ``\textit{Fracking}" occurs when one district is split into two or more non-contiguous pieces of a single county. It was not mentioned in the \textit{LWV} decision, but too can be used as an indicia of partisan (and racial) manipulation. We call this ``\textit{fracking}" to parallel the terminology of other districting terms of ``\textit{cracking}" and ``\textit{packing}". Examples of ``fracking" in the 2011 enacted map include districts 14 \& 18 in Allegheny County (Pittsburgh).}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
\subsubsection*{Compactness}
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
    The measurement of compactness has long been a central issue in the districting literature \citep[see e.g.,][]{Reock1961, Niemi1990, Polsby1991, Webster2013, Kaufman2018}. Indeed, the original gerrymander was assumed to be evidence of political manipulation because of its irregular shape \citep{Martis2008}. Some minimal level of compactness can make it harder to gerrymander \citep[][cf. \citealt{Webster2013}]{Reock1961, Altman1998}. But compactness is not a magic wand that rules out the possibility of political manipulation. Moreover, as many authors have shown, compactness has multiple dimensions, and these different dimensions need not move in parallel with one another; thus a plan appears compact on one dimension might appear ill-compact on another. The two most common ideas of compactness refer, on the one hand, to how close a legislative district’s boundaries are to its geographic center and, on the other, to how ``regular" in shape a district appears to be \citep{Niemi1990, Kaufman2018}. In Table \ref{tab:summaries} we report two well-known measures which tap respectively each of these two dimensions.  The \textit{Polsby-Popper} measure looks at perimeter irregularity by examining the area of the district compared to that of a circle with similar perimeter, while the \textit{Reock} measure compares the area of a district with that of the district’s circumscribing circle \citep{Reock1961, Polsby1991}. 
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{$ A_D $ = area of district, $P_D$= perimeter of district, $Circle_D$= minimum circumscribing circle; \\
        \textit{POLSBYPOPPER} = $\frac{4 \pi A_D}{{P_{D}}^{2}}$ \quad
        \textit{REOCK} = $\frac{A_D}{A(Circle_D)}$}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
\par
    The 2011 Enacted map looks awful by both compactness measures compared to the Court-ordered map, or even compared to the two proposed remedial maps; while both the Republican and the Democratic proposed remedial maps look reasonable by both compactness measures, even though the Democratic plan is superior to the Republican plan and the court-ordered map is superior to both. However, until we examine the likely partisan effects of these four plans we are not in a position to conclusively rule out the possibility that one or more of the remedial plans is a ``\textit{stealth gerrymander}". As is now well recognized, gerrymandering can be found even in plans with compact appearing districts, or in a plan with districts that preserve most county borders.
\par
    Each of the three measures in Table \ref{tab:summaries} indicates a very dramatic deviation from good government standards in the 2011 Enacted map. And, each of three metrics order the plans in good government terms from worst to best in exactly the same way, with the worst plan being the 2011 Enacted map, the next worst being the Republican remedy plan, followed by the the Democratic remedy plan, and the very best being the court-ordered map. Moreover, as we will see later, the same ordering of plans is reflected when we examine the partisan implications of the various maps, with the map ordered into place as a remedy by the Pennsylvania Supreme Court close to perfectly neutral in its expected partisan effects once we recognize the implications of the partisan geography in Pennsylvania in ``naturally” tilting outcomes slightly in a pro-Republican direction because of differences in the degree of geographic concentration of each party’s supporters. (see below) Moreover, we show that the proposed Republican remedy plan can be characterized as a ``disguised” pro-Republican partisan gerrymander -- what we call a ``stealth gerrymander” -- which is nearly as pernicious in its expected pro-Republican partisan consequences as the 2011 plan whose effects it claims to ``remedy".
\par
    To give a more intuitive feel for how egregiously ill-compact the 2011 map is compared to the other alternatives we are considering, we show all four maps in the same scale, in a gray and white format. What is visually apparent from these maps is how aesthetically ``ugly" the 2011 map is compared to all of its alternatives. But it also clear that three remedy maps that are similar in having relatively few county cuts nonetheless can look quite different in terms of how each configures districts.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
    \footnote{This latter point is reinforced if we examine good government maps prepared by \href{https://www.dailykos.com/stories/2018/2/8/1739648/-Pennsylvania-will-soon-redraw-its-House-map-to-end-GOP-gerrymandering-How-would-you-draw-the-lines}{DailyKos} and by the present authors (available upon request).}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
%
% =================================================================
% -- FIGURE -- FIGURE -- FIGURE -- FIGURE -- FIGURE -- FIGURE --   % 
% ----------------------------------------------------------------- 
% Maps Of The Proposed Pennsylvania Congressional Districts
 \begin{center} INSERT FIGURE \ref{fig:maps} ABOUT HERE \end{center}
        \
% -----------------------------------------------------------------
% -- END FIGURE -- END FIGURE -- END FIGURE -- END FIGURE -- END FI %
% ================================================================= 
\par
    Now we turn to a consideration of probable partisan consequences of the four maps. We begin with a discussion of the 2011 map.
\par
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
\subsubsection*{Partisan results in actual elections held under the 2011 map}
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
%
    The best evidence about a plan’s partisan consequences is, of course, evidence derived from actual elections under the plan -- if there are any -- though we do need to check if there are highly anomalous aspects to one or more of these elections which would make results from them hard to generalize. Also, the existence of incumbency advantage can have a distorting effect on the translation of votes into seats, since incumbents may deter strong challengers, thus reducing the vote-share of the minority party \citep{Jacobson_Kernell_1981, Abramowitz1991}. Nonetheless, simply from examining outcomes in the three elections held under the plan (2012, 2014, 2016), the claim that the 2011 Pennsylvania congressional plan is an egregious partisan gerrymander is very hard to refute since (a) regardless of voter preferences the plan returned 13 Republicans and 5 Democrats, with no district changing hands over the course of the three elections, and (b) in those three elections, Democrats won only 28\% of the seats (5 out of 18), despite winning a substantial share of the vote in all three elections. In 2012, the Democratic share of the average two-party vote in the congressional elections in Pennsylvania was 48.0\%, in 2014 it was 44.5\%, and in 2016 it was 45.8\%. 
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
    \footnote{For consistency, we report the average vote share after imputing 0.25 and 0.75 for uncontested districts. In raw votes, Democrats won more votes than Republicans in 2012, but by taking the imputed average, it shows they won a minority of votes due to disproportionate uncontested districts. The Republican raw vote shares are 49.2\% in 2012, 55.5\% in 2014, and 54.1\% in 2016.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================    
    Thus, the discrepancy between Republican vote share and Republican seat share was huge, and yet the seat share appeared to be immutable.
\par
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
%        
\subsection*{Asymmetry measures for partisan outcomes in the 2011 Plan based on statewide election outcomes}
%
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
%
    There are other indicators of statewide partisan gerrymandering effects that also provide a compelling picture of the 2011 map as an (enduring) partisan gerrymander. There are several metrics that have been proposed to look at the degree to which there is asymmetry in seats-votes relationships that might be indicative of partisan gerrymandering: (1) \textit{partisan bias} \citep{Tufte1973}, (2) the \textit{mean-minus-median} gap \citep{Mcdonald_Best_2015_ELJ} (3) the \textit{efficiency gap} \citep{Stephanopoulos2014_UofChicagoLaw}, and most recently (4) declination \citep{Warrington2018}. Each of these measures can be directly calculated for the 2011 map based on district election outcomes. 
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
    \footnote{Another recently proposed measure is a simplified version of partisan bias which lacks the advantage of separately calculating bias and responsiveness (swing ratio). It is calculated by taking the average two-party vote for the dominant party and adjusting it uniformly downward in each district so that the two parties receive exactly 50\% of the overall vote; then we look to see how far the seat share the dominant party now receives is from 50\% \citep{Wang2016_SLR}.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
\par
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
\subsubsection*{Partisan Bias}
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
%
    \textit{Partisan Bias} indicates asymmetry in the translation of each party’s votes into seats. Customarily, partisan bias is measured at the (hypothetical) point where each party’s share of the two-party vote is exactly 50\%. At that point, both parties should, by symmetry, receive identical seats shares. Thus, the difference between a party’s projected seat-share with 50\% of the vote and exactly 50\% can be taken as a measure of partisan bias. Note, however, that although a plan must be proportional at the ($50,50$) point, a partisan bias of zero does not imply proportional representational. What partisan bias taps is differential treatment of the parties? Thus, for example, if party R gets 56\% of the seats with 52\% of the vote, this is not a sign of bias as long as party L also can be expected to get 56\% of the seats were it to have 52\% of the vote.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
    \footnote{Although the basic idea goes back at least as far as \citet{Dahl1956}, the modern \textit{locus classicus} in political science for the idea of partisan bias(and the parallel concept of responsiveness/swing ratio) is \citet{Tufte1973}, with important recent work done by Gary King (see e.g., \citep{GelmanKing1994_unifiedAJPS}; and review in \citet{GrofmanKing2007}). A closely related line of research is found in the political geography literature (with \textit{locus classicus} \citet{Brookes1959, Brookes1960} -- with some further methodological improvements by scholars such as \citet{Johnston1994, Rossiter1997} and \citet{Johnston2002}.}   
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
\par
    There are a number of different ways to estimate partisan bias based on the shape of the votes to seats distribution \citep{Tufte1973, Grofman1983, Browning_King_1987_seats_votes, GelmanKing1994_unifiedAJPS, Grofman_et_al_1997_SwingRatio_Bias, Zingher2016_bias_swingratio_JEPP}. The method we use here is basically that of \citet{Tufte1973} with some additional complexity building in ideas of \citet{GelmanKing1994_unifiedAJPS} on simulating potential outcomes in a probabilistic fashion. We adapted the code from \citet{Kastellec_et_al_2008_PS}, which parallels that used by \texttt{JudgeIt}. It models inter-election shifts and adjusts for differences over time in vote shares among and between districts in a probabilistic fashion, using a simulation with 1000 repetitions (see more below). We then apply the following procedure to find the mean partisan bias and range of expected bias over the course of the decade.
\par
    Hypothetical elections are constructed by incrementally adding (or subtracting) one percentage point (on average) at a time to find aggregate seat outcomes under differing mean vote shares. The resulting ($s$, $v$) points on the votes-to-seats curve are then converted to a log odds form by using $ log(\frac{s}{1-s}) $ as the dependent variable, and $ log(\frac{v}{1-v}) $ as the independent variable. A regression on the transformed variables is then run, with transformed $s$ as the dependent variable, for $v$ values in some range near a 50\% vote share. 
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------    
    \footnote{Where to center the \textit{seats/votes} curve is an unresolved issue in the literature. In \citet{GelmanKing1994_unifiedAJPS} \texttt{JudgeIt} program, the choice is to center the election at the average Democratic vote share over recent elections. \citet{Kastellec_et_al_2008_PS}, similarly, simulates data for average vote shares between 45\% and 55\%. We checked for consistency between the methods by running the numbers in three ways: centering the simulated outcomes at an average election vote share, centering the simulated outcomes at a tied 50/50 election, and centering the simulated outcomes at the actual election results in the particular election. All three variations provide similar estimates, as do the results from \texttt{JudgeIt}. Readers should note that with only 18 districts, too much weight should not be place exact bias coefficients reported [see the discussion in \citet{Browning_King_1987_seats_votes}]; but the relative magnitude of the bias values across the different plans is fully congruent with the evaluations of probable partisan gerrymandering yielded by the other measures. Because of the limited number of districts in the plans, the same caution is urged for all the measures of gerrymandering reported in this essay.} 
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
    \textit{Partisan Bias} is then calculated from the intercept of this regression using an exponential transformation (for details see \citet{Grofman1983}). Partisan bias is calculated from the Republican perspective, and \underline{negative} values indicates bias against Republicans.
\par
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
        \subsubsection*{Mean-Median Gap}
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
    When districts are sorted according to two-party vote share, the \textit{Mean-Median Gap} is the difference between the average vote percentage for a party and its vote share in the median district \citep[see][]{Best2018}. This measure is a variant of \textit{skewness}. When the mean is substantially higher or lower than the median, this is taken to be indicative of partisan bias.
\par
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
        \subsubsection*{Efficiency Gap}
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
    The \textit{Efficiency Gap} is calculated as defined in \citet{Stephanopoulos2014_UofChicagoLaw}, where all the party’s votes are wasted if they lose the district, and all the winner’s votes over 50\% are wasted. The difference between each party’s wasted votes is then divided by the total votes cast to produce the \textit{Efficiency Gap}, with a value of zero denoting what is regarded as ideal. As noted in \citet[][p. 13]{Best2018}, this is equivalent to taking an aggregate responsiveness (swing ratio) of two as ideal. 
\par
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
        \subsubsection*{Declination}
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
    Among the more recent measures of gerrymandering is the declination. Declination is an astronomical term referring to the angle on a compass, with a Latin root meaning ``a bending away". For gerrymandering, it uses the angles created from ordering all of the districts by vote share, computing the mean vote-share for each party separately for the seats they won, and then comparing the differences in their distribution. To compare, a line is draw from the mean vote-share to the 0.5 line separately for each party. The angle formed from the different slopes of the line is called the declination. An angle indicates whether a plan is a gerrymandering, under the logic that vote-shares should be distributed in some uniform manner across districts. If partisan are \textit{packed}, the angle will be greater. All values range between -1 and 1. Though we have defined the vote-share in terms of Republican outcomes, we adjusted the values such that \underline{negative} values are favorable to the Democrats. Further information can be found in \citet{Warrington2018}.  
\par
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
        \subsubsection*{Measuring Gerrymandering}
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
    We emphasize that a partisan gerrymandering claim based on political consequences should also provide evidence that the (a) gerrymandering indicators that are found cannot be explained simply by the geographic distribution of each party’s electoral support (e.g., differential concentration of electoral support can lead to what has been called a ``natural gerrymander"), and (b) that the partisan asymmetry in translating votes into seats is not due to chance alone (e.g., many highly competitive seats won by very narrow margins) but can be expected to be durable. With only 18 districts, sensitivity to chance effects is particularly important. 
\par
    Rebuttal to any claims that the partisan asymmetries observed in the consistent outcomes of the 2011 map can be attributed primarily to differential concentration of party voting strengths, or to chance, have been decisively refuted by the expert witness testimony in the case (see esp. the simulations of potential good government maps done by Professor Chen, discussed in detail in the Pennsylvania Supreme Court opinion in the case, and on which the Court placed great reliance). With respect to the durability of the effects of the 2011 map, we have already pointed out that this plan yielded unchanged 13R-5D results for three elections in a row. To help explain this over time consistency we note that the 2011 Map has primarily safe seats (13 of 18 won with greater than a 10\% margin). But, indicative of the pro-Republican gerrymander achieved with the 2011 map, those safe seats are distributed in a highly asymmetric fashion, with nine essentially safe seats for Republicans and only four for Democrats.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{However, we should note that some authors \citep[e.g.,][]{Grofman2019_ELJ} have argued that a partisan gerrymandering claim at a jurisdiction-wide level must also be reinforced by district specific evidence of manipulation of particular district boundaries. As noted earlier, we do not discuss this type of district specific evidence here, other than to point out that such evidence was presented at the trial, and that this evidence was used to support the Court’s finding that the Pennsylvania congressional map was an unconstitutional gerrymander under Pennsylvania state law.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
\par
    We show in Table \ref{tab:congsum} the results for the 2011 plan found unconstitutional. Results of the four measures for the 2011 plan are based on using the average of actual congressional results for the 2012-16 period under that plan. \underline{Positive} values favor Republicans (add to their seat share), while \underline{negative} values favor Democrats (subtract from Republican seat share).   
\par
    In our view, none of these four measures standing alone can be taken to be proof of gerrymandering. Rather we see them as potentially reinforcing indicators. Our own analyses have found hypotheticals where these measures can generate either a false positive or a false negative, but we have also found that consistent bias in the same direction is unlikely when we use multiple indicators (data omitted for space reasons). Here, however, all four indicators, like the actual election results, point clearly and overwhelmingly in the same direction vis-à-vis the 2011 Enacted plan -- namely that it is a blatant pro-Republican partisan gerrymander.
%
% =================================================================
\subsection*{Evaluating Partisan Gerrymandering using Statewide elections to provide a normal vote baseline}
% =================================================================
%
    In order to assess expected partisan consequences for plans that have not yet been implemented we need to develop a predictive equation, and that is most readily done with data on partisan outcomes for state-wide elections projected into the new districts. For example, Professor Jowei Chen in his testimony in the Pennsylvania case used a composite of the six state-wide elections between 2008 and 2010 to estimate the political consequences of alternative maps. After extensive investigation of alternative prediction models and their accuracy in reproducing the partisan outcomes of the 2012, 2014, and 2016 congressional elections at the district level (data omitted for space reasons), we have opted for a model that takes the average of five state-wide 2016 elections that reflect a balance of Republican and Democratic success: President, U.S. Senate, PA Attorney General, PA Treasurer, and PA Auditor. Republicans won the statewide vote in two of them (President, U.S. Senate), while the Democrats won the statewide vote in three of them (Auditor, Treasurer, and Attorney General). The average vote for the Republicans across these five elections was 49.1\%.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{We compiled the composite measure using both the sum of the raw votes by party and an average of the two-party vote in the five races at the district level. Both ways result in the same district partisanship breakdowns in all plans.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
\par
    We use these five elections to calculate the same four previously introduced metrics but now for all four plans.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{We report the Republican share of the two-party vote ($\delta_{i}$). Districting plans are represented by $\mathcal D $, (e.g., $\mathcal D_{enacted}$, $\mathcal D_{remedial}$, \dots, $\mathcal D_{j}$), and each election in year $ y $ has a district vote distribution $ \Delta_{y} = [\delta_{y1}, \delta_{y2}, \dots, \delta_{yi}] $. To find the overall state-wide vote, we calculate the average district vote share, $ \bar{\Delta}_{\mathcal D y} = \frac{1}{n}\sum\limits_{i=1}^{n} \delta_{yi} $. By averaging, we reduce the influence of turnout variation among districts. This district-specific average is a useful state-wide estimate of voter sentiment though it is in part effected by the particular configuration of districts being examined \citep[see e.g.][cf. \cite{Grofman_et_al_1997__IntedgratedPerspective_ES}]{Kastellec_et_al_2008_PS}.In particular, incumbency advantage effects affect our ability to accurately predict vote outcomes. In ancillary work not reported here, we have looked at changes in congressional incumbency advantage in Pennsylvania over multiple decades.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
    We aggregate voting district level (precincts) data to the congressional district level.
%        ================================================================= 
%        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% ------------------------------------------------------------------------
		\footnote{We wrote a function in \texttt{R} to assign all VTDs to a district based on the centroid of the VTD. This method creates a slight difference between it and the actual vote, but averages just -0.06\% per district. Most of the mis-allocation is between two adjacent districts, District 1 and District 13, both overwhelmingly Democratic districts.}
% ------------------------------------------------------------------------ 
%        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
%        =================================================================
% =================================================================
% -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE  % 
% -----------------------------------------------------------------
\par
    \begin{center}    \textbf{US House Election Summaries (PA 2004-2016)} \par \end{center}
    \begin{center}   \textbf{[ INSERT TABLE \ref{tab:congsum} ABOUT HERE ]} \end{center}
% ••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••• 
\par
    We do need to check that our five-election composite offers a plausible basis for a prediction model once we take into account electoral tides. 
%        ================================================================= 
%        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% ------------------------------------------------------------------------
		\footnote{We can only validate our measure for the 2016 election and only for the 2011 enacted plan. For all the remedial plans we are analyzing, the composite is our ``best guess". We have to make some assumptions about state and district inter-election swings. For simplicity, we assume that all districts have a national shift equal to the state-level swing, ie, difference from the 2016 elections. Later in the paper, we conduct simulated elections that vary the district level variability.}
% ------------------------------------------------------------------------ 
%        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
%        =================================================================
    Encouragingly, our five-election composite results is the same 13R-5D result that the 2016 congressional races delivered.  
%        ================================================================= 
%        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% ------------------------------------------------------------------------
		\footnote{Whether we take the sum of the five elections or average the five elections, both methods deliver the same relative distributions.}
% ------------------------------------------------------------------------ 
%        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
%        ================================================================= 
    Additionally, each district's outcome is identical to the 2016 congressional district results. 
\par
    But we do wish to do a further robustness check on the plausibility of using our five election composite to calculate the various metrics designed to measure potential gerrymandering. One way to do this is to see how well that composite (calculated at a hypothetical 50\% vote share) mimics the results of the values for the four metrics for the 2011 map that we previously obtained using the actual congressional election results for that map. What we are doing is comparing a hypothetical ``normal vote” \citep{Converse1966} assumption (based on a statewide vote in which we are not taking incumbency effects into account) with the average of what we actually get in three different congressional elections. The results in Table \ref{tab:congsum} are 0.11 for bias, 0.16 for the efficiency gap, 0.06 for the value of the mean minus median gap, and 0.40 for declination. The comparable results from Table \ref{tab:gerry} are 0.09, 0.25, 0.05, and 0.40. It is apparent how close these two sets of values are to one other even though one is based on actual elections with a mean Democratic (two-party) vote share close to 47\% while the other is based on projected election results set to yield a 50-50 two-part vote distribution. If instead we calculate these four metrics using the average five election statewide vote in these elections over the period 2012-2016, a Democratic vote share of 49.1\% we get virtually identical values: 0.09, 0.24,  0.05, 0.36.
%        ================================================================= 
%        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% ------------------------------------------------------------------------
		\footnote{If we center the simulations at the 2012-2016 average statewide Republican vote of 53.3\% of the vote, we again get remarkably consistent estimates [0.09, 0.19, 0.05, 0.45].}
% ------------------------------------------------------------------------ 
%        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
%        =================================================================    
\par
    Our confidence in our five election projection method is further buttressed because, when we applied this method to predict the results of the 2018 election under the court-ordered remedial map, although the method had been developed prior to the 2018 election, using data no more recent than 2016, it correctly predicted that the court-ordered remedial plan would have (in this good Democratic year) an outcome of roughly 9R-9D.
% =================================================================
% -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE  % 
% -----------------------------------------------------------------
% Measures of Gerrymandering for the Four Considered Plans
% ••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••• 
%    
\begin{center} INSERT TABLE \ref{tab:gerry} ABOUT HERE \end{center}

% =================================================================
% -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE  % 
% -----------------------------------------------------------------
% Probabilistic Projections of Partisan Outcomes for Four Plans
% ••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••• 
%
\begin{center} INSERT TABLE \ref{tab:prob} ABOUT HERE \end{center}
\par
    The findings from Table \ref{tab:gerry} are very clear! Regardless of which metric we examine, the 2011 map is the most gerrymandered. Not surprisingly, the nature of the bias is in a pro-Republican direction. Given the cumulative weight of all the evidence, the 2011 congressional map was clearly an egregious pro-Republican gerrymander. In contrast, as we might expect, the court-ordered plan, prepared by a respected expert whose instructions were to satisfy good government criteria, was by a considerable margin the closest to a perfectly neutral plan in partisan terms according to all four measures. 
\par
    Finally, we turn to the plan offered by the Democratic governor, Governor Wolf. To our surprise, this plan turned out also to have a pro-Republican tilt, nearly to the same extent as the Joint Legislative plan for three of the four metrics in Table \ref{tab:gerry}. This puzzling finding suggests that the Governor might have been under pressure from Democratic incumbents not to reduce their expected vote margins, or that he sincerely intended it to be a compromise that the Republican legislature might agree to. However, there is one important difference between the Joint plan and that of the Governor: among the seven districts likely to be won by Democrats, the Democratic percentages are cut more narrowly in the Republican plan than in the Governor’s plan in three of the districts so that, in a good Republican year, Republicans will do better under their plan than under the Democratic plan.
\par
    Still, when we look at Table \ref{tab:gerry} we are struck by the fact that all four plans, even the Court-ordered one, still appear to have some level -- very slight in one case, extreme in others -- of pro-Republican bias. In the next section we consider why this might have happened, and we introduce a potentially important distinction between ``\textit{neutral plans}" and ``\textit{fair plans}."
\par
    In sum, looking at the evidence above, we see exactly the same ordering of plans according to their characterizability as a partisan gerrymander as we saw when we ordered plans according to the degree to which they satisfied good government criteria (see Table \ref{tab:summaries}), with the 2011 map the worst and the 2018 court-ordered map far and away the best. This conclusion is further buttressed by the data we present in Table \ref{tab:prob}. Here we create a probabilistic simulation using statewide five-election composite (set to a 50-50 vote share) to estimate the likelihood that random shocks (based on past state inter-election shift data) would change party control of the congressional delegation. 
\par
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
    \subsection*{Natural Gerrymanders}
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
    In 2018, even under a court-drawn plan, in a year when Democrats had won a clear majority of the aggregate two-party congressional vote, Democrats only succeeded in winning exactly 50\% of the seats. Trying to understand why all four of the plans we considered had some level of pro-Republican bias requires us to look in more detail at the partisan electoral geography of Pennsylvania. In Pennsylvania, Philadelphia is overwhelmingly Democratic in voting. In particular, if you draw two congressional districts entirely within Philadelphia County, one of them is very likely to give Democratic candidates around 90\% of the vote, and for sure, as long as both are wholly within the County, the average vote in the two will be around 80\% Democratic no matter how you draw the two districts. There are no other large, concentrated pockets of equivalently overwhelmingly Republican voting strength. Thus, more Democratic votes will \textit{naturally} be ``wasted" in Philadelphia than Republican votes will be ``wasted" elsewhere in the state. Additionally, the city of Philadelphia is landlocked, and its position on the state's border reduces the ability of line drawers who favor Democrats to distribute these voters into multiple districts so as to minimize the extent to which Democratic votes in the county are wasted. Additional Democratic wasted votes come in heavily Democratic Allegheny County, the county in which the city of Pittsburgh is located. However, saying that urban concentration of Democrats means that it is harder for Democrats to turn their votes into seats than is the case for Republicans does not mean that we cannot distinguish the effects of such \textit{natural} gerrymanders from \textit{intentional} political gerrymandering, since the latter generates perverse consequences for Democratic voters to a far greater extent than the likely one district or so penalty created by the nature of Pennsylvania’s present electoral geography. 
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
    \footnote{\citet{Chen2013} estimate this bias as about 1.45 seats (8\%) but Chen’s expert witness testimony in the Pennsylvania cases generates a lower estimate of natural bias until one builds in the need to protect minority opportunity districts. Here, we need to be careful since the minority population in such districts in the enacted map may be larger than actually needed to assure effective minority representation, and the act of  maintaining or even increasing high levels of minority population in the district may be a disguised way of packing Democratic voters so as to advantage the interests of the Republican party.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
    For example, in the first set of 500 simulations in his expert witness report, based only on traditional redistricting criteria, but not including voting rights protections, Professor Chen found that none resulted in the 13R-5D split as happened in 2012 and subsequent years. The mode was 9R-9D.  
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
    \footnote{Expert Witness Report, Jowei Chen (Pg. 15, 16, Figure2).}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
\par
    Recognizing the potential for so-called ``natural gerrymandering", we wish to highlight a potentially important distinction between \textit{neutral} plans and \textit{fair} plans -- each of which reflects a different approach to defining the baseline for what constitutes gerrymandering. Neutral plans refer to those that are drawn entirely with respect to traditional good government criteria, with no attention paid to partisan considerations. One way to define partisan gerrymandering is with respect to a baseline defined by the set of feasible neutral plans -- and this was how one of the plaintiff’s experts in the Pennsylvania case, Professor Jowei Chen \citep[cf.][]{Chen2013} shaped his testimony in the Pennsylvania case, and in other cases in which he has testified.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
    \footnote{Similarly, \citet{Grofman2018_ELJ} asserts that ``Gerrymandering occurs when a districting plan creates a disparate treatment of the vote share of the minority and majority voting blocs in a way that penalizes the minority in its ability to translate its voting support into seats compared to what we might expect from a plan drawn on the basis of neutral principles."}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
\par
    In contrast, we may wish to use the term ``fair" plans as ones that seek to compensate for the difference in level of partisan concentration. Fair plans do not seek to achieve proportional representation. Rather, they seek to achieve partisan symmetry. A fair plan would yield a proportional result when the vote shares are evenly split 50-50, but at points beyond 50-50, such as when party R wins 70\% of the districts with only 55\% of the vote, party L can similarly win 70\% of the districts if it had received 55\% of the vote. 
\par
    One clear finding of our analysis of the court-ordered map is that it was drawn as a ``neutral map", though not necessarily a fair map in the sense we are using the term above. In order to create a truly ``fair map", in our view, it would have been necessary to break up Philadelphia County in more than three pieces so as to avoid the ``natural" packing of Democratic voters.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
    \footnote{Note that we certainly not advocating that such a three way split of Philadelphia should have been done.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================    
    But the Pennsylvania Supreme Court opted not to do that, but to instead preserve county lines and draw a good government map. We can therefore identify the Court remedial plan as a ``neutral" plan.
  \par
  \par

%
% =================================================================
% -- SECTION -- SECTION -- SECTION -- SECTION -- SECTION -- SECTI %
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
\section{Conclusions and Lessons for the future; The Potential Impact of LWV}
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
    While the Pennsylvania court opinion is limited to Pennsylvania, and thus it might have seemed of importance only in Pennsylvania, in footnote 71 of the Opinion (slip op. pp.116-117), the court took what we regard as a rather unusual step. It issued what can only be called an invitation to other state courts to use the same logic it used to invalidate partisan gerrymanders in their own state. As we noted earlier, the Court pointed out that there are twelve states whose constitutions contain election clauses identical to the Pennsylvania charter, requiring elections to be “free and equal”: Arizona, Arkansas, Delaware, Illinois, Indiana, Kentucky, Oklahoma, South Dakota, Oregon, Tennessee, Washington, and Wyoming.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{In its Opinion, the Court provided specific citations to each of these provisions which we have not reproduced.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
    However, only a handful of these states are ripe for partisan gerrymandering challenges -- in that (ca. 2017) three lack unified party control of the state, two are single district states, and some already have a commission drawing plans and, in others, \textit{indicia} of gerrymandering are missing.
\par
    In looking to the future, we should also note that some states that are generally regarded as among the most pernicious partisan gerrymanders, Michigan, Ohio, North Carolina, Wisconsin are not included among the twelve, nor is Maryland. On the other hand, we should also note that a “free and equal” elections clause is not the only avenue state courts might use to attack partisan gerrymanders in the future. As University of Kentucky College of Law Professor Joshua Douglas has pointed out, virtually every state constitution protects voting rights more explicitly than the U.S. Constitution does \citep{Douglas2014_RightToVote}. In addition to the thirteen states that require elections to be ``free and equal”, an additional thirteen have state constitutional provisions that require elections to be ``free" or ``free and open”, and this clause could, in principle, be used in exactly the same way as the ``free and equal” clause. 
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{We are indebted to Jonathan Lai of the Philadelphia Inquirer (personal communication, April 2018) for calling this information to our attention. See a more detailed discussion in \citet{Elmendorf2018}.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
    Such a clause is found in a number of the states widely regarded as having the worst congressional gerrymandering. 
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{These states include Arizona, Arkansas, Colorado, Delaware, Illinois, Indiana, Kentucky, Maryland, Massachusetts, Missouri, Montana, Nebraska, New Hampshire, New Mexico, North Carolina, Oklahoma, Oregon, Pennsylvania, South Carolina, South Dakota, Tennessee, Utah, Vermont, Virginia, Washington, and Wyoming \citep[][footnote 86]{Douglas2014_RightToVote}.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
\par
    Accordingly, some scholars have suggested turning to turn away from the federal judiciary altogether and instead focusing on state-level litigation \citep{Wang_et_al_2019_Labortories_UPJCL}. We believe that the \textit{LWV} decision and the further comparative analyses of alternative plans presented in the previous section of this paper will help other state courts navigate their way to decisions that strike down some partisan gerrymanders as unconstitutional, while allowing others to remain in place on the grounds that either they are not that severe, or they are unlikely to be lasting in that they have sufficiently many competitive seats as to be responsive to realistic changes in voting patterns. Now that litigation has been successful in state courts on the grounds of egregious deviations from good government districting criteria, it seems plausible that in the next wave of redistricting, mapmakers may create covert gerrymanders like we've described in this essay. If so, there remains the issue of whether state courts have the legal tools to deal with these ``stealth gerrymander”.
\par
    According to all four measures, the Joint Legislative plan offered by Republican leaders is nearly as much a gerrymandered map -- in a pro-Republican direction -- as is the 2011 map. As noted above, we believe it appropriate to label this plan as a stealth gerrymander since it achieves partisan purposes while (mostly) providing the visual appearance of a good government map (see Figure \ref{fig:maps}). However, having labeled the Joint Legislative plan a \textit{stealth gerrymander}, it is worth reminding readers of two important points. 
\par
    First, had this plan been the original plan adopted by Republicans, it is very much an open question as to whether it would have been rejected by the Pennsylvania Supreme Court. Even though, in partisan terms, it is nearly identical to the 2011 plan, it is considerably more consistent with good government criteria (see Table \ref{tab:summaries}) -- not that much worse than the Court Remedial plan -- that it would have required detailed political election analyses like those we have done here to demonstrate its stealth gerrymander features, plus a court willing to rule on political grounds rather than good government grounds that it was unconstitutional. Because the Joint Legislative plan was not passed by the State of Pennsylvania legislature, the Court felt no need for particular deference to it (or to the Governor's plan, for that matter). Thus, the Court was not bound by the normal supposition with respect to districting that a plan authorized by the state need not be the ``best possible," but only unconstitutional. That ``deference" to legislative judgments and the criteria they reflect, might well have tipped the balance toward acceptance of the Joint Legislative plan were it offered as having been fully sanctioned by the State of Pennsylvania (the duly elected legislature and Governor). 
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
    \footnote{The Court picked the plan among those before it that most closely satisfied good government criteria -- which happily, thanks to Professor Persily's expertise as consultant to the Court, turned out to be its own plan, and thus a plan which the Court could know with certainty was not intended as a partisan gerrymander for either party. Whether it would have still sought to maximize compliance with good government criteria had the plan which did so been a ``stealth gerrymander" is not a question we can answer.}       
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
    On the other hand, in the presence of this \textit{stealth gerrymander} rather than the blatant gerrymander that came before it, the Pennsylvania Court might well have relied on partisan impact evidence to reach a conclusion of unconstitutionality. The language of the Final Order implementing the Court's own plan strongly suggests this possibility. There, the Court said about the 2011 plan that it 
    % =================================================================
% ------- QUOTE ---------------------------------------------------
    \begin{quote}
        ``was designed to dilute the votes of those who in prior elections voted for the party not in power in order to give the party in power a lasting electoral advantage. In stark contrast, Article I, Section 5 of the Pennsylvania Constitution provides: ‘Elections shall be free and equal; and no power, civil or military, shall at any time interfere to prevent the free exercise of the right of suffrage.’ Pa. Const. art. I, § 5. On this record, it is clear that the 2011 Plan violates Article I, Section 5, since a diluted vote is not an equal vote" (slip op. p. 2).
    \end{quote}
% ------- END QUOTE -----------------------------------------------
% =================================================================
\par
	Thus we have some optimism that state courts can provide a venue to check partisan gerrymandering, even when it comes in the form of a \textit{stealth gerrymander}. But we have also placed a number of caveats on any claim that state courts can compensate fully for any continued failure of the federal courts to act. Citizen-based reforms such as referendum and initiatives may also help with reducing egregious gerrymandering in states where the passage of such measures is feasible. Ultimately, there is no substitute for a federal standard that guarantees each citizen in the United States \textit{equal protection under the law}. Absent the United States Supreme Court setting this standard, state courts offer one promising pathway towards ending undemocratic partisan gerrymandering.

% 
% 
% 
% 
%

% % ================================================================= 
% % -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% % -----------------------------------------------------------------
%         \footnote{Because of its high number of competitive seats, a point prediction such as 9R-9D is somewhat misleading for the court-ordered map. Using the five-election model as our predictor, there are fewer safe Democratic seats (5) than safe Republican seats (7) in the Court plan, using 45\%-55\% definition of a competitive seat. This is one fewer safe seat for Republicans and one additional safe Democratic seat than is found in the 2011 plan, but the Court plan also has five districts balanced on a knife edge, with a projected winner’s margin under two percentage points, making it highly responsive to shifting voter sentiment. This is realized in our simulations where the mean seat share, though near 9R-9D, is either lower or higher depending on where the state-wide vote is centered.}
% % ----------------------------------------------------------------- 
% % -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% % =================================================================

% 	Past congressional results alone often are departures from the underlying district partisanship several reasons. The vote share of an incumbent has historically been greater than it is for a non-incumbent or two non-incumbents running for an open seat.
% %        ================================================================= 
% %        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% % ------------------------------------------------------------------------
% 		\footnote{A long literature in political science has evaluated and confirmed the presence and increased incumbency advantage \citep{Abramowitz1991,Levitt1997,Ansolabehere2002b,Gelman_Huang_2008}. Recent scholarship, however, has shown a decrease in the advantage towards zero \citep{Jacobson2015_JOP}.}
% % ------------------------------------------------------------------------ 
% %        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% %        =================================================================
%     A well-financed incumbent tends to suppress quality competition \citep{Jacobson2015_JOP}, vulnerable incumbents who can sense the political winds opt against running for re-election \citep{Jacobson_Kernell_1981,Ansolabehere_Snyder_2004_LSQ}, and electoral tides and varying turnout can distort the state-wide average. Midterm elections often have vastly different electorates than presidential elections \citep{Abramowitz_et_al_1986_AJPS}, while the midterm pattern of ``surge and decline" (what goes up must come down principle) reduces the sitting Presidents' party's vote share \citep{CampbellA1960}. 
%  \par
% 	What we are after is a measure of the ``normal vote''. The normal vote is the underlying partisanship of a district \citep{Converse1966}. It is sort of a mystical number since it is uncovered and unobserved, but powerful theoretically. Map builders implicitly assume a normal vote when constructing a district plan. We believe that state-wide elections offer the most useful approximation of the normal vote.
% %        ================================================================= 
% %        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% % ------------------------------------------------------------------------
% 		\footnote{We particularly believe this is true when the votes between elections are highly correlated, and the district-level variation is low, as it was in Pennsylvania from 2012 to 2016. The more stable the electorate is, the more confidence we have that state-wide elections are revealing the normal vote \citep{Ansolabehere_et_al_2000_AJPS}.}
% % ------------------------------------------------------------------------ 
% %        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% %        =================================================================
% 	 All districts vote on the same candidates; hence outcome affects all voters equally. The effects of incumbency can distort the vote in congressional races in a way that doesn't resemble the actual partisan preferences at the district-level. State-wide elections eliminate the need to impute or otherwise deal with uncontested districts. Taking an average of the state-wide elections provides additional reason to believe it's capturing a normal vote due to a regression to the mean effects. Justices in \textit{Vieth} offered skepticism in using elections that have contrasting results; we do not share this view, instead seeing it as an important source of variation that helps to recognize the underlying partisanship in which \textit{should} be used in measuring bias in districting plans. Additionally, averaging over multiple races and years helps to reduce the bias imposed by candidate effects.
% \par
% 	We construct the five-election composite of state-wide elections by aggregating voting district level (precincts) data to the congressional district level.
% %        ================================================================= 
% %        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% % ------------------------------------------------------------------------
% 		\footnote{We wrote a function in R to assign all VTDs to a district based on the centroid of the VTD. This method creates a slight difference between it and the actual vote, but averages just -0.06\% per district. Most of the mis-allocation is between two adjacent districts, District 1 and District 13, both overwhelmingly Democratic districts.}
% % ------------------------------------------------------------------------ 
% %        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% %        =================================================================
% 	Encouragingly, our five-election composite results in the same 13R-5D result that the 2016 congressional races delivered.
% %        ================================================================= 
% %        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% % ------------------------------------------------------------------------
% 		\footnote{Whether we take the sum of the five elections or average the five elections, both methods deliver the same relative distributions.}
% % ------------------------------------------------------------------------ 
% %        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% %        =================================================================
% 	Additionally, each district's outcome is identical to the 2016 congressional district results. Indeed, after accounting for incumbency, our composite measure is remarkably accurate in its predictions of the actual congressional results validating our measure. The average difference between our composite and the congressional results was just 2.16\%. 
% %        ================================================================= 
% %        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% % ------------------------------------------------------------------------	
% 	We calculate this average only for districts contested in 2016, since uncontested results have significant deviations. Including the uncontested districts increases this mean difference to just 3.93\%. We believe our composite measure offers better representation of the true partisanship in these districts. Individually, only the Auditor's race correctly predicts all the districts correctly, but by either summing or by taking the average of the five races, both methods correctly predict each district's outcome.
% % ------------------------------------------------------------------------ 
% %        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% %        =================================================================
%     We imputed the values 25\% and 75\% for uncontested or essentially uncontested races. Rather than taking the projected outcome we arrive at from this method as a single value, we looked at the sensitivity of this projection to the empirical range of inter-election swing, simulating outcomes 1,000 times. To arrive at our estimate for inter-election variation that can't be attributed to the normal vote or incumbency advantage, we estimate the equation using congressional elections in Pennsylvania in the 2004-16 time period:
%         \begin{align*}
%         \mathrm{Vote}_{ty} = \alpha + \beta_{1} \mathrm{Vote}_{yt-1} + \beta_{2} \mathrm{Incumbency}_{yt} + \beta_{4} \mathrm{Uncontestedness}_y{t} + \epsilon
%         \end{align*}
%     We are interested in $\epsilon$, which is the unexplained variance, which includes inter-election swing. We run this equation separately for each of the non-redistricting years and then average the coefficients. We estimate incumbency advantage to average 4.27\% for this time period, while inter-election variance is 4.46\%. When simulating elections, we use the composite as the normal vote, but add in for each district a random draw from a normal distribution centered at the district vote with a standard deviation of $\epsilon$. We can then calculate statistics for each of these simulated elections to find a distribution for each measure of gerrymandering.
% \par
%     Table \ref{tab:gerry} reports the same four measures of partisan effects found in Table \ref{tab:congsum} but now for all four maps, using the projected statewide vote from our five-election composite, under the methodology described above. We also provide a measure of variability to show what range of reasonable outcomes for each of the plans and identify when gerrymandering effects are persistent.

% \par
%     As we might also expect, again according to all four measures, the next most gerrymandered map -- also in a pro-Republican direction -- is the \textit{Joint Legislative} plan offered by Republican legislative leaders as a remedy plan. The partisan gerrymandering indicators for this plan are sufficiently close to those of the 2011 map, and sufficiently far from those of the court-ordered map, that we believe it appropriate to label this plan as a ``\textit{stealth gerrymander}" since it achieves partisan purposes while (mostly) providing the visual appearance of a good government map (see \ref{fig:maps}). 
% \par
%     However, having labeled the Joint Legislative plan a \textit{stealth gerrymander}, it is worth reminding readers that, had this plan been the original plan adopted by Republicans, it is very much an open question as to whether it would have been rejected by the Pennsylvania Supreme Court. Even though, in partisan terms, it is nearly identical to the 2011 plan, it is considerably more consistent with good government criteria (see Table \ref{tab:summaries}) -- not that much worse than the Court Remedial plan -- that it would have required detailed political election analyses like those we have done here to demonstrate its \textit{stealth gerrymander} features, plus a court willing to rule specifically on grounds that it was a \textit{partisan gerrymander}.
% \par
%     Because the Joint Legislative plan was not passed by the State of Pennsylvania legislature, the Court felt no need for particular deference to it (or to the Governor’s plan, for that matter). Thus, the Court was not bound by the normal supposition with respect to districting that a plan authorized by the state need not be the ``best possible," but only \underline{NOT} unconstitutional. That ``deference" to legislative judgments and the criteria they reflect, might well have tipped the balance toward acceptance of the Joint Legislative plan were it offered as having been fully sanctioned by the State of Pennsylvania (the duly elected legislature and Governor).

### 2019-06-05 Latex/anon.tex
draft revisions

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% =================================================================
\title{Tools for Identifying Partisan Gerrymandering (PA)}
\author{Identifying Information Removed}
% 
%
% Jonathan R. Cervas - jcervas@uci.edu - University of California Irvine
% Bernard Grofman - bgrofman@uci.edu - University of California Irvine
%
% 
% First Submitted: 
% Revise and Resubmit: 
% Accepted: 
% =================================================================
\documentclass[12pt, a4paper, twoside]{article}
\RequirePackage{cervasanon}
% _________________________________________________________________
%
% _________________________________________________________________
%
% =================================================================
% =================================================================
%                       - Paper Information -
% =================================================================
% =================================================================
%
\begin{document}
    \authorone{Identifying Information Removed}
        \institone{Identifying Information Removed}
    \authortwo{}
        \instittwo{}
    \titlerunning{Tools for Identifying Partisan Gerrymandering}
    \subtitle{}
    \status{Under Review}
    \data{REDACTED}
    \abstract{In League of Women Voters v. Commonwealth of Pennsylvania (2018) the Pennsylvania Supreme Court struck down as a “severe and durable” partisan gerrymander the congressional map drawn by Republicans in 2011 and used in elections from 2012-2016.  It did so entirely on state law grounds after a three-judge federal court had rejected issuing a preliminary injunction against the plan. After Pennsylvania failed to enact a lawful remedy plan of its own (due to total disagreement as to how to proceed between the newly elected Democratic governor and the still Republican-controlled legislature), the Court then ordered into place for the 2018 election a map of its own drawn for it by a court-appointed consultant. In a split court, the Court map was endorsed only by judges with Democratic affiliations. Here we compare and contrast the 2011 and 2018 maps in terms of a variety of proposed metrics for detecting partisan gerrymandering. For further comparison purposes, we examine six proposed remedy plans. These include the remedy map proposed by the Republican legislators and that proposed by the Democratic governor, along with four other maps drawn conforming either entirely or substantially to what are often referred to as ``good government” standards, and with no election outcome or party registration data used in the creation of three of them. We argue that the 2011 map was a blatant and undisguised pro-Republican gerrymander, while the 2018 remedy map proposed by Republican legislators was a covert gerrymander (what we refer to as a stealth gerrymander); but that, as we would expect, the 2018 court drawn map can not be classified as a gerrymander.}

% ================================================================= 
    % ------- TITLE PAGE ------------------------------------------
        \maketitle
        %\input{anon_title.tex}
    % ------- END TITLE PAGE --------------------------------------
% =================================================================
%        \doublespacing
%        \linenumbers
%
\input{maintext.tex}

\singlespacing
\iffalse
\clearpage
\input{Tables/table1.tex}
\clearpage
\input{Tables/table2.tex}
\clearpage
\input{Tables/table3.tex}
\clearpage
\input{Figures/figure1.tex}
\clearpage
\input{Figures/figure2.tex}
\fi
% - Bibliography
\clearpage
    \nolinenumbers
    \newgeometry{
		top=1in,
		bottom=1in,
		inner=0.85in,
		outer=0.85in,
		ignorehead,
		ignorefoot,
		nomarginpar,
	}
\begin{multicols}{2}
    \footnotesize{\bibliography{bib}}   % Bibliography linked to bib.bib
\end{multicols}

\setcounter{section}{0} \renewcommand{\thesection}{Appendix \Alph{section}}

\clearpage

%\section{\ref{app:A} -- Methodological Appendix}
    %\linenumbers
    %\doublespacing
    %\vspace{.5in}
        %\input{appendixA.tex}
    %\clearpage
        %\input{Tables/tablea1.tex}
    %\clearpage
        %\input{Figures/figureA1.tex}
    %\clearpage

\end{document}

### 2019-06-05 Latex/appendixA.tex
draft revisions

\label{app:A}
\setcounter{table}{0} \renewcommand{\thetable}{A.\arabic{table}}
\setcounter{figure}{0} \renewcommand{\thefigure}{A.\arabic{figure}}
\setcounter{footnote}{0}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% =================================================================
        %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% =================================================================
% - Methodological Appendix
% =================================================================
        %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% =================================================================
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

This appendix describes in more detail how we measure different indicia found in the literature. These measures include the \textit{Efficiency Gap} \citep{McGhee2017}, the \textit{Mean/Median Gap} \citep{Best2018}, and partisan symmetry calculations \citep{Grofman2007, Mcgann_et_al_2015_ELJ, McGann_et_al_2016_gerrymandering}. The first two of these measures are straightforward, though estimates will differ depending upon the elections from which they are calculated.

\subsubsection*{Efficiency Gap}
The \textit{Efficiency Gap} is calculated as defined in \citet{Mcghee2014}, where all the party’s votes are wasted if they lose the district, and all the winner’s votes over 50\% are wasted.  The difference between each party’s wasted votes is then divided by the total votes cast to produce the \textit{Efficiency Gap}, with a value of zero denoting what is regarded as ideal.  As noted in the text, this is equivalent to taking an aggregate swing ratio of two as ideal.

\subsubsection*{Mean/Median Gap}
The \textit{Mean/Median Gap} is the difference between the average vote percentage for a party and its share in the median district when districts are sorted according to two-party vote share.  This measure is a variant of \textit{skewness}, such that when the mean is substantially higher or lower than the median, this is indicative of bias.

\subsubsection*{Partisan Bias}
There are a number of different ways to estimate \textit{Partisan Bias} based on the shape of the votes to seats distribution \citep{Grofman1983, Browning_King_1987_seats_votes, GelmanKing1994_unifiedAJPS, Grofman_et_al_1997_SwingRatio_Bias, Zingher2016_bias_swingratio_JEPP}. \new{ \textit{Partisan Bias} tells us if an electoral system asymmetrically awards seats to one party at the expense of another. An election is not biased if at 50\% of the vote, each party gets 50\% of the seats, and at any percentage of the vote, the other party would receive the same share of the seats had they received that vote percentage.} Hypothetical elections are constructed by incrementally adding (or subtracting) one percentage point to find aggregate seat outcomes under differing mean vote shares.  The resulting (s,v) points on the votes to seats curve are then converted to a log odds form by using $ log(\frac{s}{1-s}) $ as the dependent variable, and $ log(\frac{v}{1-v}) $ as the independent variable, and the points that fall between some range, usually 45\% and 55\% vote share are entered into a regression.  \textit{Partisan Bias} is then calculated from the intercept of this regression using an exponential transformation (for details see \citet{Grofman1983}). \textit{Partisan Bias} is calculated from the Democratic perspective, so a negative bias indicates bias against Democrats. We also examine the standard errors to determine the probability that the observed bias is not due to random chance. Table \ref{tab:a1} also reports the \textit{Swing Ratio}, a measure of responsiveness that is defined as the slope of the same regression used to generate \textit{Partisan Bias}. \new{Where to center the seats/votes curve is an unresolved question in the literature \citep{Kastellec_et_al_2008_PS}. In \citet{GelmanKing1994_unifiedAJPS}'s \texttt{JudgeIt} program, the choice is to center the election at the average Democratic vote share. We've run the numbers in three ways, the aforementioned way, centering at a tied 50/50 election, and at the actual election results.  All three equations provide similar estimates as do the results from \texttt{JudgeIt}. We report here the average bias and swing ratios obtained from \texttt{JudgeIt II} in \texttt{R}. For more information on the method and uncertainty around \texttt{JudgeIt} estimates, see \citet{GelmanKing1994_unifiedAJPS}.}

\subsubsection*{Results}
Table \ref{tab:a1} in reports the measures of gerrymandering described above. No matter what measure one choices to score the plans on, the \textit{2011} unconstitutional plan is the worst, by far. The \textit{Joint Legislative} plan proposed by the Republican leaders of the state legislature is second worst on two of the three measures, and third worse on the other. By way of contrast, the \textit{Court Remedial} plan of 2018 scores the best on two of three measures, and second best on the other. The information provided in Table \ref{tab:a1} show that bias, any way measured, always points in the direction of pro-Republican.  The magnitude of the bias varies by plan and by measure, though the unconstitutional \textit{2011} plan stands out as egregious. We can not tell much about the amount of gerrymandering based solely on any of the included measures. Though the unconstitutional plan of \textit{2011} scores higher levels of bias then any other plan we have analyzed, it is difficult to derive meaning from these numbers since they don't include measures of uncertainty and the court has yet to establish a baseline, which when crossed, is \textit{indicia} of partisan gerrymandering. \new{This type of analysis does allow use to measure the relative consequences of gerrymandering among plans, on an equal playing field, using identical electoral data. The results of these analyses provide additional evidence that the \textit{2011} plans is without doubt an egregious gerrymander, the \textit{Joint legislative} plans is a \textit{Stealth Gerrymander}, and the \textit{2018 Court Remedial plan} is \underline{not} a partisan gerrymander.}

We also offer an alternative way of presenting district specific vote share or projected vote share data, of the sort provided in Table \ref{tab:2} of the main text. Figure \ref{fig:a1}’s shadings indicate the extent to which the districts are tilted toward one or the other party.

### 2019-06-05 Latex/method_appendix.tex
draft revisions

% To assist us in our evaluations of evidenced for partisan gerrymandering in these plans we also consider, for comparison purposes, four other maps drawn conforming either entirely or almost entirely to what the Pennsylvania Supreme Court referred to as traditional districting criteria (often referred to as ``good government" criteria). One is taken from an on-line blog the \textit{DailyKos} (\href{https://bit.ly/2o3PeuX}{Brief in Support of Proposed Remedial Districting Plans as Amici Curiae Adele Schneider and Stephen Wolf, LWV, PA. No. 159 MM 2017}); the other three were drawn by the present authors. Three of the four maps were drawn with no election outcome or party registration data used in their creation.
\par
%The criteria we examine in evaluating plans are only those highlighted by the Pennsylvania Supreme Court or found in the expert witness testimony in the case, such as the \textit{mean/median gap} \citep{Best2018} and the \textit{efficiency gap} \citep{McGhee2017_ELJ}. \new{Thus, we do not provide evidence about partisan bias as defined in \citet{Tufte1973}, though this was once thought to be the most promising approach to detecting and measuring gerrymandering \citep{Grofman2007}, or more recent measures introduced, such as \textit{declination} \citep{Warrington2018} in the main text, though we provide this data in the Appendix}. 
% We do not discuss district-specific criteria for identifying gerrymandering, since, although some district-specific evidence was presented about the 2011 map in the trial testimony by an expert who was highly knowledgeable about communities of interest in the state, the \textit{LWV} opinion itself offer a statewide standard for a finding of unconstitutional partisan gerrymandering. Consequently, we limit ourselves metrics that report results for a jurisdiction as whole.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        % \footnote{Given the U.S. Supreme Court's remand of the \textit{Gill} case, it appears that any federal standard would need to be district specific. \new{The court heard oral arguments in two partisan gerrymandering cases on March 26, 2018 (Rucho, et. al. v. Common Cause, et. al [1:16-CV-1026]) \& (Lamone, et. al. v. Benisek, et. al. [1:13-cv-03233-JKB]), with decisions expected in June 2019. In oral arguments, Justices Gorsuch and Kavanaugh directed their questions asking if measures of partisan gerrymandering are really an attempt at creating proportional representation, which the court previously stated is not guaranteed by the US Constitution in \textit{Davis v. Bandemer} (1986). In fact, unlike in \textit{Gill}, the words \textit{efficiency gap} were uttered only once in \textit{Rucho} and not at all in \textit{Lamone}, by Justice Gorsuch in a comment comparing the proportionality claims being made in the current case essentially being the same as those in \textit{Gill}.}}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
\par
% In our discussion of how to judge whether a plan as a whole is a partisan gerrymander, we highlight an important distinction between \textit{neutral plans} and \textit{fair plans} -- each of which reflects a different approach to defining what constitutes gerrymandering. \textit{Neutral plans} refer to those that are drawn entirely with respect to traditional good government criteria with no attention paid to partisan considerations. One way to define partisan gerrymandering is with respect to a baseline defined by the set of feasible \textit{neutral plans} -– and this was how one of the plaintiff’s experts in the case, Professor Jowei Chen \cite[cf. ][]{Chen2013} shaped his testimony.
% % ================================================================= 
% % -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% % -----------------------------------------------------------------
%         \footnote{Similarly, \citet{Grofman2018_ELJ} asserts that ``Gerrymandering occurs when a districting plan creates a disparate treatment of the vote share of the minority and majority voting blocs in a way that penalizes the minority in its ability to translate its voting support into seats compared to what we might expect from a plan drawn on the basis of neutral principles."}
% % ----------------------------------------------------------------- 
% % -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% % =================================================================
% In contrast, as we use the term, \textit{fair plans} are ones that recognize the partisan realities of electoral geography, and seek to limit the expected disproportionality between a party’s vote share and its seat share.
% % ================================================================= 
% % -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% % -----------------------------------------------------------------
%         \footnote{There have been two approaches to measuring ``fairness" in the academic literature on gerrymandering. The best known is based on the idea of \textit{partisan asymmetry}, i.e., equals being treated unequally. For two-party competition, the intuition underlying this approach is that, if a given vote share for party A will result in a different seat share for that party than what party B would receive with an identical vote share, this can be taken to be evidence of \textit{partisan bias}.  \textit{Partisan bias} can be calculated from a (hypothetical) seats-votes curve, with the \textit{locus classicus} \citet{Tufte1973}, and important recent work done by \citet{GelmanKing1994_unifiedAJPS} among others \citep[see review in][cf. \citealt{Brookes1959, Brookes1960, Johnston2002}]{Grofman2007}, as implemented with some improvements by scholars such as \citet{Johnston1994, Rossiter1997}, although the basic idea goes back at least as far as \citet{Dahl1956}. 
%         \par 
%         A second approach to fairness is based on having some reasonable relationship between the proportion (or expected proportion) of congressional or legislative seats captured by a given party and that party’s share (or expected share) of the \underline{statewide} two-party vote -- with proportionality of vote share and seat share the obvious ideal.  However, the U.S. Supreme Court has repeatedly rejected the claim that any group has a right to proportional representation (\textit{Davis v. Bandemer}, 1986). The \textit{efficiency gap}, which compares the aggregate cracking and packing of each of the two major party’s voting strength \citep{McGhee2017_ELJ, Mcghee2018_Rejoinder_ELJ}, and was proposed as a measure of partisan gerrymandering in testimony by Professor Christopher Warshaw at the trial phase of the \textit{LWV} litigation, and \new{by others} in other recent litigation at the federal level, can be viewed as a fairness-based approach to gerrymandering. But it does not have perfect proportionality as its ideal. As its inventor, \citet[][pg. 69]{McGhee2014_Bias_LSQ}, \citep[see also][pg. 6]{McGhee2017_ELJ}, has pointed out, the \textit{efficiency gap} shows the absence of gerrymandering only when the gap between a party’s seat share and 50 percent is twice the difference between its vote share and 50 percent. Higher values of the \textit{efficiency gap} are taken as undesirable/evidence of partisan gerrymandering. 
%         \linebreak
%         \par

%         \bg{Should we exclude this per reviewers saying not necessary to pick fights?}
%         \new{
%       These two approaches to fairness can yield quite different conclusion about gerrymandering, and we believe only the first is appropriate, especially since the literature on the \textit{efficiency gap} provides no justification for why a value of two is the ideal. Our view is that disproportionality \textit{per se} is not an indicator of gerrymandering, since it is well recognized that, in plurality single member district elections, the smaller party will be disadvantaged \citep{Gudgin2012}. Rather, gerrymandering is best tapped by measures of disparate partisan impact such as partisan bias a la \citet{Tufte1973}, or the \textit{mean-median gap} proposed by \citet{Best2018}, or other asymmetry measures, such as those proposed by \citet{Wang2016_SLR, Wang2016_ELJ}. The \textit{efficiency gap}, in contrast, is a very close relative of what \citet{Tufte1973} refers to as \textit{responsiveness} (swing ratio).  Of course, as we discuss below in the context of Pennsylvania congressional districting, even if we find evidence of partisan gerrymandering, we must ask whether it can be explained by either chance factors or the partisan electoral geography of the jurisdiction rather than intentional gerrymandering, and courts must decide when a given level of demonstrated gerrymandering rises to the level of a constitutional violation.}
%       }
% % ----------------------------------------------------------------- 
% % -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% % =================================================================
% In so doing, they will seek to compensate for any biases that may be imposed because one party’s voting strength is more geographically concentrated than another’s. When gerrymandering is judged from the standpoint of political \textit{fairness} rather than \textit{neutrality}, such geographically-linked biases can create what has been called a ``natural gerrymander."
% \par
% In Pennsylvania, for example, Philadelphia is overwhelmingly Democratic in voting. In particular, if you draw two congressional districts entirely within Philadelphia County, one of them is very likely to give Democratic candidates around 90\% of the vote, and for sure, as long as both are wholly within the County, the average vote in the two will be around 80\% Democratic no matter how you draw the two districts. There are no other large, concentrated pockets of equivalently overwhelmingly Republican voting strength. Thus, more Democratic votes will ``naturally" be ``wasted" in Philadelphia than Republican votes will be ``wasted" elsewhere in the state. \new{Additionally, the city of Philadelphia is landlocked, ie, it's position on the border reduces the ability to effectively distribute these voters among many districts in ways to create \textit{fair} plans.} Additional Democratic wasted votes come in heavily Democratic Allegheny County, the county in which the city of Pittsburgh is located. Such geographic features need to be taken into account in assessing the extent of intentional partisan gerrymandering, and the expert witness testimony in the case by experts for plaintiffs by and large did so.
% \par
% Because electoral geography can matter, distinguishing purposive effects of line drawing from consequences of ``natural gerrymandering" can be critical for courts that are seeking to assess partisan intent indirectly by looking at partisan consequences and positing that what is achieved/expected, and foreseeable by any reasonable person, can be seen as intended. But even if partisan \underline{intent} is not at issue, a court that is paying attention to partisan \underline{consequences} when it is drawing a remedy plan might have to decide to what extent it is seeking to draw a ``neutral" plan as opposed to drawing a ``fair" plan as its remedy -- using the definition of those terms given above.
% \par
% In the discussion that follows in the next section, even after taking the partisan geography into account, we argue that the \textit{2011 Enacted map} was a blatant and undisguised pro-Republican gerrymander, while the 2018 \textit{Joint legislative} map proposed by Republican legislators was a covert or \textit{stealth gerrymander}. However, notwithstanding claims to the contrary, just as we would expect, the 2018 court drawn \textit{Remedial} map was not a gerrymander. Furthermore and rather unexpectedly, the plan proposed by the \textit{Gov. Wolf}(D) also could not be labeled as a pro-Democratic partisan gerrymander.
% 
% \par \bg{We should formally define what a stealth gerrymander is and how we measure it, for future citations. Here's my first crack.} \par
% \new{\textit{Stealth Gerrymander} - A plan that is indistinguishable from ``neutral" plans in terms of traditional redistricting criteria, but non-the-less creates a partisan advantage for one party. We can test for a \textit{stealth gerrymander} by comparing plan statistics; to do so, a counterfactual set of districts must be established. These counterfactual plans must also follow traditional districting criteria, especially measures of compactness and splits of political subdivisions. Partisan bias can then be tested by simulating a seats/votes curve and applying a regression, such as the one described in \citet{Tufte1973}. This test will measure both the partisan bias in the plan, along with it's responsiveness to changes in vote shares. A \textit{Stealth Gerrymander} will return statistics on traditional districting criteria that are indistinguishable from the set of counterfactual districts, but have partisan bias and responsiveness numbers that lie on the exteriors of the range.}

% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
%        \footnote{As noted above, there is dispute in the academic literature about what elections to use in making projections, and in what other factors, if any, to include in the projections \citep{Rush2000}. Nonetheless, there can be no dispute that those doing the redistricting make use of past election data to make judgments about expected partisan outcomes and commonly do so with considerable confidence. Our view is that there is no \textit{a priori} way to determine optimal projection methods prior to actually running analyses of their predictive accuracy. \new{Our composite measure has a mean discrepancy of 3.4\% after removing noncompetitive elections, and predicts every district accurately. There is also a 94\% correlation between the projected outcomes and the actual outcomes, after adjusting for noncompetitive districts.} We also recognize that projection methods that appear deterministic are nonetheless probabilistic in their predictions, even if that point is not made \underline{explicitly}. Moreover, for most plans, we would expect to get different outcomes in very good years for Republicans than in very good years for Democrats. What makes for a durable gerrymander is the absence of competitive districts \citep{Owen1988}. This absence operates to constrain the responsiveness of the plan to changes in voter preferences and puts limits on the possible success of the minority party in other than tsunami–like wave elections. \new{Figure \ref{fig:densityplots_congressional} explicitly demonstrates this.} For alternative perspectives see e.g., \citet{Mcghee2018_Rejoinder_ELJ}, \citet[][esp. at p. 83]{Best2018}.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
%
\new{While we would like to test the accuracy of the measure for 2018 result, a number of circumstances make direct comparison unavailable. Voter Tabulation Districts (VTDs) constant from election to election, and in particular they would differ from 2016 since the congressional districts themselves have changed, requiring subsequent VTD changes. Second, the district names have changed for the court plan, making direct comparison impossible. Still, we can identify the \textit{predicted} number of districts in 2018, understanding it as a probabilistic estimation. As Table \ref{tab:plansummary} shows, our composite measure averages 51.4\% of the seats (9.26R-8.74D seats at 48.5\% of the vote), once we account for the difference in state-wide support the parties actually received in 2018. Nearly a quarter of the simulations resulted in the actual outcome of 9R-9D, and 65.6\% were within one seat.} 
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
    \footnote{\new{We can characterize the overall validity of our proxy as strong. Of the three observable elections, only 2012 can be classified as a miss. 2012 was the first election after redistricting, and changes in incumbency and other factors might be contributing to the slight miss, along with any electorate shifts or effects of candidates for the Presidency. Additionally, partisan bias seemed to be particularly large given that the Democrats recieved over 50\% of the aggregated votes, but were elected in only 5 of the 18 districts. Since we are only interested in the effects of the plan post-2016, we are confident that the composite measure is appropriate for measuring gerrymandering under different plans.}}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================

% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{ \href{https://bit.ly/2NsvXyv}{The precinct level data come from Harvard Dataverse for 2008}, and from \href{https://bit.ly/2PJMyzn}{Nathaniel Kelso’s github for 2016.}}
    % ------- END FOOTNOTE ----------------------------------------
% =================================================================
Using the 2016 composite of five state-wide elections, among the 18 districts in the Court map, five are strongly favored by the Democrats, seven are strongly favored by the Republicans, and six are competitive (\rpm 5\%).
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{For a discussion about thresholds of competitiveness, see \citet{CervasGrofman2017_PC}. \new{For the partisan breakdowns, we utilize the actual sum of the five races instead of the simulated hypothetical elections we use to measure gerrymandering.}}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
In the 2011 map, Democrats were strongly favored in only four districts, while Republicans were favored in eight. In the map adopted in 2011, the Republicans were able to consistently win 13 of the districts in every election, despite a statewide partisan split that straddled the 50\% line.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{In 2012, the Republican share of the state-wide total popular vote in the congressional elections in Pennsylvania was 49.2\%, in 2014 it was 55.5\%, and in 2016 it was 54.1\%. In 2018, it was 44.9\%.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
%
% =================================================================
% ------- TABLE ---------------------------------------------------
    \begin{center}\textbf{[INSERT FIGURE \ref{fig:barplot_partisanship} ABOUT HERE]}\end{center}
                    \input{Figures/_barplot_partisanship.tex}
% ------- END TABLE -----------------------------------------------
% =================================================================
\par
We show projections under this composite five election measure in Figure \ref{fig:barplot_partisanship}.  Here, and in Table \ref{tab:plan_estimates} in the Appendix, we provide more detail than in Table \ref{tab:plansummary} by ranking the districts from least Democratic to most Democratic in projected two-party vote share. We see that the Court plan, \textit{ceteris paribus}, is expected to yield a 9D-9R split \new{for the 2016 election}. However, because of its high number of competitive seats, this single number is somewhat misleading. Using the five-election model as our predictor, there are fewer safe Democratic seats (5) than safe Republican seats (7) in the Court plan, using a five-percentage point definition of a competitive seat. This is one fewer safe seat for Republicans and one additional safe Democratic seat than is found in the 2011 plan, but the Court plan also has five districts balanced on a knife edge, with a projected winner’s margin under two percentage points, making it highly responsive to changes in voting patterns. Thus, in the Court drawn plan, changes in vote choice or turnout
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{We do not take incumbency into account in our projections. If there is an incumbent in place, and if the district boundaries are not that much changed by redistricting, districts with highly competitive baseline partisan support will be biased toward the reelection of that incumbent. However, incumbency advantage has declined in recent years to lows which have not been seen since the 1950s \citep{Jacobson2015}. The 2016 general election average we used to calculate the projected votes include three state-wide races where the Democratic candidates did particularly well comparedd both to Hillary Clinton and to the Democratic candidate for U.S. Senate. Similarly, in Pennsylvania in 2016, Republican congressional incumbents out-polled President Trump in 9 of the 13 districts won by Republicans.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
might result in different outcomes, especially in the most marginal of districts. \new{The court plan, and those with similar relative district positions will likely see `all or nothing' seat share shifts, where the parties are limited in their ability to create disproportionate delegations.} Most importantly, changes in voter sentiment will result in incumbents losing their seats, \new{especially in those competitive seats,} precisely as envisioned by the founders.
\par
\new{The first draft of this essay was submitted before the 2018 election, so it is reasonable to ask how well our predictions did after an election was held in 2018. Indeed, as predicted in our model, the 2018 election, using the Court's Remedial map, result in a 9R-9D balance. Democrats had a good year, winning the popular vote by a 2.66 percentage point advantage. We would characterize a 9R-9D result as unbiased. We acknowledge that the Republicans won three districts by less than three percentage points, meaning the election could have easily been 6R-12D. Under the old map, measured by the composite 2016 data, it too predicts a 9R-9D split, but most tellingly, the average seat share is 52\% for the Republicans in the old map, compared to 47\% under the Remedial map. Moreover, the median share under the enacted map would have been 9R-9D, whereas under the Court Remedial plan, it was 8R-10D. At the time of submission, precinct level data was not available in a form that would facilitate evaluation of the five other maps analyzed in this essay.}
%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ================================================================ %
\subsubsection*{Evaluation of the Court Remedy in comparison with Republican and Democratic alternatives}
% ================================================================ %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
The \textit{Joint} Submission, which was \new{floated by the Republican legislature} but not passed by the legislature, was a collaboration of the Republican leaders of the two chambers, Rep. Turzai and Sen. Scarnati. While the plan adhered far more closely to the principles laid out by the court than the \textit{2011} map, the governor called it a partisan gerrymander and rejected it. He issued a statement saying ``\dots my preference would have been for the General Assembly to send me a fair map" and submitted his own map to the Court."
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{Official statement from the office of the Governor: \href{https://bit.ly/2CqIg88}{Governor Wolf Submits a Fairer Congressional Map to Supreme Court}}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
The Court rejected both in favor of the plan drawn by the Court’s consultant.
\par
Using the 2016 five election composite projection, the \textit{Joint Legislative} plan would have delivered an average of seven Democratic seats, the same as the \textit{Governor}’s plan. However, there are very important differences between the two plans once one looks at the details. One main difference between the \textit{Joint} plan and the \textit{Governor}’s is that among the seven districts likely to be won by Democrats, the Democratic percentages are cut more narrowly in the Republican plan than in the \textit{Governor}’s plan in three districts so that, in a good Republican year, Republicans will do better under their plan than under the \textit{Governor}’s plan. The second main difference between the plans operates in the same direction. In the next four most competitive seats where a Republican is projected to win under our composite measure, the Democratic vote margins are lower under the \textit{Joint legislative} plan than under the \textit{Governor}’s plan. Thus, in a good Democratic year, the Democrats will likely not fare as well under the Republican plan as under the \textit{Governor}’s plan. In sum, whether it’s a good year for Republican or a bad year for Republicans, the Republicans will fare better under their own \textit{Joint} plan than under the \textit{Governor}’s plan.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ================================================================ %
\subsubsection*{Evaluation of the Court Remedy in Comparison with DailyKos Plan}
% ================================================================ %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
The liberal blog, the \textit{DailyKos}, also submitted a plan seeking to implement the criteria that the Court laid out.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{DailyKos has prepared a number of other plans, but for space reasons we will only consider their non-partisan map. The DailyKos has long advocated for redistricting reform that creates good government plans that resemble proportional representation.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
Using our composite measure, as shown in Table \ref{tab:plan_estimates}, their plan created five safe Democratic districts, eight safe Republican districts, and five competitive districts. Based on the five-election model, in 2016 the Democrats would have been predicted to win eight seats and the Republicans 10 under this plan. Though this result is slightly worse for the Democrats in an absolute sense then the \textit{Court Remedy}, the ninth and tenth districts have margins so small that small permutations in votes could end up with an 8R-10D split. The Democrats might have preferred a plan like this over the \textit{Court's Remedy}. Like the other non-partisan plan analyzed here and those of the expert witnesses, the \textit{DailyKos} plans is highly responsive to public sentiment.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{There are also illustrative maps proposed by academics, by various interest groups, and by journalists (such as \textit{FiveThirtyEight}) that we do not analyze due to space constraints and their similarity to other plans.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
%
%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ================================================================ %
\subsubsection*{Evaluation of the Court Remedy to Plans Drawn by the Authors}
% ================================================================ %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
We have drawn two alternative plans for the House using the good government criteria required by the court in which we paid absolutely no attention to partisan information, and a third plan that started with plan \textit{V1} as a base and attempted to increase the number of Democratic districts by switching out precincts and counties between districts using 2016 election results at the presidential level. All three plans score well on good government criteria. These plans have between five and six competitive seats. In a good Democratic year such as 2008, as shown in Table \ref{tab:plansummary}, the projected results in these plans would have been 7R-11D in Plan \textit{V1} and 8R-10D in Plans \textit{V2} \& \textit{V3} using presidential results. In a good Republican year such as 2016, we estimate the result would have been 11R-7D in Plan \textit{V1}, 10R-8D in Plan \textit{V2}, and 11R-7D in Plan \textit{V3}, using only presidential results. Using the 2016 five-election synthetic projection method, we find results of 10R-8D in Plan \textit{V1}, 11R -7D in Plan \textit{V2}, and 9R -9D in Plan \textit{V3}.  Comparing 2008 and 2016 projections, all three of our plans demonstrate responsiveness to shifting preferences of the electorate. As such, these plans, along with the \textit{DailyKos} and \textit{Governor's} plan can be classified as both \textit{neutral} and \textit{fair} plans.
%
%
% =================================================================
% -- SUBSECTION -- SUBSECTION -- SUBSECTION -- SUBSECTION -- SUBS %

\par
% =================================================================
% -- SUBSECTION -- SUBSECTION -- SUBSECTION -- SUBSECTION -- SUBS %
% =================================================================
    \subsection*{Results}
% =================================================================
% -- SUBSECTION -- SUBSECTION -- SUBSECTION -- SUBSECTION -- SUBS %
% =================================================================
% FOCUS RESULTS ON ENACTED, JOINT, AND REMEDIAL PLANS. OTHER 5 PLANS HAVE DATA IN TABLE

Table \ref{tab:gerry} in reports the measures of gerrymandering described above. No matter what measure one chooses to score the plans on, the \textit{2011 Enacted} unconstitutional plan is the worst, by far. The \textit{Joint Legislative} plan proposed by the Republican leaders of the state legislature is second worst on three of the four measures, and third worse on the other. By way of contrast, the \textit{Court Remedial} plan of 2018 scores the best on three of four measures, and second best on the other. The information provided in Table \ref{tab:gerry} show that bias, any way measured, always points in the direction of pro-Republican.  The magnitude of the bias varies by plan and by measure, though the unconstitutional \textit{2011 Enacted} plan stands out as especially egregious. \new{We cannot tell much about the process by which these districts were created or the intentions to gerrymandering based solely on any of the included measures. Though the unconstitutional plan of 2011 scores higher levels of bias then any other plan we have analyzed, it is difficult to derive meaning from these numbers since they don't include measures of uncertainty and the court has yet to establish a baseline, which when crossed, is \textit{indicia} of partisan gerrymandering. This type of analysis does allow use to measure the relative consequences of gerrymandering among plans, on an equal playing field, using identical electoral data. The results of these analyses provide additional evidence that the \textit{2011 Enacted} plans is without doubt an egregious gerrymander, the Joint legislative plans is a \textit{stealth gerrymander}, and the 2018 Court Remedial plan is \underline{not} a partisan gerrymander.}

% We also offer an alternative way of presenting district specific vote share or projected vote share data, of the sort provided in Table \ref{tab:plan_estimates} of the main text. Figure \ref{fig:a1}’s shadings indicate the extent to which the districts are tilted toward one or the other party.    
% 
% 
% =================================================================
% -- SECTION -- SECTION -- SECTION -- SECTION -- SECTION -- SECTI %
% =================================================================
\section{Discussion about Partisan Gerrymandering in the Various Plans}
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
What conclusions can we draw from the analyses in the previous section? Let us first turn to the \textit{2011 Enacted} plan. Here, there is simply no doubt that, using our standards and definition, it is a partisan gerrymander no matter what features of the map we examine. Of course, as the \textit{LWV} opinion emphasizes, this map demonstrate an almost unbelievable failure to satisfy traditional districting principles -- suggesting strongly the belief of its' authors that, given the U.S. Supreme Court’s history of abdication regarding partisan gerrymandering, there were no legal checks on how egregious a map could be drawn in search of partisan gain in the 2010 redistricting round \citep{McGann_et_al_2016_gerrymandering}. Indeed, it would be difficult to create a more egregious gerrymander. District 7 of the \textit{2011} map, otherwise known as ``Goofy kicking Donald Duck", best encapsulates this idea. The district was so specifically draw as to maximize Republican advantage that it narrows to just one road, and is splinter among so many municipalities that counting them by hand would be troublesome for even the most fluent mathematicians. The mere fact that the Democrats won over 50\% of the two-party vote in 2012 congressional elections but only won 27\% of the seats speaks to the efficiency of the gerrymander. Additionally, given that in the subsequent elections the two-party vote changed but the number of seats the Democrats won did not, the success of the plan is equally characterized by its unresponsive.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{\citet{Owen1988} list two types of `optimal gerrymanders', in which we can classify the 2011 Pennsylvania plan as one that seeks to maximize expected number of seats, as opposed to one that seeks maximize the chances of a working majority.  The latter of the two is not likely to happen at the congressional district level, since, in most cases, the former actually can help lead to the latter when all 435 congressional districts are aggregated.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
\par
Let us now consider the Court adopted 2018 \textit{Remedial} plan. The antagonism of Republican state legislators to this map took on an extreme character. This included not just litigation to block the implementation of the map, but also threats to impeach the judges who voted in the majority, and labeling of the opinion as simply an illegitimate power grab by judges who are Democrats. House Speaker Mike Turzai, R-Allegheny, and Senate President Pro Tempore Joe Scarnati, R-Jefferson, in a joint statement issued on February 20, 2018, denounced the Pennsylvania Supreme Court’s map. After vowing to fight it in court, they asserted: ``The League of Women Voters maintains that it filed this suit in order to take partisan politics out of the Congressional redistricting process. This map illustrates that the definition of fair is simply code for a desire to elect more Democrats." Were the Republicans correct in this assertion? The evidence provided above suggests not. Given the probabilistic nature of projections, our own simulations reinforce the view that the partisan consequences of the \textit{Court Remedial} plan are within the parameters expected of a non-partisan plan. \new{As the results of the 2018 midterm elections verified, the court map demonstrated symmetry in the treatment of the two parties.}
\par
We turn next to the issue of whether the proposed Republican \textit{Joint Legislative} Plan and the map offered by the Democratic Governor could be considered partisan gerrymanders. Remarkably, this question was not really addressed by the Pennsylvania Supreme Court. Indeed, the Court’s endorsement of the plan that it ultimately adopted is couched entirely in good government terms (Opinion and Order, February 19, 2018, at pp. 6-7, with internal footnotes omitted), with no discussion of its partisan implications, or those of the other alternatives proposed.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
    \footnote{We should also note that the Court in its adopted plan pairs two Democratic incumbents in Philadelphia (though one, Rep. Brady, was already intending to retire,), three others who are retiring (Reps. Meehan, Dent, Shuster), and the vacant incumbent seat in old district 18 with an incumbent now running for the Senate (Rep. Barletta). Its final Order does not even bother to mention the consequences for incumbents of the adopted plan. Presumably this is because the 2011 was so tainted overall by what we might call ``partisan greed" that no deference is required to the contorted lines in it that yielded those incumbencies. But, since the Court says nothing about its reasons for not mentioning incumbent effects, we can only guess.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
Describing the adopted plan, the Court says:
% =================================================================
% -- QUOTE -- QUOTE -- QUOTE -- QUOTE -- QUOTE -- QUOTE -- QUOTE  %
% -----------------------------------------------------------------
        \begin{quote}
            ``It is composed of congressional districts which follow the traditional redistricting criteria of compactness, contiguity, equality of population, and respect for the integrity of political subdivisions. The Remedial Plan splits only 13 counties. Of those, four counties are split into three districts and nine are split into two districts. The parties, intervenors, and amici differ in how they calculate municipal and precinct splits, and, as noted earlier, the Legislative Respondents suggest that updated data on precinct and municipal boundaries does not exist. The Remedial Plan is superior or comparable to all plans submitted by the parties, the intervenors, and amici, by whichever Census-provided definition one employs (Minor Civil Divisions, Cities, Boroughs, Townships, and Census Places)

            The compactness of the plan is superior or comparable to the other submissions, according to the Reock, Schwartzberg, Polsby-Popper, Population Polygon, and Minimum Convex Polygon measures described in the Court’s January 26 Order. Here, too, the parties, intervenors, and amici disagree on the precise ways to calculate these measures, and some failed to deliver compactness scores with their submissions. By whichever calculation methodology employed, the Remedial Plan is superior or comparable. Finally, no district has more than a one-person difference in population from any other district, and, therefore, the Remedial Plan achieves the constitutional guarantee of one person, one vote.

            It is composed of congressional districts which follow the traditional redistricting criteria of compactness, contiguity, equality of population, and respect for the integrity of political subdivisions. The Remedial Plan splits only 13 counties. Of those, four counties are split into three districts and nine are split into two districts. The parties, intervenors, and amici differ in how they calculate municipal and precinct splits, and, as noted earlier, the Legislative Respondents suggest that updated data on precinct and municipal boundaries does not exist. The Remedial Plan is superior or comparable to all plans submitted by the parties, the intervenors, and amici, by whichever Census-provided definition one employs (Minor Civil Divisions, Cities, Boroughs, Townships, and Census Places)

            Accordingly, this 19th day of February 2018, the Court orders as follows: First, the Pennsylvania primary and general elections for seats in the United States House of Representatives commencing in the year 2018 shall be conducted in accordance with the Remedial Plan\dots"
          \end{quote}
% -----------------------------------------------------------------
% -- END QUOTE -- END QUOTE -- END QUOTE -- END QUOTE -- END QUOT %
% =================================================================
\par
We believe that the evidence provided above is fully consistent with the claim by the Democratic governor that \textit{Joint Legislative} was a partisan gerrymander. Indeed, in the terminology previously defined, we would call it a \textit{stealth gerrymander}. \new{The effects of the Joint plan more closely resembled the unconstitutional plan than any of the other maps analyzed here.} The degree to which the \textit{Joint Legislative} plan is a \textit{stealth gerrymander} can be best understood by looking at the good Democratic year of 2008 in Table \ref{tab:plansummary} where, in the non-partisan plans (\textit{Court Remedial}, the first two Authors’ Plans [\textit{V1} \& \textit{V2}], \textit{DailyKos}), the Republicans averaged under 7 districts using 2008 presidential and composite election projections, but in the 2018 \textit{Joint Legislative} plan as in the \textit{2011 Enacted} plan determined to be unconstitutional, they win 11 seats. \textit{Ceteris paribus}, because of inefficiency in the distribution of Democratic voters, Republicans might pick up about a one seat advantage from ``neutral" redistricting using good government criteria so, if the state is 50-50 in partisan vote share, a ``neutral" plan might not be 9R-9D, but more like 10R-8D, but certainly not 13R-5D.
% =================================================================
    % ------- FOOTNOTE --------------------------------------------
        \footnote{\citet{Chen2013}, for example, estimate this bias as about 1.45 seats (8\%).  }
    % ------- END FOOTNOTE ----------------------------------------
% =================================================================
However, there are \underline{many} different possible \textit{neutral} plans.
% =================================================================
    % ------- FOOTNOTE --------------------------------------------
        \footnote{Indeed, many proposals were submitted to the court or released on the internet that were not taken into consideration by the court.}
    % ------- END FOOTNOTE ----------------------------------------
% =================================================================
In some reasonable proportion of the ones drawn by Professor Chen for his expert witness testimony in the case, when the state votes shares are proportional, the Democratic seat shares would be closer to 9D. In his expert witness report, Professor Chen shows in his first set of simulations using only traditional redistricting criteria that compared to the actual 13R-5D districts of the \textit{2011} plan, of the 500 simulations, \underline{none} of them resulted in a 13R-5D split. The mode was 9R-9D, while average was 8.53 seats for the Republicans.
% =================================================================
    % ------- FOOTNOTE --------------------------------------------
        \footnote{\href{https://www.pubintlaw.org/wp-content/uploads/2017/06/Expert-Report-Jowei-Chen.pdf}{Expert Witness Report, Jowei Chen} (Pg. 15, 16, Figure 2).}
    % ------- END FOOTNOTE ----------------------------------------
% =================================================================
The 11R-7D outcome in the proposed \textit{Joint Legislative} plan is beyond \new{all but 93} of the \new{1,000} outcomes Professor Chen simulated, and the 13R-5D in the \textit{2011 Enacted} plan was found to be simply outside the bounds of statistical probability. \new{The partisan outcomes can be evaluated along with adherence to traditional redistricting criteria. We determine a plan to be a \textit{stealth gerrymander} when outcomes suggests asymmetry between party's ability to elect candidates, but traditional criteria is indistinguishable from random simulations. Additionally, the asymmetry can not be attribured to geographical concentration of partisans. On page 15 of Dr. Chen's expert witness report, he shows that none of his randomly simulated plans had compactness scores as low as the \textit{2011 Enacted} plan. Additionally, none split as many municipalities, and the most amount of counties split was 19, compared to the 28 in the \textit{Enacted} plan. Alternatively, the \textit{Joint Submission} fits within the bounds of expected compactness and expected number of county splits. It also is second worse all \underline{ALL} measures of gerrymandering in Table \ref{tab:gerry}, with only the 2011 enacted plan preforming worse. It, therefore, fits our definition as a \textit{stealth gerrymander}.}
\par
However, having labeled the \textit{Joint Legislative} plan a \textit{stealth gerrymander}, it is worth reminding readers that, had this plan been the original plan adopted by Republicans, it is very much an open question as to whether it would have been rejected by the Pennsylvania Supreme Court. Even though, in partisan terms, it is virtually identical to the \textit{2011} plan, it is considerably more consistent with good government criteria (see Table \ref{tab:plansummary}) -- undoubtedly, not that much worse than the \textit{Court Remedial} plan -- that it would have required detailed political election analyses like those we have done here to demonstrate its \textit{stealth gerrymander} features.
\par
Because the \textit{Joint Legislative} plan was not passed by the State of Pennsylvania legislature, the Court felt no need for particular deference to it (or to the \textit{Governor’s} plan, for that matter.) Thus, it was not bound by the normal supposition with respect to districting that a plan authorized by the state need not be the ``best possible," but only \underline{NOT} unconstitutional. That ``deference" to legislative judgments and the criteria they reflect, might well have tipped the balance toward acceptance of the \textit{Joint Legislative} plan were it offered as having been fully sanctioned by the State of Pennsylvania (the duly elected legislature and governor).
% =================================================================
    % ------- FOOTNOTE --------------------------------------------
        \footnote{The Court picked the plan among those before it that \underline{most closely satisfied good government criteria} -- which happily, thanks to Professor Persily’s expertise as consultant to the Court, turned out to be its own plan, and thus a plan which the Court could know with certainty was not intended as a partisan gerrymander for either party.}
    % ------- END FOOTNOTE ----------------------------------------
% =================================================================
On the other hand, in the presence of this \textit{stealth gerrymander} rather than the blatant gerrymander that came before it, the Pennsylvania Court might well have relied on \underline{partisan impact} evidence to reach a conclusion of unconstitutionality. The language of the Final Order implementing the Court’s own plan strongly suggests this possibility. There, the Court said about the 2011 plan that it

\par
Finally, we consider the \textit{Governor’s} plan. That plan, unlike the \textit{Joint Legislative} plan, shows itself to be responsive to changes in voting patterns. In particular, the Democrats are projected to win three fewer seats in a good Republican year (2016) than in a good Democratic year (2008). See Table \ref{tab:plansummary} and discussion of alternative plans above. There are no clear reasons to consider this plan a partisan gerrymander. The best evidence that the Governor’s plan was \underline{not} a pro-Democratic gerrymander is evidenced by the shape of its districts.  Given the geographic concentration of Democrats in Pittsburgh and Philadelphia, had this truly been a pro-Democratic gerrymander, it would have had unnecessary splits of county and municipality boundaries in the areas of Allegheny and Philadelphia counties. These are also absent in the Court plan, along with our plans and that of the DailyKos. We have no way of knowing why the Democrats took a different tack than the Republicans in framing their remedy plan. In light of the court's remedy plan, Republicans would have been better off passing the Governor's plan. In failing to do so, the new maps will be less responsive to pro-Republican changes in state-wide vote shares.
%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% =================================================================
        %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% =================================================================
\section{Conclusions and Lessons for the future; The Potential Impact of LWV}
% =================================================================
        %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% =================================================================
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
While the Pennsylvania court opinion is limited to Pennsylvania, and thus it might have seemed of importance only in Pennsylvania, in footnote 71 of the Opinion (slip op. pp.116-117), the court took what we regard as a rather unusual step. It issued what can only be called an invitation to other state courts to use the same logic it used to invalidate partisan gerrymanders in their own state. As we noted earlier, the Court pointed out that there are twelve states whose constitutions contain election clauses identical to the Pennsylvania charter, requiring elections to be ``free and equal": Arizona, Arkansas, Delaware, Illinois, Indiana, Kentucky, Oklahoma, South Dakota, Oregon, Tennessee, Washington, and Wyoming.

Only a handful of these states are ripe for partisan gerrymandering challenges -- in that (ca. 2017) three lack unified party control of the state, two are single district states, and some already have a commission drawing plans and, in others, \textit{indicia} of gerrymandering are missing. We show in Table \ref{tab:compare_measures} various information that can help us form beliefs about whether or not the congressional plans in these dozen states are partisan gerrymanders. However, as pointed out in the table notes, some measures are not well suited for calculation when the seats at issue are few in number.
%
%
% =================================================================
% ------- TABLE ---------------------------------------------------
    \begin{center}\textbf{TABLE \ref{tab:compare_measures} ABOUT HERE}\end{center}
                    \input{Tables/_tab_compare_measures.tex}
% ------- END TABLE -----------------------------------------------
% =================================================================
%
%
In looking to the future, we should also note that some states that are generally regarded as among the most pernicious partisan gerrymanders, Michigan, Ohio,  North Carolina, Wisconsin are not included among the twelve, nor is Maryland.  On the other hand, we should also note that a ``free and equal" elections clause is not the only avenue state courts might use to attack partisan gerrymanders in the future. As University of Kentucky College of Law Professor Joshua Douglas has pointed out, virtually every state constitution protects voting rights more explicitly than the U.S. Constitution does. In addition to the thirteen states that require elections to be ``free and equal", an additional thirteen have state constitutional provisions that require elections to be ``free and open", and this clause could, in principle, be used in exactly the same way as the ``free and equal" clause.
% =================================================================
    % ------- FOOTNOTE --------------------------------------------
        \footnote{We are indebted to Jonathan Lai of the Philadelphia Inquirer (personal communication, April 2018) for calling this information to our attention. See a more detailed discussion in \citet{Elmendorf2018}.}
    % ------- END FOOTNOTE ----------------------------------------
% =================================================================
Such a clause is found in a number of the states widely regarded as having the worst congressional gerrymandering. \new{Additionally, other routes besides the courts are recognized as potential solutions to egregious gerrymandering \citep{Wang_et_al_2019_Labortories_UPJCL}.} We believe that the \textit{LWV} decision and the further comparative analyses of alternative plans presented in the previous section of this paper will help other state courts navigate their way to decisions that strike down some partisan gerrymanders as unconstitutional, while allowing others to remain in place on the grounds that either they are not that severe, or they are unlikely to be lasting in that they have sufficiently many competitive seats as to be responsive to changes in voting patterns.

	To simplify the exposition, we report the Republican share of the two-party vote ($\delta_{i}$). Districting plans are represented by $\mathcal D $, (e.g., $\mathcal D_{enacted}$, $\mathcal D_{remedial}$, \dots, $\mathcal D_{j}$), and each election in year $ y $ has a district vote distribution $ \Delta_{y} = [\delta_{y1}, \delta_{y2}, \dots, \delta_{yi}] $. To find the overall state-wide vote, we calculate the average district vote share, $ \bar{\Delta}_{\mathcal D y} = \frac{1}{n}\sum\limits_{i=1}^{n} \delta_{yi} $. By averaging, we reduce the influence of turnout variation between districts,. This average is a useful state-wide estimate of voter sentiment \citep[see e.g.][]{Kastellec_et_al_2008_PS}.

### 2019-06-05 Latex/_maintext.tex
2018 composite estimation

@@
 %        =================================================================
-% =================================================================
-% -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE  % 
-% -----------------------------------------------------------------
+
 \par
@@
     But we do wish to do a further robustness check on the plausibility of using our five election composite to calculate the various metrics designed to measure potential gerrymandering.
-One way to do this is to see how well that composite (calculated at a hypothetical 50\% vote share) mimics the results of the values for the four metrics for the 2011 map that we previously obtained using the actual congressional election results for that map.
+One way to do this is to see how well the composite (calculated at a hypothetical 50\% vote share) mimics the results of the values for the four metrics for the 2011 map that we previously obtained using the actual congressional election results for that map.
 What we are doing is comparing a hypothetical ``normal vote” \citep{Converse1966} assumption (based on a statewide vote in which we are not taking incumbency effects into account) with the average of what we actually get in three different congressional elections.
@@
 The comparable results from Table \ref{tab:gerry} are 0.09, 0.25, 0.05, and 0.40.
-It is apparent how close these two sets of values are to one other even though one is based on actual elections with a mean Democratic (two-party) vote share close to 47\% while the other is based on projected election results set to yield a 50-50 two-part vote distribution.
+It is apparent how close these two sets of values are to one other even though one is based on actual elections with a mean Republican (two-party) vote share close to 53\% while the other is based on projected election results set to yield a 50-50 two-part vote distribution.
 If instead we calculate these four metrics using the average five election statewide vote in these elections over the period 2012-2016, a Democratic vote share of 49.1\% we get virtually identical values: 0.09, 0.24,  0.05, 0.36.
@@
 % ------------------------------------------------------------------------
-		\footnote{If we center the simulations at the 2012-2016 average statewide Republican vote of 53.3\% of the vote, we again get remarkably consistent estimates [0.09, 0.19, 0.05, 0.45].}
+		\footnote{If we center the simulations at the 2012-2016 average statewide Republican vote of 53.3\% of the vote, we again get remarkably consistent estimates [0.09, 0.19, 0.05, 0.46 ].}
 % ------------------------------------------------------------------------ 
@@
 \par
-    Our confidence in our five election projection method is further buttressed because, when we applied this method to predict the results of the 2018 election under the court-ordered remedial map, although the method had been developed prior to the 2018 election, using data no more recent than 2016, it correctly predicted that the court-ordered remedial plan would have (in this good Democratic year) an outcome of roughly 9R-9D.
+    Our confidence in our five election projection method is further buttressed because, when we applied this method to predict the results of the 2018 election under the court-ordered remedial map.
+Although the method had been developed prior to the 2018 election, using data no more recent than 2016, and not including any incumbency advantage, it predicted that the court-ordered remedial plan would have produced an outcome of between 5R-12D and 11R-7D.
+%        ================================================================= 
+%        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
+% ------------------------------------------------------------------------
+        \footnote{After uniformly shifting each district's composite such that the statewide average equalled the actual 2018 statewide average.
+The large range is indicative of the number of competitive districts that Republican candidates won.
+Had we modeled incumbency advantage instead of the `normal vote', the two Republican incumbents who won by less than 2 percentage points, along with the one who won by less than 3 percentage points would have made the actual 9R-9D outcome the likely composite outcome.}
+% ------------------------------------------------------------------------ 
+%        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
+%        =================================================================
+
 % =================================================================

### 2019-06-05 R/Simulations.R

@@ -267,11 +267,10 @@ enacted533 <- sim.election(
 	yr=50, sims=1000, 
 	sigma=sigma)
 
-		enacted533seats <- (unlist(lapply(enacted.sims.5050, seats)))
-		enacted533votes <- (unlist(lapply(enacted.sims.5050, mean)))
+		enacted533seats <- (unlist(lapply(enacted533, seats)))
+		enacted533votes <- (unlist(lapply(enacted533, mean)))
 out.sims.stats("Enacted 53.3%", enacted533seats, enacted533votes)
 
-bias()
 
 enacted533.list.bias <- lapply(enacted533, bias)
 enacted533.unlist.bias <- do.call(rbind, enacted533.list.bias)
@@ -283,7 +282,28 @@ enacted533.meanmedian <- simulation(enacted533, meanmedian)
 enacted533.declination <- simulation(enacted533, declination)
 	-1 * mean(enacted533.declination)
 enacted533.bias <- enacted533.unlist.bias
-                               
+  
+
+remedial.at.2018 <- sim.election(
+	votes=court.comp, 
+	center = mean.w(house.2018.votes, house.2018.turnout), 
+	yr=2018, sims=1000, 
+	sigma=sigma)
+
+		remedial.at.2018.seats <- (unlist(lapply(remedial.at.2018, seats)))
+		remedial.at.2018.votes <- (unlist(lapply(remedial.at.2018, mean)))
+out.sims.stats("Remedial Centered at 2018 Actual", remedial.at.2018.seats, remedial.at.2018.votes)
+
+enacted533.list.bias <- lapply(enacted533, bias)
+enacted533.unlist.bias <- do.call(rbind, enacted533.list.bias)
+	mean(enacted533.unlist.bias)
+enacted533.eg <- simulation(enacted533, eg_TP)
+	-1 * mean(enacted533.eg)
+enacted533.meanmedian <- simulation(enacted533, meanmedian)
+	mean(enacted533.meanmedian)
+enacted533.declination <- simulation(enacted533, declination)
+	-1 * mean(enacted533.declination)
+enacted533.bias <- enacted533.unlist.bias                             
 # gerry.tab <- function(x) {
 # 	rbind(
 # 		r(mean(get(paste0(x, ".eg")))),

### 2019-06-05 Latex/_maintext.tex
Update _maintext.tex

@@
 \par
-    Our confidence in our five election projection method is further buttressed because, when we applied this method to predict the results of the 2018 election under the court-ordered remedial map.
+    Our confidence in the five election projection method is further buttressed because, when we applied this method to predict the results of the 2018 election under the court-ordered remedial map.
 Although the method had been developed prior to the 2018 election, using data no more recent than 2016, and not including any incumbency advantage, it predicted that the court-ordered remedial plan would have produced an outcome of between 5R-12D and 11R-7D.

### 2019-06-06 Latex/_maintext.tex
Update _maintext.tex
Latex files update

@@
 \par
-    In addition to the rejected 2011 and the court-ordered 2018 map, we also address the partisan and other features of two maps offered to the Pennsylvania Court as a proposed remedy for constitutional violation found in the 2011 map: the \textit{Joint Plan} offered by Republican legislators from both chambers, and the congressional map proposed by the Democratic governor (\textit{Gov.
+    In addition to the rejected 2011 and the court-ordered 2018 map, we also address the partisan and other features of two maps offered to the Pennsylvania Court as a proposed remedy for constitutional violation found in the 2011 map: the \textit{joint legislative} plan offered by Republican legislators from both chambers, and the congressional map proposed by the Democratic governor (\textit{Gov.
 Wolf}).
@@
 % •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-    As noted above, to contrast the Court-ordered Remedial plan with the 2011 Enacted plan, and to make further comparisons to the two most politically important remedy plans offered to the Court, the Joint Plan proposed by the Republicans leaders of the two legislative chambers, and the Governor’s proposed plan as a Democratic alternative, we will look at a number of features of each of these four plans as a whole.
+    As noted above, to contrast the Court-ordered Remedial plan with the 2011 Enacted plan, and to make further comparisons to the two most politically important remedy plans offered to the Court, the joint plan proposed by the Republicans leaders of the two legislative chambers, and the Governor’s proposed plan as a Democratic alternative, we will look at a number of features of each of these four plans as a whole.
 We do not, however, discuss district-specific criteria for identifying gerrymandering in these maps, since, although some district-specific evidence was presented about the 2011 map in the trial testimony by an expert who was highly knowledgeable about communities of interest in the state, the Pennsylvania Supreme Court opinion itself offers a \underline{statewide} standard for a finding of unconstitutional partisan gerrymandering.
@@
 % -----------------------------------------------------------------
-% County Splits, Compactness Scores of the Plans
+\input{Tables/_tab_summaries.tex}
+\begin{center} INSERT TABLE \ref{tab:summaries} ABOUT HERE \end{center}
 % •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
- \begin{center} INSERT TABLE \ref{tab:summaries} ABOUT HERE \end{center}
-% -----------------------------------------------------------------
-% -- END TABLE -- END TABLE -- END TABLE -- END TABLE -- END TABL %
-% =================================================================
 \par
@@
 %        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
+% ------------------------------------------------------------------------
         \footnote{However, we should note that some authors \citep[e.g.,][]{Grofman2019_ELJ} have argued that a partisan gerrymandering claim at a jurisdiction-wide level must also be reinforced by district specific evidence of manipulation of particular district boundaries.
@@
 %        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
+% ------------------------------------------------------------------------
         \footnote{We compiled the composite measure using both the sum of the raw votes by party and an average of the two-party vote in the five races at the district level.
@@
 %        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
+% ------------------------------------------------------------------------
         \footnote{We report the Republican share of the two-party vote ($\delta_{i}$).
@@
 This district-specific average is a useful state-wide estimate of voter sentiment though it is in part effected by the particular configuration of districts being examined \citep[see e.g.][cf.
-\cite{Grofman_et_al_1997__IntedgratedPerspective_ES}]{Kastellec_et_al_2008_PS}.In particular, incumbency advantage effects affect our ability to accurately predict vote outcomes.
+\cite{Grofman_et_al_1997__IntedgratedPerspective_ES}]{Kastellec_et_al_2008_PS}.
+In particular, incumbency advantage effects affect our ability to accurately predict vote outcomes.
 In ancillary work not reported here, we have looked at changes in congressional incumbency advantage in Pennsylvania over multiple decades.}
@@
     Our confidence in the five election projection method is further buttressed because, when we applied this method to predict the results of the 2018 election under the court-ordered remedial map.
-Although the method had been developed prior to the 2018 election, using data no more recent than 2016, and not including any incumbency advantage, it predicted that the court-ordered remedial plan would have produced an outcome of between 5R-12D and 11R-7D.
+Although the method had been developed prior to the 2018 election, using data no more recent than 2016, and not including any incumbency advantage, it predicted that the court-ordered remedial plan would have produced an outcome of 7R-11D.
 %        ================================================================= 
@@
 The large range is indicative of the number of competitive districts that Republican candidates won.
-Had we modeled incumbency advantage instead of the `normal vote', the two Republican incumbents who won by less than 2 percentage points, along with the one who won by less than 3 percentage points would have made the actual 9R-9D outcome the likely composite outcome.}
+Had we modeled incumbency advantage instead of the `normal vote', the two Republican incumbents who won re-election by less than 2 percentage points, along with the one who won by less than 3 percentage points would have made the actual 9R-9D outcome the likely predicted outcome.}
 % ------------------------------------------------------------------------ 
@@
 %        =================================================================
-
+%
 % =================================================================
@@
 % -----------------------------------------------------------------
-% Measures of Gerrymandering for the Four Considered Plans
+\input{Tables/_tab_gerry.tex}
+\begin{center}\textbf{INSERT TABLE \ref{tab:gerry} ABOUT HERE} \end{center}
 % •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
 %
-\begin{center} INSERT TABLE \ref{tab:gerry} ABOUT HERE \end{center}
-
 % =================================================================
@@
 % -----------------------------------------------------------------
-% Probabilistic Projections of Partisan Outcomes for Four Plans
+\input{Tables/_tab_prob.tex}
+\begin{center}\textbf{INSERT TABLE \ref{tab:prob} ABOUT HERE} \end{center}
 % •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
 %
-\begin{center} INSERT TABLE \ref{tab:prob} ABOUT HERE \end{center}
 \par
-    The findings from Table \ref{tab:gerry} are very clear!
-Regardless of which metric we examine, the 2011 map is the most gerrymandered.
+    The findings from Table \ref{tab:gerry} are very clear; Regardless of which metric we examine, the 2011 map is the most gerrymandered.
 Not surprisingly, the nature of the bias is in a pro-Republican direction.
@@
     Finally, we turn to the plan offered by the Democratic governor, Governor Wolf.
-To our surprise, this plan turned out also to have a pro-Republican tilt, nearly to the same extent as the Joint Legislative plan for three of the four metrics in Table \ref{tab:gerry}.
+To our surprise, this plan turned out also to have a pro-Republican tilt, nearly to the same extent as the joint legislative plan for three of the four metrics in Table \ref{tab:gerry}.
 This puzzling finding suggests that the Governor might have been under pressure from Democratic incumbents not to reduce their expected vote margins, or that he sincerely intended it to be a compromise that the Republican legislature might agree to.
-However, there is one important difference between the Joint plan and that of the Governor: among the seven districts likely to be won by Democrats, the Democratic percentages are cut more narrowly in the Republican plan than in the Governor’s plan in three of the districts so that, in a good Republican year, Republicans will do better under their plan than under the Democratic plan.
+However, there is one important difference between the joint plan and that of the Governor: among the seven districts likely to be won by Democrats, the Democratic percentages are cut more narrowly in the Republican plan than in the Governor’s plan in three of the districts so that, in a good Republican year, Republicans will do better under their plan than under the Democratic plan.
 \par
@@
 %        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
+% ------------------------------------------------------------------------
         \footnote{\citet{Chen2013} estimate this bias as about 1.45 seats (8\%) but Chen’s expert witness testimony in the Pennsylvania cases generates a lower estimate of natural bias until one builds in the need to protect minority opportunity districts.
 Here, we need to be careful since the minority population in such districts in the enacted map may be larger than actually needed to assure effective minority representation, and the act of maintaining or even increasing high levels of minority population in the district may be a disguised way of packing Democratic voters so as to advantage the interests of the Republican party.}
-% ----------------------------------------------------------------- 
+% ------------------------------------------------------------------------ 
 %        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
 %        ================================================================= 
-    For example, in the first set of 500 simulations in his expert witness report, based only on traditional redistricting criteria, but not including voting rights protections, Professor Chen found that none resulted in the 13R-5D split as happened in 2012 and subsequent years.
+    For example, in the first set of 500 map simulations in his expert witness report, based only on traditional redistricting criteria, but not including voting rights protections, Professor Chen found that none resulted in the 13R-5D split as happened in 2012 and subsequent years.
 The mode was 9R-9D.
@@
 %        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
+% ------------------------------------------------------------------------
         \footnote{Expert Witness Report, Jowei Chen (Pg. 15, 16, Figure2).}
-% ----------------------------------------------------------------- 
+% ------------------------------------------------------------------------ 
 %        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
@@
 Neutral plans refer to those that are drawn entirely with respect to traditional good government criteria, with no attention paid to partisan considerations.
-One way to define partisan gerrymandering is with respect to a baseline defined by the set of feasible neutral plans -- and this was how one of the plaintiff’s experts in the Pennsylvania case, Professor Jowei Chen \citep[cf.][]{Chen2013} shaped his testimony in the Pennsylvania case, and in other cases in which he has testified.
+One way to define partisan gerrymandering is with respect to a baseline defined by the set of feasible neutral plans -- and this was how one of the plaintiff’s experts in the Pennsylvania case, Professor Jowei Chen \citep[cf.][]{Chen2015_JOP} shaped his testimony in the Pennsylvania case, and in other cases in which he has testified.
 %        ================================================================= 
 %        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
+% ------------------------------------------------------------------------
         \footnote{Similarly, \citet{Grofman2018_ELJ} asserts that ``Gerrymandering occurs when a districting plan creates a disparate treatment of the vote share of the minority and majority voting blocs in a way that penalizes the minority in its ability to translate its voting support into seats compared to what we might expect from a plan drawn on the basis of neutral principles."}
-% ----------------------------------------------------------------- 
+% ------------------------------------------------------------------------ 
 %        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
@@
 \par
-    In contrast, we may wish to use the term ``fair" plans as ones that seek to compensate for the difference in level of partisan concentration.
+    In contrast, we may wish to use the term ``fair" as plans that seek to compensate for the difference in level of partisan concentration.
 Fair plans do not seek to achieve proportional representation.
@@
 %        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
+% ------------------------------------------------------------------------
     \footnote{Note that we certainly not advocating that such a three way split of Philadelphia should have been done.}
-% ----------------------------------------------------------------- 
+% ------------------------------------------------------------------------ 
 %        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
 %        ================================================================= 
-    But the Pennsylvania Supreme Court opted not to do that, but to instead preserve county lines and draw a good government map.
-We can therefore identify the Court remedial plan as a ``neutral" plan.
+    But the Pennsylvania Supreme Court opted not to do that, but to instead preserve county lines and draw a good government map. We therefore identify the Court remedial plan as a ``neutral" plan.
 \par
-  \par
-
 %
@@
 It issued what can only be called an invitation to other state courts to use the same logic it used to invalidate partisan gerrymanders in their own state.
-As we noted earlier, the Court pointed out that there are twelve states whose constitutions contain election clauses identical to the Pennsylvania charter, requiring elections to be “free and equal”: Arizona, Arkansas, Delaware, Illinois, Indiana, Kentucky, Oklahoma, South Dakota, Oregon, Tennessee, Washington, and Wyoming.
+As we noted earlier, the Court pointed out that there are twelve additional states whose constitutions contain election clauses identical to the Pennsylvania charter, requiring elections to be ``free and equal”: Arizona, Arkansas, Delaware, Illinois, Indiana, Kentucky, Oklahoma, South Dakota, Oregon, Tennessee, Washington, and Wyoming.
 % ================================================================= 
@@
     However, only a handful of these states are ripe for partisan gerrymandering challenges -- in that (ca.
-2017) three lack unified party control of the state, two are single district states, and some already have a commission drawing plans and, in others, \textit{indicia} of gerrymandering are missing.
+2017) three lack unified party control of the state, two are single district states, and some already have a commission drawing plans and in others, \textit{indicia} of gerrymandering are missing.
 \par
     In looking to the future, we should also note that some states that are generally regarded as among the most pernicious partisan gerrymanders, Michigan, Ohio, North Carolina, Wisconsin are not included among the twelve, nor is Maryland.
-On the other hand, we should also note that a “free and equal” elections clause is not the only avenue state courts might use to attack partisan gerrymanders in the future.
+On the other hand, we should also note that a ``''free and equal” elections clause is not the only avenue state courts might use to attack partisan gerrymanders in the future.
 As University of Kentucky College of Law Professor Joshua Douglas has pointed out, virtually every state constitution protects voting rights more explicitly than the U.S.
 Constitution does \citep{Douglas2014_RightToVote}.
-In addition to the thirteen states that require elections to be ``free and equal”, an additional thirteen have state constitutional provisions that require elections to be ``free" or ``free and open”, and this clause could, in principle, be used in exactly the same way as the ``free and equal” clause.
+In addition to the thirteen states that require elections to be ``free and equal”, an additional thirteen have state constitutional provisions that require elections to be ``free" or ``free and open”, and these clauses could, in principle, be used in exactly the same way as the ``free and equal” clause.
 % ================================================================= 
@@
 % =================================================================
-    Such a clause is found in a number of the states widely regarded as having the worst congressional gerrymandering. 
+    Such clauses are found in a number of the states widely regarded as having the worst congressional gerrymandering. 
 % ================================================================= 
@@
     Accordingly, some scholars have suggested turning to turn away from the federal judiciary altogether and instead focusing on state-level litigation \citep{Wang_et_al_2019_Labortories_UPJCL}.
-We believe that the \textit{LWV} decision and the further comparative analyses of alternative plans presented in the previous section of this paper will help other state courts navigate their way to decisions that strike down some partisan gerrymanders as unconstitutional, while allowing others to remain in place on the grounds that either they are not that severe, or they are unlikely to be lasting in that they have sufficiently many competitive seats as to be responsive to realistic changes in voting patterns.
+We believe that the \textit{LWV} decision and the further comparative analyses of alternative plans presented in the previous sections of this paper will help other state courts navigate their way to decisions that strike down some partisan gerrymanders as unconstitutional, while allowing others to remain in place on the grounds that either they are not that severe, or they are unlikely to be lasting in that they have sufficiently many competitive seats as to be responsive to realistic changes in voting patterns.
 Now that litigation has been successful in state courts on the grounds of egregious deviations from good government districting criteria, it seems plausible that in the next wave of redistricting, mapmakers may create covert gerrymanders like we've described in this essay.
@@
 \par
-    According to all four measures, the Joint Legislative plan offered by Republican leaders is nearly as much a gerrymandered map -- in a pro-Republican direction -- as is the 2011 map.
-As noted above, we believe it appropriate to label this plan as a stealth gerrymander since it achieves partisan purposes while (mostly) providing the visual appearance of a good government map (see Figure \ref{fig:maps}).
-However, having labeled the Joint Legislative plan a \textit{stealth gerrymander}, it is worth reminding readers of two important points.
-\par
+    According to all four measures, the joint legislative plan offered by Republican leaders is nearly as much a gerrymandered map -- in a pro-Republican direction -- as is the 2011 map.
+As noted above, we believe it appropriate to label this plan as a \textit{stealth gerrymander} since it creates partisan assymetries while (mostly) providing the visual appearance of a good government map (see Figure \ref{fig:maps}).
+However, having labeled the joint legislative plan a \textit{stealth gerrymander}, it is worth reminding readers of two important points.
 First, had this plan been the original plan adopted by Republicans, it is very much an open question as to whether it would have been rejected by the Pennsylvania Supreme Court.
-Even though, in partisan terms, it is nearly identical to the 2011 plan, it is considerably more consistent with good government criteria (see Table \ref{tab:summaries}) -- not that much worse than the Court Remedial plan -- that it would have required detailed political election analyses like those we have done here to demonstrate its stealth gerrymander features, plus a court willing to rule on political grounds rather than good government grounds that it was unconstitutional.
-Because the Joint Legislative plan was not passed by the State of Pennsylvania legislature, the Court felt no need for particular deference to it (or to the Governor's plan, for that matter).
-Thus, the Court was not bound by the normal supposition with respect to districting that a plan authorized by the state need not be the ``best possible," but only unconstitutional.
-That ``deference" to legislative judgments and the criteria they reflect, might well have tipped the balance toward acceptance of the Joint Legislative plan were it offered as having been fully sanctioned by the State of Pennsylvania (the duly elected legislature and Governor).
+Even though, in partisan terms, it is nearly identical to the 2011 plan, it is considerably more consistent with good government criteria (see Table \ref{tab:summaries}) -- not that much worse than the Court Remedial plan -- that it would have required detailed political election analyses like those we have done here to demonstrate its \textit{stealth gerrymander} features, plus a court willing to rule on political grounds, rather than good government grounds that it was unconstitutional.
+Because the joint legislative plan was not passed by the State of Pennsylvania's legislature, the Court felt no need for particular deference to it (or to the Governor's plan, for that matter).
+Thus, the Court was not bound by the normal supposition with respect to districting that a plan authorized by the state need not be the ``best possible", but only unconstitutional.
+That ``deference" to legislative judgments and the criteria they reflect, might well have tipped the balance toward acceptance of the Republican joint legislative plan were it offered as having been fully sanctioned by the State of Pennsylvania (the duly elected legislature and Governor).
 % ================================================================= 
@@
 There, the Court said about the 2011 plan that it
-    % =================================================================
+% =====================================================================
     % ------- QUOTE ---------------------------------------------------
@@
     % ------- END QUOTE -----------------------------------------------
-% =================================================================
+% =====================================================================
 \par
-	Thus we have some optimism that state courts can provide a venue to check partisan gerrymandering, even when it comes in the form of a \textit{stealth gerrymander}.
+	Thus, we have some optimism that state courts can provide a venue to check partisan gerrymandering, even when it comes in the form of a \textit{stealth gerrymander}.
 But we have also placed a number of caveats on any claim that state courts can compensate fully for any continued failure of the federal courts to act.
@@
 The vote share of an incumbent has historically been greater than it is for a non-incumbent or two non-incumbents running for an open seat.
+
 % %        ================================================================= 
@@
 % %        =================================================================
+
 %     A well-financed incumbent tends to suppress quality competition \citep{Jacobson2015_JOP}, vulnerable incumbents who can sense the political winds opt against running for re-election \citep{Jacobson_Kernell_1981,Ansolabehere_Snyder_2004_LSQ}, and electoral tides and varying turnout can distort the state-wide average.
@@
 % 	We construct the five-election composite of state-wide elections by aggregating voting district level (precincts) data to the congressional district level.
-% %        ================================================================= 
-% %        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% % ------------------------------------------------------------------------
-% 		\footnote{We wrote a function in R to assign all VTDs to a district based on the centroid of the VTD.
-This method creates a slight difference between it and the actual vote, but averages just -0.06\% per district.
-Most of the mis-allocation is between two adjacent districts, District 1 and District 13, both overwhelmingly Democratic districts.}
-% % ------------------------------------------------------------------------ 
-% %        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% %        =================================================================
-% 	Encouragingly, our five-election composite results in the same 13R-5D result that the 2016 congressional races delivered.
-% %        ================================================================= 
-% %        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% % ------------------------------------------------------------------------
-% 		\footnote{Whether we take the sum of the five elections or average the five elections, both methods deliver the same relative distributions.}
-% % ------------------------------------------------------------------------ 
-% %        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% %        =================================================================
+
 % 	Additionally, each district's outcome is identical to the 2016 congressional district results.
@@
 The average difference between our composite and the congressional results was just 2.16\%.
+
 % %        ================================================================= 
@@
 % %        =================================================================
+
 %     We imputed the values 25\% and 75\% for uncontested or essentially uncontested races.
@@
 We can then calculate statistics for each of these simulated elections to find a distribution for each measure of gerrymandering.
-% \par
-%     Table \ref{tab:gerry} reports the same four measures of partisan effects found in Table \ref{tab:congsum} but now for all four maps, using the projected statewide vote from our five-election composite, under the methodology described above.
-We also provide a measure of variability to show what range of reasonable outcomes for each of the plans and identify when gerrymandering effects are persistent.
-
-% \par
-%     As we might also expect, again according to all four measures, the next most gerrymandered map -- also in a pro-Republican direction -- is the \textit{Joint Legislative} plan offered by Republican legislative leaders as a remedy plan.
-The partisan gerrymandering indicators for this plan are sufficiently close to those of the 2011 map, and sufficiently far from those of the court-ordered map, that we believe it appropriate to label this plan as a ``\textit{stealth gerrymander}" since it achieves partisan purposes while (mostly) providing the visual appearance of a good government map (see \ref{fig:maps}).
-% \par
-%     However, having labeled the Joint Legislative plan a \textit{stealth gerrymander}, it is worth reminding readers that, had this plan been the original plan adopted by Republicans, it is very much an open question as to whether it would have been rejected by the Pennsylvania Supreme Court.
-Even though, in partisan terms, it is nearly identical to the 2011 plan, it is considerably more consistent with good government criteria (see Table \ref{tab:summaries}) -- not that much worse than the Court Remedial plan -- that it would have required detailed political election analyses like those we have done here to demonstrate its \textit{stealth gerrymander} features, plus a court willing to rule specifically on grounds that it was a \textit{partisan gerrymander}.
-% \par
-%     Because the Joint Legislative plan was not passed by the State of Pennsylvania legislature, the Court felt no need for particular deference to it (or to the Governor’s plan, for that matter).
-Thus, the Court was not bound by the normal supposition with respect to districting that a plan authorized by the state need not be the ``best possible," but only \underline{NOT} unconstitutional.
-That ``deference" to legislative judgments and the criteria they reflect, might well have tipped the balance toward acceptance of the Joint Legislative plan were it offered as having been fully sanctioned by the State of Pennsylvania (the duly elected legislature and Governor).

### 2019-06-06 R/GERRYfunctions.R

@@ -3,29 +3,37 @@ cat(
 •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
     LOADING FUNCTIONS. . . . . . . . . . . . . . . . . . . . . . 
 •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••\n")
-`Table` <- function(x, caption="", label="", footnote="", out=NULL)
+`Table` <- function(x, path=NULL, caption="", label="", footnote="", landscape=FALSE, out=NULL)
   {
     x <- append(x,
 paste0(
 "\n% =================================================================
 % -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE  % 
-% -----------------------------------------------------------------
-% ",
-caption,
-"\n% ••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••• \n%"),
+% -----------------------------------------------------------------\n",
+ifelse(is.null(path),
+paste0("\\begin{center} \\textbf{", caption, "} \\end{center}"),
+paste0("\\input{", path, "}")),
+
+"\n\\begin{center}\\textbf{INSERT TABLE \\ref{", label, "} ABOUT HERE} \\end{center}
+% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••\n
+\n% =================================================================
+% -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE  % 
+% -----------------------------------------------------------------",
+ifelse(landscape==T, "\n\\begin{landscape}", "")),
  after = 1)
 
     x <- append(x,
 paste0("\\end{tabular}
 \\tabnotes{", footnote, "}
-\\end{table} 
-%
+\\end{table}",
+ifelse(landscape==T, "\n\\end{landscape}", ""),
+"
 % -----------------------------------------------------------------
 % -- END TABLE -- END TABLE -- END TABLE -- END TABLE -- END TABL %
 % ================================================================= \n \n"),
 after = length(x))
 cat(paste(x, collapse = "\n"), "\n")
-cat(paste0("\n \\begin{center} INSERT TABLE \\ref{", label, "} ABOUT HERE \\end{center}\n"))
+cat(paste0())
   }
 
 

### 2019-06-06 R/Tables.R

@@ -214,7 +214,7 @@ tab_prop.gen <-
 
 prop.caption = "Probabilistic Projections of Partisan Outcomes for Four Plans"
 prop.label = "tab:prob"
-prop.footnote = "Using a Composite of Five Statewide Elections (adjusted to a 50\% Vote Share)"
+prop.footnote = "Using a Composite of Five Statewide Elections (adjusted to a 50\\% Vote Share)"
 
 tab_prop.tex <- stargazer(tab_prop.gen,
     style = "apsr", 
@@ -400,26 +400,30 @@ tab_congress_predict <- tab_congress_predict[c(-6, -26, -28, -29, -30, -31, -32)
 sink()
 
 
-Table(
-	tab_plan_summary,
+Table(tab_plan_summary,
+	path = "Tables/_tab_summaries.tex",
 	caption = plan_summary.caption,
 	label = plan_summary.label,
 	footnote = plan_summary.footnote)
 
 Table(congsum.tmp.tex,
+	path = "Tables/_tab_congsum.tex",
 	caption = congsum.caption,
 	label = congsum.label,
 	footnote = congsum.footnote)
 
-Table(tab_gerry,  
+Table(tab_gerry, 
+	path = "Tables/_tab_gerry.tex", 
 	caption = gerry.caption,
 	label = gerry.label, 
 	footnote = gerry.footnote)
 
 Table(tab_prop,
+	path = "Tables/_tab_prob.tex",
 	caption = prop.caption,
 	label = prop.label, 
-	footnote = prop.footnote)
+	footnote = prop.footnote,
+	landscape = TRUE)
 
 Table(district_summary,
 	caption = district_summary.caption,

### 2019-06-06 R/GERRYfunctions.R
Update GERRYfunctions.R

@@ -7,18 +7,18 @@ cat(
   {
     x <- append(x,
 paste0(
-"\n% =================================================================
-% -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE  % 
-% -----------------------------------------------------------------\n",
+"\n% =====================================================================
+% ▀▄▀▄▀▄ T̟A̟B̟L̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ T̟A̟B̟L̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ T̟A̟B̟L̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ T̟A̟B̟L̟E̟
+% ---------------------------------------------------------------------\n",
 ifelse(is.null(path),
 paste0("\\begin{center} \\textbf{", caption, "} \\end{center}"),
 paste0("\\input{", path, "}")),
 
-"\n\\begin{center}\\textbf{INSERT TABLE \\ref{", label, "} ABOUT HERE} \\end{center}
+"\n \\begin{center}\\textbf{INSERT TABLE \\ref{", label, "} ABOUT HERE} \\end{center}
 % •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••\n
-\n% =================================================================
-% -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE  % 
-% -----------------------------------------------------------------",
+\n% =====================================================================
+% ▀▄▀▄▀▄ T̟A̟B̟L̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ T̟A̟B̟L̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ T̟A̟B̟L̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ T̟A̟B̟L̟E̟
+% ---------------------------------------------------------------------",
 ifelse(landscape==T, "\n\\begin{landscape}", "")),
  after = 1)
 
@@ -28,9 +28,9 @@ paste0("\\end{tabular}
 \\end{table}",
 ifelse(landscape==T, "\n\\end{landscape}", ""),
 "
-% -----------------------------------------------------------------
-% -- END TABLE -- END TABLE -- END TABLE -- END TABLE -- END TABL %
[…]

### 2019-06-09 R/DataSetup.R
revising text and code

@@ -84,7 +84,7 @@ base.coef <- array(NA, c(length(yrs.nondistricting),4))
 for (y in 1:length(yrs.nondistricting))
   {
 reg <- lm(update(base.form, as.formula(. ~ .)), 
-  data=pahouse.non.redist, 
+  data = pahouse.non.redist, 
   subset = year == yrs.nondistricting[y])
 
     base.coef[y,1:2] <- coef(reg)

### 2019-06-09 R/GERRYfunctions.R

@@ -344,13 +344,13 @@ r <- function(r, d=2) round(r, digits=d)
     gamma = atan((2*mean(abo)-1) / (length(abo) / length(votes)))
   # normalize from radians to values betwen -1 and 1
   # A little extra precision just in case :)
-    return(2.0*(gamma-theta)/3.1415926535)}
+    return(-1 * (2.0*(gamma-theta)/3.1415926535))}
 
 `declin2` <- # Simplified, not transformed from: Katz, Jonathan N. et al. 2018. Theoretical Foundations and Empirical Evaluations of Partisan Fairness in District-Based Democracies *. https://gking.harvard.edu/files/psym_2.pdf (Accessed March 16, 2019).
   function(vs) {
     abo = vs[vs > 0.5]   # districts won by party A
     bel = vs[vs <= 0.5]  # districts won by party B
-    ((mean(abo) - 0.5) / (sum(find.winner(abo))/length(vs))) - ((0.5 - mean(bel)) / (1 - sum(find.winner(abo))/length(vs)))
+    (-1 * (mean(abo) - 0.5) / (sum(find.winner(abo))/length(vs))) - ((0.5 - mean(bel)) / (1 - sum(find.winner(abo))/length(vs)))
   }
 # •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
 # ••• MEAN/MEDIAN••••••••••••••••••••••••••••••••••••••••••••••••••
@@ -370,18 +370,20 @@ r <- function(r, d=2) round(r, digits=d)
 eg_TP <- function(votes) 
   {
     y <- cbind.data.frame(votes, 1-votes)
-      return(eff_gap(y[1], y[2]))
+      return(-1 * eff_gap(y[1], y[2]))
   }
-
+# •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
+# •••• GERRY DISPLAY ••••••••••••••••••••••••••••••••••••••••••••••
+# •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
   gerry <- function(x, toggle=TRUE)
     {
     Seats = paste0(" [", seats.print(x), "]")
     SeatPER = percent(seats(x))
     Votes = percent(mean(default.unc(x)))
[…]

### 2019-06-09 R/Plots.R

@@ -418,7 +418,7 @@ text(.55, .55, plan_names[plan], cex = 1.2, srt = 90, xpd = NA)
 dev.off()
 
 Figure(
-	path="Figures/fig_maps.pdf", 
+	path="Figures/_maps.tex", 
 	caption="Maps Of The Proposed Pennsylvania Congressional Districts", 
 	label="fig:maps", 
 	footnote="Maps a drawn with a Mercator projection. Shapefiles were obtained from the Pennsylvania Supreme Court website for the four government plans.")

### 2019-06-09 R/Simulations.R

@@ -149,24 +149,21 @@ return(
 # SIMULATE MAPS
 	head(pa.redist.dta)
 	enacted.comp <- composite(pa.redist.dta, "enacted")$COMPOSITE
-	court.comp <- composite(pa.redist.dta, "court")$COMPOSITE
 	joint.comp <- composite(pa.redist.dta, "joint")$COMPOSITE
 	govwolf.comp <- composite(pa.redist.dta, "gov")$COMPOSITE
-
+	court.comp <- composite(pa.redist.dta, "court")$COMPOSITE
+cat("Court-remedial 2018 set to Congressional 2018 vote-share and adjusted for incumbency\n",
+"	", percent(seats(court.comp+(incum_advant*house.inc.2018))), "\n")
 	enacted.pres <- composite(pa.redist.dta, "enacted")$PRES
-	court.pres <- composite(pa.redist.dta, "court")$PRES
 	joint.pres <- composite(pa.redist.dta, "joint")$PRES
 	govwolf.pres <- composite(pa.redist.dta, "gov")$PRES
+	court.pres <- composite(pa.redist.dta, "court")$PRES
 
 enacted.sims.5050 <- sim.election(
 	votes=enacted.comp, 
 	center = 0.50, 
-	yr=50, sims=1000, 
-	sigma=sigma)
-court.sims.5050 <- sim.election(
-	votes=court.comp, 
-	center = 0.50, 
-	yr=50, sims=1000, 
+	yr=50, 
+	sims=1000, 
 	sigma=sigma)
 joint.sims.5050 <- sim.election(
 	votes=joint.comp, 
@@ -176,6 +173,12 @@ joint.sims.5050 <- sim.election(
 	sigma=sigma)
 govwolf.sims.5050 <- sim.election(
 	votes=govwolf.comp, 
+	center = 0.50, 
+	yr=50, 
+	sims=1000, 
+	sigma=sigma)
+court.sims.5050 <- sim.election(
+	votes=court.comp, 
 	center = 0.50, 
 	yr=50, 
 	sims=1000, 
@@ -199,7 +202,7 @@ maps.sims.votes.5050 <- new.list(4)
 
 
 	out.sims.stats("Enacted 2011", maps.sims.seats.5050$enacted, maps.sims.votes.5050$enacted)
-	out.sims.stats("Court court", maps.sims.seats.5050$court, maps.sims.votes.5050$court)
+	out.sims.stats("Court remedial", maps.sims.seats.5050$court, maps.sims.votes.5050$court)
[…]

### 2019-06-09 R/Tables.R

@@ -154,83 +154,6 @@ district_summary <- district_summary[c(-6, -28, -29, -30, -31)]
 # =================================================================
 # Measures of Gerrymandering for the Eight Considered Plans
 # =================================================================
-
-nintyfive <- function(x, percent=FALSE) {
-	if (percent==TRUE){ return(paste0("(", percent(r(quantile(x[!is.na(x)], 0.025),d=1)), ", ", percent(r(quantile(x[!is.na(x)], 0.975),d=1)), ")")) }
-	return(paste0("(", r(quantile(x[!is.na(x)], 0.025)), ", ", r(quantile(x[!is.na(x)], 0.975)), ")"))
-	}
-
-gtab <- function(x) {
-	p <- (paste0(x, ".sims.5050"))
-	p.tmp <- get(p)
-	s.tmp <- maps.sims.seats.5050[[x]]
-	v.tmp <- maps.sims.votes.5050[[x]]
-	x.bias <- get(paste0(p, ".bias"))
-	x.eg <- get(paste0(p, ".eg"))
-	x.mm <- get(paste0(p, ".meanmedian"))
-	x.declin <- get(paste0(p, ".declination"))
-		return(c(
-			r(mean(x.bias)), 
-				nintyfive(x.bias), 
-			r(mean(x.eg)), 
-				nintyfive(x.eg), 
-			r(mean(x.mm)), 
-				nintyfive(x.mm),
-			r(mean(x.declin)), 
-				nintyfive(x.declin)
-			))
-		}
-
-ptab <- function(x) {
-	p <- (paste0(x, ".sims.5050"))
-	p.tmp <- get(p)
-	s.tmp <- maps.sims.seats.5050[[x]]
-	v.tmp <- maps.sims.votes.5050[[x]]
-		return(c(
-			paste0(r(sum(s.tmp*18)/(1000*18)*18), "R-", r(18-sum(s.tmp*18)/(1000*18)*18), "D"),
-			paste0("(", r(quantile(s.tmp, 0.025),d=1)*18, "R-", 18-r(quantile(s.tmp, 0.025),d=1)*18, "D, ", r(quantile(s.tmp, 0.975),d=1)*18, "R-", 18-r(quantile(s.tmp, 0.975),d=1)*18, "D)"),
-			paste0(median(s.tmp)*18, "R-", 18-median(s.tmp)*18, "D"),
-			percent(sum(1 * do.call(rbind, lapply(p.tmp, function(x) mean(find.winner(x)))) > 0.5) / 1000),
[…]

### 2019-06-09 R/Tables.R
Tables update

@@ -165,13 +165,13 @@ gerry.table.gen <-
 	colnames(gerry.table.gen) <- c(plan_names)
 	rownames(gerry.table.gen) <- c(
 		"Partisan Bias", 
-		"SD1",
+		"CI1",
 		"Efficiency Gap",
-		"SD2",
+		"CI2",
 		"Mean/Median",
-		"SD3",
+		"CI3",
 		"Declination",
-		"SD4")
+		"CI4")
 
 gerry.caption = "Measures of Gerrymandering for the Four Considered Plans"
 gerry.label = "tab:gerry"
@@ -187,9 +187,9 @@ tab_gerry.tex <- stargazer(gerry.table.gen,
     title= gerry.caption, 
     label= gerry.label,
 	notes = gerry.footnote)
+	tab_gerry.tex <- gsub("CI[1-9]", "", tab_gerry.tex)
 tab_gerry <- tab_gerry.tex[c(-6, -17, -18, -19, -20)]
 # ••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-
 # ••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
 
 tab_prop.gen <- 

### 2019-06-12 R/GERRYfunctions.R
updates

@@ -208,13 +208,13 @@ std <- function(x) {
   (mean(x[!is.na(x)]) - x) / sd(x[!is.na(x)])
   }
 
-ci <- function(x)
-  {
-    r <- summary(lm(x~1))
-      m <- coef(r)[1]
-      se <- coef(r)[2]
-      return(qt(0.975, df=length(x)-1) * se)
-  }
+# ci <- function(x)
+#   {
+#     r <- summary(lm(x~1))
+#       m <- coef(r)[1]
+#       se <- coef(r)[2]
+#       return(qt(0.975, df=length(x)-1) * se)
+#   }
 
 `quick.summary` <- function (x) 
   {
@@ -326,6 +326,9 @@ sim.election <- function(votes= NULL, center=house.2016.votes, incumbency=NULL, 
   }
 
 r <- function(r, d=2) round(r, digits=d)
+
+
+
 # =================================================================
 # -- GERRYMANDER MEASURES -- GERRYMANDER MEASURES -- GERRYMANDER ME
 # =================================================================
@@ -402,6 +405,12 @@ eg_TP <- function(votes)
     return(a)
     }
 
+p_value <- function(x){
+  l <- quantile(x[!is.na(x)], 0.025)
+  u <- quantile(x[!is.na(x)], 0.975)
+  t <- mean(x)/sd(x)
+    return(2*pt(-abs(t),df=length(x)-1))
+  }
 
 nintyfive <- function(x, percent=FALSE) {
   if (percent==TRUE){ return(paste0("(", percent(r(quantile(x[!is.na(x)], 0.025),d=1)), ", ", percent(r(quantile(x[!is.na(x)], 0.975),d=1)), ")")) }

### 2019-06-12 R/Tables.R

@@ -84,9 +84,13 @@ tab_plan_summary <- append(tab_plan_summary,
 # =================================================================
 # ••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
 # SUMMARY OF ELECTIONS
-congsum.caption = "US House Election Summaries (PA 2012-2016)"
+percent(sum(house.rep.2012)/sum(house.2012.turnout))
+percent(sum(house.rep.2014)/sum(house.2014.turnout))
+percent(sum(house.rep.2016)/sum(house.2016.turnout))
+
+congsum.caption = "U.S. House Election Summaries \\\\ \\hspace{2cm}(PA 2012-2016 Enacted Map)"
 congsum.label = "tab:congsum"
-congsum.footnote = "Uncontested races are imputed with 0.25 and 0.75 for the respective winners. All votes are calculated from the Republican perspective of the two-party vote. We've adjusted all gerrymandering measures such that negative numbers indicate bias in favor of the Democrats."
+congsum.footnote = "Calculations based on actual congressional elections in Pennsylvania under the map found unconstitutional in 2018. Uncontested races are imputed with 0.25 and 0.75 for the respective winners. Un-adjusted Republican two-party vote totals are 49.2\\% for 2012, 55.5\\% for 2014, and 54.1\\% for 2016. All votes are calculated from the Republican perspective of the two-party vote. We've adjusted all gerrymandering measures such that negative numbers indicate bias in favor of the Democrats."
 
 
 congsum.tmp <- cbind.data.frame(
@@ -103,7 +107,8 @@ congsum.tmp <- cbind.data.frame(
   	gerry(default.unc(house.2014.votes), toggle=F),
   	gerry(default.unc(house.2016.votes), toggle=F)))
   )
-congsum.tmp 
+congsum.tmp[2:3,4]  <- percent(congsum.tmp[2:3,4])
+congsum.tmp[4:7,4] <- r(as.numeric(congsum.tmp[4:7,4]))
 colnames(congsum.tmp) <- c(seq(2012,2016,2),"AVE")
 congsum.tmp.tex <- stargazer(congsum.tmp,
 	style = "apsr", 

### 2019-06-13 Latex/Tables/_tab_congsum.tex
Updates to code including automatic .tex files

@@
-% =================================================================
-% -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE  % 
-% -----------------------------------------------------------------
-% US House Election Summaries (PA 2012-2016)
+% =====================================================================
+% ▀▄▀▄▀▄ T̟A̟B̟L̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ T̟A̟B̟L̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ T̟A̟B̟L̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ T̟A̟B̟L̟E̟
+% ---------------------------------------------------------------------
+\input{Tables/_tab_congsum.tex}
+ \begin{center}\textbf{INSERT TABLE \ref{tab:congsum} ABOUT HERE} \end{center}
 % •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-%
+
+% =====================================================================
+% ▀▄▀▄▀▄ T̟A̟B̟L̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ T̟A̟B̟L̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ T̟A̟B̟L̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ T̟A̟B̟L̟E̟
+% ---------------------------------------------------------------------
 \begin{table}[!htbp] \centering 
-  \caption{US House Election Summaries (PA 2012-2016)} 
+  \caption{U.S. House Election Summaries \\ {\Large\hspace{4cm}(PA 2012-2016 Enacted Map)}} 
   \label{tab:congsum} 
@@
 Seats &  [13R-5D] &  [13R-5D] &  [13R-5D] & [13R-5D] \\ 
-Seat \% & 72.2\% & 72.2\% & 72.2\% & $0.720$ \\ 
-Votes & 51.1\% & 55.5\% & 54.2\% & $0.533$ \\ 
-Bias & 0.13 & 0.1 & 0.11 & $0.113$ \\ 
-Efficiency Gap & 0.21 & 0.11 & 0.17 & $0.163$ \\ 
-Mean/Median & 0.06 & 0.06 & 0.06 & $0.060$ \\ 
-Declination & 0.46 & 0.36 & 0.39 & $0.403$ \\ 
+Seat \% & 72.2\% & 72.2\% & 72.2\% & 72.0\% \\ 
+Votes & 51.1\% & 55.5\% & 54.2\% & 53.3\% \\ 
+Bias & 0.13 & 0.1 & 0.11 & 0.11 \\ 
+Efficiency Gap & 0.21 & 0.11 & 0.17 & 0.16 \\ 
+Mean/Median & 0.06 & 0.06 & 0.06 & 0.06 \\ 
+Declination & 0.46 & 0.36 & 0.39 & 0.4 \\ 
 \end{tabular}
-\tabnotes{Uncontested races are imputed with 0.25 and 0.75 for the respective winners.
+\tabnotes{Calculations based on actual congressional elections in Pennsylvania under the map found unconstitutional in 2018.
+Uncontested races are imputed with 0.25 and 0.75 for the respective winners.
+Un-adjusted Republican two-party vote totals are 49.2\% for 2012, 55.5\% for 2014, and 54.1\% for 2016.
 All votes are calculated from the Republican perspective of the two-party vote.
@@
 \end{table}
-%
-% -----------------------------------------------------------------
-% -- END TABLE -- END TABLE -- END TABLE -- END TABLE -- END TABL %
-% =================================================================
+% ---------------------------------------------------------------------
+% ▀▄▀▄▀▄ E͎N͎D͎ T͎A͎B͎L͎E͎ ▄▀▄▀▄▀▀▄▀▄▀▄ E͎N͎D͎ T͎A͎B͎L͎E͎ ▄▀▄▀▄▀▀▄▀▄▀▄ E͎N͎D͎ T͎A͎B͎L͎E͎ ▄▀▄▀▄▀
+% =====================================================================

### 2019-06-13 R/GERRYfunctions.R

@@ -32,11 +32,18 @@ ifelse(landscape==T, "\n\\end{landscape}", ""),
 % ▀▄▀▄▀▄ E͎N͎D͎ T͎A͎B͎L͎E͎ ▄▀▄▀▄▀▀▄▀▄▀▄ E͎N͎D͎ T͎A͎B͎L͎E͎ ▄▀▄▀▄▀▀▄▀▄▀▄ E͎N͎D͎ T͎A͎B͎L͎E͎ ▄▀▄▀▄▀
 % ===================================================================== \n \n"),
 after = length(x))
-cat(paste(x, collapse = "\n"), "\n")
-cat(paste0())
+cat(paste(latex.special.chr(x), collapse = "\n"), "\n")
+cat(paste(latex.special.chr(x), collapse = "\n"), "\n", file = paste0("Latex/", path))
   }
 
-
+latex.special.chr <- function(x) {
+  x <- gsub("$\\hat{\\mkern6mu}$", "^", x,fixed=TRUE)
+  x <- gsub("\\textasteriskcentered ", "*", x,fixed=TRUE)
+  x <- gsub("\\textbackslash ", "\\", x,fixed=TRUE)
+  x <- gsub("\\{", "{", x,fixed=TRUE)
+  x <- gsub("\\}", "}", x,fixed=TRUE)
+  x <- gsub("\\$", "$", x,fixed=TRUE)
+}
 
  `Figure` <- function(path=NULL, caption="", label="", footnote="")
   {
@@ -412,9 +419,17 @@ p_value <- function(x){
     return(2*pt(-abs(t),df=length(x)-1))
   }
 
+strs <- function(s) {
+  strs <- 
+  ifelse(s <= 0.001, "***", 
+    ifelse(0.001 < s & s <= 0.01, "**", 
+      ifelse(0.01 < s & s <= 0.05, "*", "")))
+  return(paste0("$^{", strs, "}$"))
+}
+
 nintyfive <- function(x, percent=FALSE) {
   if (percent==TRUE){ return(paste0("(", percent(r(quantile(x[!is.na(x)], 0.025),d=1)), ", ", percent(r(quantile(x[!is.na(x)], 0.975),d=1)), ")")) }
-  return(paste0("(", r(quantile(x[!is.na(x)], 0.025)), ", ", r(quantile(x[!is.na(x)], 0.975)), ")"))
[…]

### 2019-06-13 R/Tables.R

@@ -88,7 +88,7 @@ percent(sum(house.rep.2012)/sum(house.2012.turnout))
 percent(sum(house.rep.2014)/sum(house.2014.turnout))
 percent(sum(house.rep.2016)/sum(house.2016.turnout))
 
-congsum.caption = "U.S. House Election Summaries \\\\ \\hspace{2cm}(PA 2012-2016 Enacted Map)"
+congsum.caption = "U.S. House Election Summaries \\\\ {\\large\\hspace{4cm}(PA 2012-2016 Enacted Map)}"
 congsum.label = "tab:congsum"
 congsum.footnote = "Calculations based on actual congressional elections in Pennsylvania under the map found unconstitutional in 2018. Uncontested races are imputed with 0.25 and 0.75 for the respective winners. Un-adjusted Republican two-party vote totals are 49.2\\% for 2012, 55.5\\% for 2014, and 54.1\\% for 2016. All votes are calculated from the Republican perspective of the two-party vote. We've adjusted all gerrymandering measures such that negative numbers indicate bias in favor of the Democrats."
 
@@ -180,8 +180,8 @@ gerry.table.gen <-
 
 gerry.caption = "Measures of Gerrymandering for the Four Considered Plans"
 gerry.label = "tab:gerry"
-gerry.footnote = "Measures are averages of 1,000 simulations for each map using the 2016 composite. Brackets numbers are the 95\\% range."
-
+gerry.footnote = "$^{*}$p $<$ 0.05; $^{**}$p $<$ .01; $^{***}$p $<$ 0.001. Measures are averages of 1,000 simulations for each map using the 2016 composite. Brackets numbers are the 95\\% range."
+ 
 tab_gerry.tex <- stargazer(gerry.table.gen,
     style = "apsr", 
     summary=F,
@@ -207,7 +207,7 @@ tab_prop.gen <-
 	colnames(tab_prop.gen) <- c(plan_names)
 	rownames(tab_prop.gen) <- c(
 		"Mean Seat Share",
-		"SDS", 
+		"CI1", 
 		"Median Seat Share",
 		"Probability Republican Majority",
 		"Probability Democratic Majority",
@@ -227,6 +227,7 @@ tab_prop.tex <- stargazer(tab_prop.gen,
     title= prop.caption, 
     label= prop.label,
 	notes = prop.footnote)
+tab_prop.tex <- gsub("CI[1-9]", "", tab_prop.tex)
[…]

### 2019-06-13 Latex/Tables/_tab_states_compare.tex
Updated LaTeX files

        \begin{table}[!htbp]
        \caption{Information about Twelve States with Constitutional Provisions similar to Pennsylvania}
        \label{tab:states_compare}
        \begin{tabular}{
        >{\centering\arraybackslash}
        M{0.1\linewidth}|
        M{.025\linewidth}|
        M{.075\linewidth}
        M{.075\linewidth}
        M{.075\linewidth}
        M{.075\linewidth}
        M{.075\linewidth}
        M{.075\linewidth}
        M{.075\linewidth}
        M{.3\linewidth}}
        \textbf{State} & \textbf{\# CDs} & \textbf{Unified Control (2011)} & \textbf{Unified Control (2016)} & \textbf{Seats (2016)} & \textbf{Votes (2016)} & \textbf{Mean District Vote Share (Dem)} & \textbf{Median District Vote Share (Dem)} & \textbf{Efficiency Gap} & \textbf{Who does districting} \\
         \hline
        Arizona & 9 & $\surd$ (R) & $\surd$ (R) & 44.4\% & 48.1\% & 50.4\% & 49.4\% & 0.08 & Independent commission \\
        Arkansas & 4 & $\surd$ (D) & $\surd$ (R) & 0\% & 35.7\% & 35.5\% &  &  & state legislature \\
        Delaware & 1 & $\surd$ (D) & $\surd$ (D) & 100\% & 55.9\% & 56\% &  &  & NA \\
        Illinois & 18 & $\surd$ (D) &  & 61.1\% & 59.0\% & 59.6\% & 59.8\% & 0.08 & state legislature \\
        Indiana & 9 & $\surd$ (R) & $\surd$ (R) & 22.2\% & 39.9\% & 40.1\% & 35.9\% & 0.09 & state legislature \\
        Kentucky & 6 &  & $\surd$ (R) & 16.7\% & 34.3\% & 33.8\% &  &  & state legislature \\
        Oklahoma & 5 & $\surd$ (R) & $\surd$ (R) & 0\% & 30.7\% & 30.7\% &  &  & state legislature \\
        Oregon & 5 &  & $\surd$ (D) & 80\% & 56.2\% & 56\% &  &  & state legislature \\
        South Dakota & 1 & $\surd$ (R) & $\surd$ (R) & 0\% & 34.0\% & 34\% &  &  & NA \\
        Tennessee & 9 & $\surd$ (R) & $\surd$ (R) & 22.2\% & 36.4\% & 37.4\% & 31.3\% & 0.03 & state legislature \\
        Washington & 10 & $\surd$ (D) & $\surd$ (D) & 70\% & 58.8\% & 56\% & 52.3\% & -0.05 & 5-member independent commission \\
        Wyoming & 1 & $\surd$ (R) & $\surd$ (R) & 0\% & 24.3\% & 24.3\% &  &  & NA

% =================================================================
% -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE  % 
% -----------------------------------------------------------------
% Information about Twelve States with Constitutional Provisions similar to Pennsylvania
% ••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••• 
%
\end{tabular}
\tabnotes{Seats and votes are based on the 2016 five-election projection (to deal with the existence of non-contested congressional districts). Percentages are of the of district level results. This difference is why the percentages reported in columns 6 and 8 are not identical. Data from DailyKos, All about Redistricting, and Ballotpedia web sites.States with fewer than 9 districts do not have efficiency gap or median values reported because of the potential unreliability of those calculations given the small number of districts involved.}
\end{table} 
%
% -----------------------------------------------------------------
% -- END TABLE -- END TABLE -- END TABLE -- END TABLE -- END TABL %
% =================================================================

### 2019-06-13 Latex/Tables/_tabplan_estimates.tex
Updated LaTeX files

\begin{landscape}

\begin{table}
\caption{Projected Partisanship of the 18 Districts among the Different Plans According to a Five-Election Composite of Statewide 2016 elections (districts arranged from least to most Democratic)}
\label{tab:plan_estimates}
\centering
\scalebox{0.9}{
\begin{tabular}{c|cccccccc}
& \textbf{2011 Map} & \textbf{Court Remedial} & \textbf{Joint Legislative} & \textbf{Governor} & \textbf{DailyKos} & \textbf{Authors’ V1} & \textbf{Authors’ V2} & \textbf{Authors’ V3}  \\ \hline
\multirow{8}{*}{\rotatebox{90}{Least Democratic -----\textgreater{}}} & 34.30\% & 30.80\% & 34.70\% & 33.60\% & 32.10\% & 29.60\% & 34.90\% & 29.60\% \\
 & 34.80\% & 33.90\% & 34.80\% & 34.50\% & 34.00\% & 36.90\% & 34.90\% & 37.20\% \\
 & 38.50\% & 34.70\% & 35.30\% & 36.10\% & 35.00\% & 37.50\% & 36.30\% & 37.50\% \\
 & 40.20\% & 36.80\% & 35.50\% & 37.10\% & 35.90\% & 39.20\% & 37.00\% & 39.20\% \\
 & 40.30\% & 37.10\% & 41.40\% & 38.60\% & 38.80\% & 40.60\% & 37.90\% & 40.60\% \\
 & 42.10\% & 42.10\% & 42.30\% & 41.60\% & 41.50\% & 41.00\% & 41.20\% & 40.90\% \\
 & 44.40\% & 42.70\% & 43.00\% & 41.80\% & 42.90\% & 41.90\% & 42.80\% & 41.00\% \\
 & 44.70\% & 46.10\% & 43.80\% & 44.10\% & 43.30\% & 45.60\% & 44.90\% & 45.70\% \\
 & 45.50\% & 49.80\% & 46.20\% & 48.00\% & 47.90\% & 47.00\% & 46.00\% & 46.90\% \\
\multirow{9}{*}{\rotatebox{270}{Most Democratic -----\textgreater{}}} & 46.70\% & 51.20\% & 47.30\% & 48.10\% & 49.90\% & 49.60\% & 49.30\% & 50.10\% \\
 & 47.80\% & 51.50\% & 49.20\% & 49.10\% & 50.10\% & 50.30\% & 49.80\% & 50.30\% \\
 & 47.90\% & 51.50\% & 50.10\% & 52.20\% & 51.30\% & 50.50\% & 50.40\% & 50.40\% \\
 & 48.70\% & 51.50\% & 50.90\% & 53.10\% & 52.20\% & 50.90\% & 50.60\% & 50.90\% \\
 & 50.70\% & 57.30\% & 51.00\% & 56.90\% & 55.70\% & 59.00\% & 59.30\% & 59.00\% \\
 & 66.80\% & 61.90\% & 65.90\% & 67.40\% & 60.20\% & 60.20\% & 60.50\% & 60.20\% \\
 & 71.40\% & 67.00\% & 71.30\% & 69.40\% & 69.20\% & 66.30\% & 70.80\% & 66.30\% \\
 & 81.30\% & 75.90\% & 82.30\% & 74.90\% & 85.80\% & 75.90\% & 74.80\% & 75.90\% \\
 & 90.50\% & 91.40\% & 91.10\% & 88.40\% & 89.60\% & 91.40\% & 92.70\% & 91.40\% \\ \hline
 & 13R-5D & 9R-9D & 11R-7D & 11R-7D & 10R-8D & 10R-8D & 11R-7D & 9R-9D
\end{tabular}}
\end{table}

\noindent
\tabnotes{Since plans were developed \textit{de novo}, district number varies by plan. As such, we list districts by there \textit{partisanship}, from least Democratic to most Democratic. This way we can compare the relative \textit{packing} and \textit{cracking} of each plan, identify the median district, and make predictions about future elections.}

\end{landscape}

### 2019-06-16 R/Plots.R
Update Plots.R

@@ -10,344 +10,335 @@ cat(
 		CREATING PLOTS . . . . . . . . . . . . . . . . . . . . . 
 •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••\n")
 
-# % -----------------------------------------------------------------
-yrs <- seq(2012,2016,2)
-pdf(paste0("Figures/fig_sims_actual_density.pdf"), 
-	width = 5, 
-	height = 6)
-par(mfrow=c(3,1), 
-	oma= c(3,1.5,0,0), 
-	mar=c(3,1,1,1), 
-	mgp= c(2,1,0))
-for (k in 1:3)
-	{		
-		sims.tmp.yrs <- sim.election(
-			votes = default.unc(house.2016.votes), 
-			center = default.unc(get(paste0("house.", yrs[k],".votes"))), 
-			yr=yrs[k], sims=1000, 
-			sigma=sigma)
+# # % -----------------------------------------------------------------
+# yrs <- seq(2012,2016,2)
+# pdf(paste0("Figures/fig_sims_actual_density.pdf"), 
+# 	width = 5, 
+# 	height = 6)
+# par(mfrow=c(3,1), 
+# 	oma= c(3,1.5,0,0), 
+# 	mar=c(3,1,1,1), 
+# 	mgp= c(2,1,0))
+# for (k in 1:3)
+# 	{		
+# 		sims.tmp.yrs <- sim.election(
+# 			votes = default.unc(house.2016.votes), 
+# 			center = default.unc(get(paste0("house.", yrs[k],".votes"))), 
+# 			yr=yrs[k], sims=1000, 
+# 			sigma=sigma)
 
-		x <- (unlist(lapply(sims.tmp.yrs, seats))) # Vector of Seats from Simulations for each year
-		x.sd <- sd(x)
-		x.mean <- mean(x)
-		a <- density(x, na.rm=TRUE, from=x.mean-x.sd, to=x.mean+x.sd)
-		area <- abs(sum(a$y)*(a$x[1]-a$x[2]))
-		# twec <- sum(d$ecvotes[d$p>.5])/sum(d$ecvotes)
-		e1 <- density(x, na.rm=TRUE, adjust=2)
+# 		x <- (unlist(lapply(sims.tmp.yrs, seats))) # Vector of Seats from Simulations for each year
+# 		x.sd <- sd(x)
+# 		x.mean <- mean(x)
+# 		a <- density(x, na.rm=TRUE, from=x.mean-x.sd, to=x.mean+x.sd)
[…]

### 2019-06-16 R/Plots.R
Map Figure

@@ -360,7 +360,7 @@ radius <- function(x) perimeter(x) / (2 * 3.1415926535)
 		colnames(rim) <- c("x", "y")
 		c <- rbind(rim,c, c[1,])
 		dev.off()
-pdf(paste0("Figures/fig_maps.pdf"), 
+pdf(paste0("Latex/Figures/fig_maps.pdf"), 
   	width=8, 
   	height=10)
 par(mar=c(0,0,0,0),
@@ -369,7 +369,7 @@ nf <- layout(
 	rbind(mat=
 		c(0,1:2), c(3,7,8), c(4,9,10), c(5,11,12), c(6,13,14)
 		),
-	widths= c(.1,3,1.5),
+	widths= c(.15,3,1.45),
 	heights= c(.1,1,1,1,1), respect=F)
 # layout.show(nf)
 par(mar=c(0,0,0,0))
@@ -380,7 +380,7 @@ text(.55, .55, "Philadelphia", cex = 1.2, srt = 0, xpd = NA)
 for (plan in 1:4){
 par(mar=c(0,0.5,0,0))
 plot(.5,.5, xlim = c(.4, .7), ylim = c(.4, .7), type = "n", axes = F, xlab = "", ylab = "")#call empty plot to keep on same scale
-text(.55, .55, plan_names[plan], cex = 1.2, srt = 90, xpd = NA)
+text(.55, .55, plan_names[plan], cex = 2, srt = 90, xpd = NA)
 }
 	for (plan in 1:length(plans.compressed)) {
 		shapeFile.compressed <- get(plans.compressed[plan])
@@ -409,7 +409,7 @@ text(.55, .55, plan_names[plan], cex = 1.2, srt = 90, xpd = NA)
 dev.off()
 
 Figure(
-	path="Figures/_maps.tex", 
+	path="Figures/fig_maps.pdf", 
 	caption="Maps Of The Proposed Pennsylvania Congressional Districts", 
 	label="fig:maps", 
 	footnote="Maps a drawn with a Mercator projection. Shapefiles were obtained from the Pennsylvania Supreme Court website for the four government plans.")

### 2019-06-17 R/Plots.R
Cleaned up files

@@ -354,7 +354,7 @@ radius <- function(x) perimeter(x) / (2 * 3.1415926535)
 		e2 <- e + c(-.1, .1, -.1, .1)
 		x <- e[2] - (e[2]-e[1])/2
 		y <- e[4] - (e[4]-e[3])/2
-		cir <-  circle(x,y, radius=(e[2]-e[1])/2.5)
+		cir <-  circle(x,y, radius=(e[2]-e[1])/2.4)
 		c <- cbind.data.frame(x=rev(cir$x),y=rev(cir$y))
 		rim <- rbind(cbind(e2[2],e2[3]),cbind(e2[2],e2[4]), cbind(e2[1],e2[4]), cbind(e2[1],e2[3]), cbind(e2[2],e2[3])) 
 		colnames(rim) <- c("x", "y")
@@ -369,7 +369,8 @@ nf <- layout(
 	rbind(mat=
 		c(0,1:2), c(3,7,8), c(4,9,10), c(5,11,12), c(6,13,14)
 		),
-	widths= c(.15,3,1.45),
+	# widths= c(.15,3,1.45),
+	widths= c(1, 3, 2),
 	heights= c(.1,1,1,1,1), respect=F)
 # layout.show(nf)
 par(mar=c(0,0,0,0))
@@ -401,7 +402,7 @@ text(.55, .55, plan_names[plan], cex = 2, srt = 90, xpd = NA)
 		plot(rem.philly, border=NA)
 		plot(raster::crop(shapeFile.compressed,c), add=T, col="gray30", border="gray70", lwd=1)
 		plot(pa_counties.compressed, add=T, col="NA", border="gray10", lwd=3)
-		plot(raster::crop(shapeFile.compressed,c), add=T, col=NA, border="white", lwd=0.25)
+		plot(raster::crop(shapeFile.compressed,c), add=T, col=NA, border="white", lwd=0.20)
 		polygon(cir, border="black", lwd=15)
 		polygon(cir, border="white", lwd=5)
 		polygon(c, col="white", border="white")
@@ -412,7 +413,7 @@ Figure(
 	path="Figures/fig_maps.pdf", 
 	caption="Maps Of The Proposed Pennsylvania Congressional Districts", 
 	label="fig:maps", 
-	footnote="Maps a drawn with a Mercator projection. Shapefiles were obtained from the Pennsylvania Supreme Court website for the four government plans.")
[…]

### 2019-06-18 Latex/_maintext.tex
Update _maintext.tex
revising text and code

@@
+% =================================================================
 % •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-% -- SECTION -- SECTION -- SECTION -- SECTION -- SECTION -- SECTI %
+% █▀▀ █▀▀ █▀▀ ▀▀█▀▀ ░▀░ █▀▀█ █▀▀▄
+% ▀▀█ █▀▀ █░░ ░░█░░ ▀█▀ █░░█ █░░█
+% ▀▀▀ ▀▀▀ ▀▀▀ ░░▀░░ ▀▀▀ ▀▀▀▀ ▀░░▀
 % •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
@@
     \lettrine[lines=3, lraise=0.1, nindent=0em]{\color{lightgray}I}{n} \textit{League of Women Voters v.
-Pennsylvania}, decided January 22, 2018 (henceforth abbreviated \textit{LWV}), the Pennsylvania Supreme Court offered a new and distinctive test for unconstitutional partisan gerrymandering under which it found the state’s congressional districts to be a ``severe and durable" partisan gerrymander that violated a long-standing provision of the state’s constitution.
+Pennsylvania}, decided January 22, 2018 (henceforth abbreviated \textit{LWV}), the Pennsylvania Supreme Court offered a new and distinctive test for unconstitutional partisan gerrymandering under which it found the state’s congressional districts to be a ``severe and durable'' partisan gerrymander that violated a long-standing provision of the state’s constitution.
 The congressional map that was struck down was drawn in 2011 by a Republican-controlled legislature and signed into law by the then Republican governor.
@@
 \par
-    The Pennsylvania Supreme Court held that ``the fact that the 2011 Plan cannot, as a statistical matter, be a plan directed at complying with traditional redistricting requirements is \underline{sufficient} to establish that it violates the Free and Equal Elections Clause" [of the Pennsylvania State Constitution] (slip op.
+    The Pennsylvania Supreme Court held that ``the fact that the 2011 Plan cannot, as a statistical matter, be a plan directed at complying with traditional redistricting requirements is \underline{sufficient} to establish that it violates the Free and Equal Elections Clause'' [of the Pennsylvania State Constitution] (slip op.
 at p.128, emphasis added).
@@
 % -----------------------------------------------------------------
-        \footnote{``Elections shall be free and equal; and no power, civil or military, shall at any time interfere to prevent the free exercise of the right of suffrage." Pa. Const. art. I, § 5.}
+        \footnote{``Elections shall be free and equal; and no power, civil or military, shall at any time interfere to prevent the free exercise of the right of suffrage.'' Pa. Const. art. I, § 5.}
 % ----------------------------------------------------------------- 
@@
 % =================================================================
-    However, the Court also considered a variety of other indicators involving the partisan aspects of the plan in coming to the conclusion that the map created a ``severe and durable" partisan gerrymander.
+    However, the Court also considered a variety of other indicators involving the partisan aspects of the plan in coming to the conclusion that the map created a ``severe and durable'' partisan gerrymander.
 % ================================================================= 
@@
 In his review, the trial judge found, in essence, that the legislative process through which the plan had been chosen demonstrated the indicia of a partisan gerrymander, that the weird shapes of a number of the districts demonstrated the indicia of a partisan gerrymander, that the dismemberment of municipalities, townships and counties in the state demonstrated the indicia of a partisan gerrymander, that the cracking and packing of Democrat-leaning voters demonstrated the indicia of a partisan gerrymander, and that the frozen 13R-5D results over the course of three elections in a state that is highly politically competitive statewide demonstrated the indicia of a partisan gerrymander.
-And the trial judge also found that none of these features of the map could be explained as necessitated by the electoral geography and demography of the state, though he suggested the possibility that consideration such as incumbency protection might have mattered.
+The trial judge also found that none of these features of the map could be explained as necessitated by the electoral geography and demography of the state, though he suggested the possibility that consideration such as incumbency protection might have mattered.
 However, the trial court then, rather regretfully, concluded that it did not have a legal basis to declare the plan unconstitutional under either the Pennsylvania State Constitution or under then existing federal law precedents.
@@
 S.
-\_\_\_ (2018)}.}, which had overturned as a partisan gerrymander a Wisconsin legislative map, would finally adopt a (manageable) standard for identifying unconstitutional partisan gerrymandering that all courts could apply.
+\underline{\hspace{3em}} (2018)}.}, which had overturned as a partisan gerrymander a Wisconsin legislative map, would finally adopt a (manageable) standard for identifying unconstitutional partisan gerrymandering that all courts could apply.
 The Pennsylvania Supreme Court, for all practical purposes, adopted in \textit{toto} the factual finding of the trial court magistrate, Judge Brobson, but came to a diametrically opposite legal conclusion about the constitutionality of the congressional plan and the legal grounding for that finding.}
@@
 \par
-    Having invalidated the \textit{2011 Enacted} map, the Pennsylvania Supreme Court offered deferrence to the State to draw a remedy plan.
+    Having invalidated the \textit{2011 Enacted} map, the Pennsylvania Supreme Court offered deference to the State to draw a remedy plan.
 The Pennsylvania legislature failed to enact a lawful remedy plan of its own due mostly to divided government that led to disagreement as to how to proceed between the newly elected Democratic governor and the still Republican-controlled legislature.
@@
         \footnote{Pennsylvania Supreme Court Judges are elected or retained to ten-year terms in elections with party labels on the ballot.
-Judges with Democratic affiliations now constitute the majority of the Court In 2011 this was not the case.}
+Judges with Democratic affiliations now constitute the majority of the Court.
+In 2011 this was not the case.}
 % ----------------------------------------------------------------- 
@@
 Others saw this as simply sour grapes: \\ \href{https://wapo.st/2LlEVvE}{``Pennsylvania Republicans lost the redistricting battle.
-Now, they’re declaring war on the courts", https://wapo.st/2LlEVvE (Christopher Ingraham, February 22, 2018).;} \\ \href{https://www.philly.com/philly/news/politics/will-republicans-impeach-pennsylvania-supreme-court-justices-20180222.html}{``Pa.
+Now, they’re declaring war on the courts'', https://wapo.st/2LlEVvE (Christopher Ingraham, February 22, 2018).;} \\ \href{https://www.philly.com/philly/news/politics/will-republicans-impeach-pennsylvania-supreme-court-justices-20180222.html}{``Pa.
 Republicans are talking about impeaching state Supreme Court justices.
-Do they have an argument?", https://www.philly.com/philly/news/politics/will-republicans-impeach-pennsylvania-supreme-court-justices-20180222.html (Andrew Seidman \& Jonathan Lai, February 22, 2018).}}
+Do they have an argument?'', https://www.philly.com/philly/news/politics/will-republicans-impeach-pennsylvania-supreme-court-justices-20180222.html (Andrew Seidman \& Jonathan Lai, February 22, 2018).}}
 % ----------------------------------------------------------------- 
@@
 However, a 2018 remand of a Wisconsin partisan gerrymandering case, \textit{Gill v.
-Whitford} has, to a limited extent, provided clues about the current [ca.
-May 2019] state of federal partisan gerrymandering jurisprudence than anything before it.
+Whitford} (referred to interchangeably as \textit{Gill}) has, to a limited extent, provided clues about the current [ca.
+June 2019] state of federal partisan gerrymandering jurisprudence than anything before it.
 % ================================================================= 
@@
 Bandemer, 478 U.S.
-109 (1986)}, which declared partisan gerrymandering justiciable, but under which the Supreme Court has never yet (as of May 2019) struck down a plan as an unconstitutional partisan gerrymander.
-Despite the 7-2 decision on the Wisconsin remand, which rejected the view that the Court should simply bow out of the process of reviewing maps as possible unconstitutional partisan gerrymanders, the retirement of Justice Kennedy and his replacement by a more conservative Justice creates even greater uncertainty about whether, in the lifetime of the present authors, there will ever be a U.S.
-Supreme Court majority that rejects some challenged district as an unconstitutional partisan gerrymander.
-We will, relatively soon, have the answer to that question due to the inevitable appeals of partisan gerrymandering cases from North Carolina and Maryland that have already had oral argument on March 26, 2019 before the U.S.
+109 (1986)}, which declared partisan gerrymandering justiciable, but under which the Supreme Court has never yet (as of June 2019) struck down a plan as an unconstitutional partisan gerrymander.
+Despite the 7-2 decision on the Wisconsin remand, which rejected the view that the Court should simply bow out of the process of reviewing maps as possible unconstitutional partisan gerrymanders, the retirement of Justice Kennedy and his replacement by a more conservative justice creates even greater uncertainty about whether, in the lifetime of the present authors, there will ever be a U.S.
+Supreme Court majority that rejects some challenged district (or plan) as an unconstitutional partisan gerrymander.
+We will relatively soon have the answer to that question due to the inevitable appeals of partisan gerrymandering cases from North Carolina and Maryland that have already had oral argument on March 26, 2019 before the U.S.
 Supreme Court (\textit{Rucho, et.
@@
 In addition, there is a partisan gerrymandering challenge in Michigan, \textit{League of Women Voters of Michigan} v.
-\textit{Johnson} that will certainly be appealed to the U S.
+\textit{Johnson} that will certainly be appealed to the U.S.
 Supreme Court, as will the next lower court decision in \textit{Gill} itself.}
@@
 Whitford, 585 U.S.
-\_\_\_ (2018)}), the Supreme Court indicated that any standard for partisan gerrymandering must be a challenge to individual districts which demonstrates that the voters in the particular district have had their votes either ``packed" or ``cracked."
+\underline{\hspace{3em}} (2018)}), the Supreme Court indicated that any standard for partisan gerrymandering must be a challenge to individual districts which demonstrates that the voters in the particular district have had their votes either ``packed'' or ``cracked.''
 \par
@@
 Districts that have been \textit{cracked} are ones where voters of one partisan leaning have had their share of the district’s voting strength reduced so as to make it impossible (or unlikely) that members of this group will be able to carry the district.
-The standard for unconstitutional gerrymandering at the federal level -- if there is ever is one -- can thus be expected to be very different from that offered by the Pennsylvania Supreme Court, since \textit{LWV} relies on data about the features and effects of the plan as a whole, while the US Supreme Court appears to be insisting on a district-specific test of packing and cracking.
+The standard for unconstitutional gerrymandering at the federal level -- if there is ever one -- can thus be expected to be very different from that offered by the Pennsylvania Supreme Court, since \textit{LWV} relies on data about the features and effects of the plan as a whole, while the U.S.
+Supreme Court appears to be insisting on a district-specific test of packing and cracking.
 \par
@@
 \par
-    Although only directly applicable in Pennsylvania, \textit{LWV} is, we believe, nonetheless the most important partisan gerrymandering case yet decided in a final and definitive fashion this decade.
+    Although only directly applicable in Pennsylvania, \textit{LWV} is, we believe nonetheless, the most important partisan gerrymandering case yet decided in a final and definitive fashion this decade.
 % ================================================================= 
@@
 % =================================================================
-    Because the US Constitution is the minimum rights provided to all citizens, states offer additional protections.
-Language virtually identical to the provision of the state constitution that the Pennsylvania Supreme Court relied on is found in the constitutions of another dozen states, and another thirteen have very similar language \citep{Douglas2014_RightToVote}.
+    Because the U.S.
+Constitution provides for the minimum rights provided to all citizens, state laws and constitutions offer additional protections.
+Language virtually identical to the provision of the PA state constitution that the Pennsylvania Supreme Court relied on is found in the constitutions of another dozen states, and another thirteen have very similar language \citep{Douglas2014_RightToVote}.
 % ================================================================= 
@@
 \par
-    However, as we show later, some states in which possible state law challenges could be made will be insulated from those challenges by factors such as the nature of institutional control over the redistricting process in the state, or political gridlock in the state courts.
+    Some states in which possible state law challenges could be made will be insulated from those challenges by factors such as the nature of institutional control over the redistricting process in the state, or political gridlock in the state courts.
 Nonetheless, if a state court does follow the Pennsylvania Supreme Court lead, we view the expected effects on limiting partisan gerrymandering to be potentially greater than we might expect under federal law, since they would involve the overturning of whole plans, not just individual districts -- although, in practice, a finding of overall unconstitutionality is likely to lead to changes only in the most egregiously gerrymandered districts and some of the districts adjacent to them.
+\par
     Unlike all federal courts that have ruled on partisan gerrymandering to date, the Pennsylvania opinion does not require proof of intent to discriminate, which could make proving a violation easier under state law than under any likely federal standard.
-Moreover, although the Pennsylvania Supreme Court opinion does not require evidence of partisan effects, and emphasizes failure to comply with good government criteria, we believe that the \textit{LWV} opinion can be ``tweaked" to address what \citet{GrofmanCervas2018_WashPo} in their on-line essay in the \textit{Washington Post} refer to as \textit{stealth gerrymanders}, i.e., plans that satisfy good government redistricting criteria to a reasonable degree but are nonetheless carefully crafted invidious partisan gerrymanders in terms of their expected political impact.
+Moreover, although the Pennsylvania Supreme Court opinion does not require evidence of partisan effects, and emphasizes failure to comply with good government criteria, we believe that the \textit{LWV} opinion can be ``tweaked'' to address what \citet[][see also \citeyear{Grofman_Cervas_2018_ELJ}]{GrofmanCervas2018_WashPo} in their on-line essay in the \textit{Washington Post} refer to as \textit{stealth gerrymanders}, i.e., plans that satisfy good government redistricting criteria to a reasonable degree but are nonetheless carefully crafted invidious partisan gerrymanders in terms of their expected political impact.
 The reason for our optimism in this respect are the findings in the opinion that do address partisan effects of the Pennsylvania congressional map.
@@
 % -----------------------------------------------------------------
-        \footnote{See NYT Upshot: \href{https://nyti.ms/2w8b19k}{``Democrats Didn’t Even Dream of This Pennsylvania Map. How Did It Happen?", https://nyti.ms/2w8b19k (Nate Cohn, Feb. 21, 2018).;} \\
-        \href{https://nyti.ms/2wkpEFT}{``Hundreds of Simulated Maps Show How Well Democrats Fared in Pennsylvania", https://nyti.ms/2wkpEFT (Nate Cohn, Feb. 26, 2018).}}
+        \footnote{See NYT Upshot: \href{https://nyti.ms/2w8b19k}{``Democrats Didn’t Even Dream of This Pennsylvania Map. How Did It Happen?'', https://nyti.ms/2w8b19k (Nate Cohn, Feb. 21, 2018).;} \\
+        \href{https://nyti.ms/2wkpEFT}{``Hundreds of Simulated Maps Show How Well Democrats Fared in Pennsylvania'', https://nyti.ms/2wkpEFT (Nate Cohn, Feb. 26, 2018).}}
 % ----------------------------------------------------------------- 
@@
 \par
-    In addition to the rejected 2011 and the court-ordered 2018 map, we also address the partisan and other features of two maps offered to the Pennsylvania Court as a proposed remedy for constitutional violation found in the 2011 map: the \textit{joint legislative} plan offered by Republican legislators from both chambers, and the congressional map proposed by the Democratic governor (\textit{Gov.
+    In addition to the rejected 2011 and the court-ordered 2018 map, we also address the partisan and other traditional features of two maps offered to the Pennsylvania Court as a proposed remedy for constitutional violation found in the 2011 map: the \textit{Joint Legislative} plan offered by Republican legislators from both chambers, and the map proposed by the Democratic governor (\textit{Gov.
 Wolf}).
@@
 \par
+% =================================================================
 % •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-% -- SECTION -- SECTION -- SECTION -- SECTION -- SECTION -- SECTI %
+% █▀▀ █▀▀ █▀▀ ▀▀█▀▀ ░▀░ █▀▀█ █▀▀▄
+% ▀▀█ █▀▀ █░░ ░░█░░ ▀█▀ █░░█ █░░█
+% ▀▀▀ ▀▀▀ ▀▀▀ ░░▀░░ ▀▀▀ ▀▀▀▀ ▀░░▀
 % •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
@@
 % •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-    As noted above, to contrast the Court-ordered Remedial plan with the 2011 Enacted plan, and to make further comparisons to the two most politically important remedy plans offered to the Court, the joint plan proposed by the Republicans leaders of the two legislative chambers, and the Governor’s proposed plan as a Democratic alternative, we will look at a number of features of each of these four plans as a whole.
+    To facilitate comparisons to the two most politically important remedy plans offered to the Court, the joint plan proposed by the Republicans leaders of the two legislative chambers, and the Governor’s proposed plan as a Democratic alternative, we look at a number of features in each of the four plans as a whole.
 We do not, however, discuss district-specific criteria for identifying gerrymandering in these maps, since, although some district-specific evidence was presented about the 2011 map in the trial testimony by an expert who was highly knowledgeable about communities of interest in the state, the Pennsylvania Supreme Court opinion itself offers a \underline{statewide} standard for a finding of unconstitutional partisan gerrymandering.
@@
 % =================================================================
-    All the plans create contiguous districts.
-Similarly, all four plans appear to satisfy voting rights requirements.
-Thus, we do not discuss these two ``good government” criteria further.
     Consequently, we limit ourselves to metrics that report results for a jurisdiction as whole.
 \par
-    We begin our analyses describing map features distinct from projected (or actual) partisan outcomes or proposed measures of partisan effects.
+    All the plans create contiguous districts. Similarly, all four plans appear to satisfy voting rights requirements. Thus, we do not discuss these two ``good government'' criteria further.
+\par
+% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
+% ▀▄▀▄▀▄ SUBSECTION -- SUBSECTION -- SUBSECTION -- SUBSECTION -- ▄▀▄▀▄▀
+% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
+        \subsection*{Institutional Measures}
+% ------ ••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
+% =====================================================================
+% ▀▄▀▄▀▄ T̟A̟B̟L̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ T̟A̟B̟L̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ T̟A̟B̟L̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ T̟A̟B̟L̟E̟
+% ---------------------------------------------------------------------
+    \input{Tables/_tab_summaries.tex}
+        \begin{center}\textbf{INSERT TABLE \ref{tab:summaries} ABOUT HERE} \end{center}
+% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
+\par
+    We begin our analyses describing map features distinct from partisan outcomes or proposed measures of partisan asymmetries.
 Table \ref{tab:summaries} offers a comparison of two key good government features of the map.
+%
     \begin{itemize}
@@
     \end{itemize}
+%
 \par
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-% -- SUBSECTION -- SUBSECTION -- SUBSECTION -- SUBSECTION -- SUBS %
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-\subsection*{Measurement and Data}
+% ↤↤↤↤↤ -------------- S̟U̟B̟-S̟U̟B̟S̟E̟C̟T̟I̟O̟N̟ -------------- ↦↦↦↦↦
             \subsubsection*{County Splits}
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
+% ↤↤↤↤↤ -------------- S̟U̟B̟-S̟U̟B̟S̟E̟C̟T̟I̟O̟N̟ -------------- ↦↦↦↦↦
     Avoiding county splits is incorporated into the language of many state constitutions \citep{Grofman1985} and can be seen as a way of guarding against gerrymandering by making it more difficult to draw partisan maps.
@@
 Allegheny and Montgomery must each be split at least twice, while Philadelphia must be split at least three times.
-If no set of counties exactly equals the ideal congressional population, every districting plan requires a minimum of $n-1$ splits (where $n$ equals the number of congressional districts).
+If no set of counties exactly equals the ideal congressional population, every districting plan requires a minimum of $n-1$ splits (where $n$ equals the number of congressional districts; assuming that population equality between district must be one person or less).
 The Pennsylvania Supreme Court found that an unusually and egregiously large number of counties were split in the 2011 congressional map, and they found this fact deeply troubling.
-\par
-% =================================================================
-% -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE  % 
-% -----------------------------------------------------------------
-\input{Tables/_tab_summaries.tex}
-\begin{center} INSERT TABLE \ref{tab:summaries} ABOUT HERE \end{center}
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
 \par
@@
 The other way is to count the number of times a county is split, i.e., the number of separate pieces into which the county is divided.
-In the first counting method, a county can have portions of itself found in 18 different districts, but have this counted as only one county spit.
+In the first counting method, a county can have portions of itself found in 18 different districts, but have this counted as only one county split.
 Under the second method, this wholesale splitting of the county would be counted as 18 splits.
@@
 However, we must also be alert for divisions of a county into two or more discontiguous pieces found within the same district.
-We call this “fracking,” to provide a term similar sounding to ‘packing’ and ‘cracking.’ We find “fracking” a particularly egregious problem since there is no need for counties to be fracked, and fracking may be used as a gerrymandering tool.
+We call this ``fracking'', to provide a term similar sounding to ‘packing’ and ‘cracking.’ We find fracking a particularly egregious problem since there is no need for counties to be fracked, and fracking may be used as a gerrymandering tool.
 Instances where counties have multiple discontiguous pieces located in a single district are not discussed in the Pennsylvania Supreme Court ruling on the unconstitutionality of the 2011 Pennsylvania congressional plan.
 However, our analysis of the 2011 plan reveals several cases of fracking: two of the most egregious are districts 14 and 18 in Allegheny County (Pittsburgh).
-This concept has also been addressed by the US District Court in North Carolina (\textit{Common Cause v.
+This concept has also been addressed by the U.S.
+District Court in North Carolina (\textit{Common Cause v.
 Rucho} No 1:16-CV-1026 (U.S.
@@
 105 [p.
-194])).
-}
+194])).}
 % ----------------------------------------------------------------- 
@@
 That the 2011 map is an eyesore vis-a-vis unnecessary splits of county boundaries is revealed clearly in Figure \ref{fig:maps}.
+\par
+% =====================================================================
+% ▀▄▀▄▀▄ F̟I̟G̟U̟R̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ F̟I̟G̟U̟R̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ F̟I̟G̟U̟R̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ F̟I̟
+% ---------------------------------------------------------------------
+    \input{Figures/_fig_maps.tex}
+        \begin{center} INSERT FIGURE \ref{fig:maps} ABOUT HERE \end{center}
+% ---------------------------------------------------------------------
+% ▀▄▀▄▀▄ E͎N͎D͎ F͎I͎G͎U͎R͎E͎ ▄▀▄▀▄▀▀▄▀▄▀▄ E͎N͎D͎ F͎I͎G͎U͎R͎E͎ ▄▀▄▀▄▀▀▄▀▄▀▄ E͎N͎D͎ F͎I͎G͎U͎R͎E͎ ▄▀▄
+% ===================================================================== 
+\par
 %
-% ================================================================= 
-% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
-    \footnote{We find counties that are split into multiple pieces particularly egregious.
-``\textit{Fracking}" occurs when one district is split into two or more non-contiguous pieces of a single county.
-It was not mentioned in the \textit{LWV} decision, but too can be used as an indicia of partisan (and racial) manipulation.
-We call this ``\textit{fracking}" to parallel the terminology of other districting terms of ``\textit{cracking}" and ``\textit{packing}".
-Examples of ``fracking" in the 2011 enacted map include districts 14 \& 18 in Allegheny County (Pittsburgh).}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% =================================================================
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
+% ↤↤↤↤↤ -------------- S̟U̟B̟-S̟U̟B̟S̟E̟C̟T̟I̟O̟N̟ -------------- ↦↦↦↦↦
             \subsubsection*{Compactness}
@@
 Moreover, as many authors have shown, compactness has multiple dimensions, and these different dimensions need not move in parallel with one another; thus a plan appears compact on one dimension might appear ill-compact on another.
-The two most common ideas of compactness refer, on the one hand, to how close a legislative district’s boundaries are to its geographic center and, on the other, to how ``regular" in shape a district appears to be \citep{Niemi1990, Kaufman2018}.
+The two most common ideas of compactness refer, on the one hand, to how close a legislative district’s boundaries are to its geographic center and, on the other, to how ``regular'' in shape a district appears to be \citep{Niemi1990, Kaufman2018}.
 In Table \ref{tab:summaries} we report two well-known measures which tap respectively each of these two dimensions.
@@
     The 2011 Enacted map looks awful by both compactness measures compared to the Court-ordered map, or even compared to the two proposed remedial maps; while both the Republican and the Democratic proposed remedial maps look reasonable by both compactness measures, even though the Democratic plan is superior to the Republican plan and the court-ordered map is superior to both.
-However, until we examine the likely partisan effects of these four plans we are not in a position to conclusively rule out the possibility that one or more of the remedial plans is a ``\textit{stealth gerrymander}".
+However, until we examine the likely partisan effects of these four plans we are not in a position to conclusively rule out the possibility that one or more of the remedial plans is a ``\textit{stealth gerrymander}''.
 As is now well recognized, gerrymandering can be found even in plans with compact appearing districts, or in a plan with districts that preserve most county borders.
@@
 And, each of three metrics order the plans in good government terms from worst to best in exactly the same way, with the worst plan being the 2011 Enacted map, the next worst being the Republican remedy plan, followed by the the Democratic remedy plan, and the very best being the court-ordered map.
-Moreover, as we will see later, the same ordering of plans is reflected when we examine the partisan implications of the various maps, with the map ordered into place as a remedy by the Pennsylvania Supreme Court close to perfectly neutral in its expected partisan effects once we recognize the implications of the partisan geography in Pennsylvania in ``naturally” tilting outcomes slightly in a pro-Republican direction because of differences in the degree of geographic concentration of each party’s supporters.
-(see below) Moreover, we show that the proposed Republican remedy plan can be characterized as a ``disguised” pro-Republican partisan gerrymander -- what we call a ``stealth gerrymander” -- which is nearly as pernicious in its expected pro-Republican partisan consequences as the 2011 plan whose effects it claims to ``remedy".
+Moreover, as we will see later, the same ordering of plans is reflected when we examine the partisan implications of the various maps, with the map ordered into place as a remedy by the Pennsylvania Supreme Court close to perfectly neutral in its expected partisan effects once we recognize the implications of the partisan geography in Pennsylvania in ``naturally'' tilting outcomes slightly in a pro-Republican direction because of differences in the degree of geographic concentration of each party’s supporters (see below).
+Moreover, we show that the proposed Republican remedy plan can be characterized as a ``disguised'' pro-Republican partisan gerrymander -- what we call a ``stealth gerrymander'' -- which is nearly as pernicious in its expected pro-Republican partisan consequences as the 2011 plan whose effects it claims to ``remedy''.
 \par
-    To give a more intuitive feel for how egregiously ill-compact the 2011 map is compared to the other alternatives we are considering, we show all four maps in the same scale, in a gray and white format.
-What is visually apparent from these maps is how aesthetically ``ugly" the 2011 map is compared to all of its alternatives.
+    To give a more intuitive feel for how egregiously ill-compact the 2011 map is compared to the other alternatives we are considering, Figure \ref{fig:maps} shows all four maps in the same scale, in a gray and white format.
+What is visually apparent from these maps is how aesthetically ``ugly'' the 2011 map is compared to all of its alternatives.
 But it also clear that three remedy maps that are similar in having relatively few county cuts nonetheless can look quite different in terms of how each configures districts.
@@
 %
+%
+% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
+%     ↤↤↤↤↤ -------------- S̟U̟B̟-S̟U̟B̟S̟E̟C̟T̟I̟O̟N̟ -------------- ↦↦↦↦↦
+            \subsection*{Asymmetry Measures}
+% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
+    We emphasize that a partisan gerrymandering claim based on statewide political consequences should also provide evidence that the (a) gerrymandering indicators that are found cannot be explained simply by the geographic distribution of each party’s electoral support (e.g., differential concentration of electoral support can lead to what has been called a ``natural gerrymander''), and (b) that the partisan asymmetry in translating votes into seats is not due to chance alone (e.g., many highly competitive seats won by very narrow margins) but can be expected to be durable.
+With only 18 districts, sensitivity to chance effects is particularly important.
+\par
+    Rebuttal to any claims that the partisan asymmetries observed in the consistent outcomes of the 2011 map can be attributed primarily to differential concentration of party voting strengths, or to chance, is found in the expert witness testimony in the case (see esp.
+the simulations of potential good government maps done by Professor Chen, discussed in detail in the Pennsylvania Supreme Court opinion in the case, and on which the Court placed great reliance).
+With respect to the durability of the effects of the 2011 map, we have already pointed out that this plan yielded unchanged 13R-5D results for three elections in a row.
+To help explain this over time consistency we note that the 2011 Map has primarily safe seats (13 of 18 won with greater than a 10 percentage point margin).
+But, indicative of the pro-Republican gerrymander achieved with the 2011 map, those safe seats are distributed in a highly asymmetric fashion, with nine essentially safe seats for Republicans and only four for Democrats.
 %        =================================================================
-% -- FIGURE -- FIGURE -- FIGURE -- FIGURE -- FIGURE -- FIGURE --   % 
+%        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
+% ------------------------------------------------------------------------
+        \footnote{However, we should note that some authors \citep[e.g.,][]{Grofman2019_ELJ} have argued that a partisan gerrymandering claim at a jurisdiction-wide level must also be reinforced by district specific evidence of manipulation of particular district boundaries.
+As noted earlier, we do not discuss this type of district specific evidence here, other than to point out that such evidence was presented at the trial, and that this evidence was used to support the Court’s finding that the Pennsylvania congressional map was an unconstitutional gerrymander under Pennsylvania state law.}
+% ------------------------------------------------------------------------
+%        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
+%        =================================================================
+%
+\par
+    There are several metrics that have been proposed to look at the degree to which there is asymmetry in seats-votes relationships that might be indicative of partisan gerrymandering: (1) \textit{partisan bias} \citep{Tufte1973}, (2) the \textit{mean-minus-median} gap \citep{Mcdonald_Best_2015_ELJ} (3) the \textit{efficiency gap} \citep{Stephanopoulos2014_UofChicagoLaw}, and most recently (4) \textit{declination} \citep{Warrington2018} (details below).
+Each of these measures can be directly calculated for the 2011 map based on district election outcomes in 2012, 2014, and 2016.
+% ================================================================= 
+% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
 % -----------------------------------------------------------------
-% Maps Of The Proposed Pennsylvania Congressional Districts
- \begin{center} INSERT FIGURE \ref{fig:maps} ABOUT HERE \end{center}
-        \
+        \footnote{Another recently proposed measure is a simplified version of partisan bias which lacks the advantage of separately calculating bias and responsiveness (swing ratio).
+It is calculated by taking the average two-party vote for the dominant party and adjusting it uniformly downward in each district so that the two parties receive exactly 50\% of the overall vote; then we look to see how far the seat share the dominant party now receives is from 50\% \citep{Wang2016_SLR}.}
 % ----------------------------------------------------------------- 
-% -- END FIGURE -- END FIGURE -- END FIGURE -- END FIGURE -- END FI %
+% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
+% =================================================================
+    \underline{Positive} values favor Republicans, while \underline{negative} values favor Democrats. 
+\par
+    We now define these four measures of gerrymandering in which we will evaluate the four competing maps.
+%
+% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
+%     ↤↤↤↤↤ -------------- S̟U̟B̟-S̟U̟B̟S̟E̟C̟T̟I̟O̟N̟ -------------- ↦↦↦↦↦
+            \subsubsection*{Partisan Bias}
+% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
+%
+    \textit{Partisan Bias} indicates asymmetry in the translation of each party’s votes into seats.
+Customarily, partisan bias is measured at the (hypothetical) point where each party’s share of the two-party vote is exactly 50\%.
+At that point, both parties should, by symmetry, receive identical seats shares.
+Thus, the difference between a party’s projected seat-share with 50\% of the vote and exactly 50\% can be taken as a measure of partisan bias.
+Note, however, that although a plan must be proportional at the ($50,50$) point, a partisan bias of zero does \underline{not} imply proportional representational.
+What partisan bias taps is the differential treatment of parties.
+Thus, for example, if party R gets 61\% of the seats with 52\% of the vote, this is not a sign of bias as long as party L also can be expected to get 61\% of the seats were it to have 52\% of the vote.
+% ================================================================= 
+% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
+% -----------------------------------------------------------------
+        \footnote{Although the basic idea goes back at least as far as \citet{Dahl1956}, the modern \textit{locus classicus} in political science for the idea of partisan bias (and the parallel concept of responsiveness/swing ratio) is \citet{Tufte1973}, with important recent work done by Gary King (see e.g., \citep{GelmanKing1994_unifiedAJPS}; and reviewed in \citet{GrofmanKing2007_ELJ}).
+A closely related line of research is found in the political geography literature (with \textit{locus classicus} \citet{Brookes1959, Brookes1960} -- with some further methodological improvements by scholars such as \citet{Johnston1994, Rossiter1997}, and \citet{Johnston2002}.}
+% ----------------------------------------------------------------- 
+% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
 % =================================================================
 \par
-    Now we turn to a consideration of probable partisan consequences of the four maps. We begin with a discussion of the 2011 map.
+    There are a number of different ways to estimate partisan bias based on the shape of the votes to seats distribution \citep{Tufte1973, Grofman1983, Browning_King_1987_seats_votes, GelmanKing1994_unifiedAJPS, Grofman_et_al_1997_SwingRatio_Bias, Zingher2016_bias_swingratio_JEPP}.
+The method we use here is that of \citet{Tufte1973}.
+Hypothetical elections are constructed by incrementally adding (or subtracting) one percentage point (on average) at a time to find aggregate seat outcomes under differing mean vote shares.
+The resulting ($s$, $v$) points on the votes-to-seats curve are then converted to a log odds form by using $ log(\frac{s}{1-s}) $ as the dependent variable, and $ log(\frac{v}{1-v}) $ as the independent variable.
+A regression on the transformed variables is then run, with transformed $s$ as the dependent variable, for $v$ values in some range near a 50\% vote share.
+% ================================================================= 
+% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
+% -----------------------------------------------------------------    
+        \footnote{Where to center the \textit{seats/votes} curve is an unresolved issue in the literature.
+In \citet{GelmanKing1994_unifiedAJPS} \texttt{JudgeIt} program, the choice is to center the election at the average Democratic vote share over recent elections.
+\citet{Kastellec_et_al_2008_PS} similarly simulates data for average vote shares between 45\% and 55\%.
+We checked for consistency between the methods by running the numbers in three ways: centering the simulated outcomes at an average election vote share, centering the simulated outcomes at a tied 50/50 election, and centering the simulated outcomes at the actual election results in the particular election.
+All three variations provide similar estimates, as do the results from \texttt{JudgeIt}.
+Readers should note that, with only 18 districts, too much weight should not be placed on the exact bias coefficients reported [see the discussion in \citet{Browning_King_1987_seats_votes}]; but the relative magnitude of the bias values across the different plans is fully congruent with the evaluations of probable partisan gerrymandering yielded by the other measures.
+Because of the limited number of districts in the plans, the same caution is urged for all the measures of gerrymandering reported in this essay.
+We have devoted more space to partisan bias than to the other measures of gerrymandering reported in this essay due to it's complexity than to the other more straightforward measures.}
+% ----------------------------------------------------------------- 
+% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
+% =================================================================
+    \textit{Partisan Bias} is then calculated from the intercept of this regression using an exponential transformation (for details see \citet{Grofman1983}).
+Partisan bias is calculated from the Republican perspective, and \underline{negative} values indicates bias against Republicans.
 \par
+%     ↤↤↤↤↤ -------------- S̟U̟B̟-S̟U̟B̟S̟E̟C̟T̟I̟O̟N̟ -------------- ↦↦↦↦↦
+            \subsubsection*{Mean-Median Gap}
 % •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-\subsubsection*{Partisan results in actual elections held under the 2011 map}
+    When districts are sorted according to two-party vote share, the \textit{Mean-Median Gap} is the difference between the average vote percentage for a party and its vote share in the median district \citep[see][]{Best2018}.
+This measure is a variant of \textit{skewness}.
+When the mean is substantially higher or lower than the median, this is taken to be indicative of partisan bias.
+\par
+%     ↤↤↤↤↤ -------------- S̟U̟B̟-S̟U̟B̟S̟E̟C̟T̟I̟O̟N̟ -------------- ↦↦↦↦↦
+            \subsubsection*{Efficiency Gap}
+% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
+    The \textit{Efficiency Gap} is calculated as defined in \citet{Stephanopoulos2014_UofChicagoLaw}, where all the party’s votes are wasted if they lose the district, and all the winner’s votes over 50\% are wasted.
+The difference between each party’s wasted votes is then divided by the total votes cast to produce the \textit{efficiency gap}, with a value of zero denoting what is regarded as ideal.
+As noted in \citet[][p.
+13]{Best2018}, this is equivalent to taking an aggregate responsiveness (swing ratio) of two as ideal.
+\par
+%     ↤↤↤↤↤ -------------- S̟U̟B̟-S̟U̟B̟S̟E̟C̟T̟I̟O̟N̟ -------------- ↦↦↦↦↦
+            \subsubsection*{Declination}
+% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
+    Among the more recent measures of gerrymandering is the \textit{Declination}.
+\textit{Declination} is an astronomical term referring to the angle on a compass, with a Latin root meaning ``a bending away''.
+For gerrymandering, it uses the angles created from ordering all of the districts by vote share, computing the mean vote-share for each party separately for the seats they won, and then comparing the differences in their distribution.
+To compare, a line is draw from the mean vote-share to the 0.5 line separately for each party.
+The angle formed from the different slopes of the line is called the \textit{declination}.
+An angle indicates whether a plan is a gerrymandering, under the logic that vote-shares should be distributed in some uniform manner across districts.
+If partisan are \textit{packed}, the angle will be greater.
+All values range between -1 and 1.
+Though we have defined the vote-share in terms of Republican outcomes, we adjusted the values such that \underline{negative} values are favorable to the Democrats.
+Further information on declination can be found in \citet{Warrington2018}.
+\par
+    In our view, none of these four measures standing alone can be taken to be proof of gerrymandering.
+Rather we see them as potentially reinforcing indicators.
+Our own analyses have found hypotheticals where these measures can generate either a false positive or a false negative, but we have also found that consistent bias in the same direction is unlikely when we use multiple indicators (data omitted for space reasons).
+We will return to a discussion of the results later in this essay.
+Here we point out, however, all four indicators point clearly and overwhelmingly in the same direction vis-à-vis the 2011 Enacted plan -- namely that it is a blatant pro-Republican partisan gerrymander.
+\par
+% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
+% ▀▄▀▄▀▄ SUBSECTION -- SUBSECTION -- SUBSECTION -- SUBSECTION -- ▄▀▄▀▄▀
+% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
+    \subsection*{Evaluation of the 2011 enacted map}
+% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
+%
+%     ↤↤↤↤↤ -------------- S̟U̟B̟-S̟U̟B̟S̟E̟C̟T̟I̟O̟N̟ -------------- ↦↦↦↦↦
+        \subsubsection*{Partisan results in actual congressional elections held under the 2011 map}
 % •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
@@
         \footnote{For consistency, we report the average vote share after imputing 0.25 and 0.75 for uncontested districts.
-In raw votes, Democrats won more votes than Republicans in 2012, but by taking the imputed average, it shows they won a minority of votes due to disproportionate uncontested districts.
-The Republican raw vote shares are 49.2\% in 2012, 55.5\% in 2014, and 54.1\% in 2016.}
+In raw votes, Democrats won more votes than Republicans in 2012, and the difference is an artifact of the plan's `packing' and `cracking'.
+The Democratic raw vote shares are 50.8\% in 2012, 44.5\% in 2014, and 45.9\% in 2016.}
 % ----------------------------------------------------------------- 
@@
     Thus, the discrepancy between Republican vote share and Republican seat share was huge, and yet the seat share appeared to be immutable.
+Table \ref{tab:congsum} shows the indicators of statewide partisan gerrymandering effects that provides additional compelling evidence of the 2011 map as an (enduring) partisan gerrymander.
 \par
+% =====================================================================
+% ▀▄▀▄▀▄ T̟A̟B̟L̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ T̟A̟B̟L̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ T̟A̟B̟L̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ T̟A̟B̟L̟E̟
+% ---------------------------------------------------------------------
+    \input{Tables/_tab_congsum.tex}
+        \begin{center}\textbf{INSERT TABLE \ref{tab:congsum} ABOUT HERE} \end{center}
 % •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
 %
-\subsection*{Asymmetry measures for partisan outcomes in the 2011 Plan based on statewide election outcomes}
+\par
+% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
+% ▀▄▀▄▀▄ SUBSECTION -- SUBSECTION -- SUBSECTION -- SUBSECTION -- ▄▀▄▀▄▀
+% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
+    \subsection*{Evaluating district plans using Statewide elections to provide a normal vote baseline}
+% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
 %
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-%
-    There are other indicators of statewide partisan gerrymandering effects that also provide a compelling picture of the 2011 map as an (enduring) partisan gerrymander.
-There are several metrics that have been proposed to look at the degree to which there is asymmetry in seats-votes relationships that might be indicative of partisan gerrymandering: (1) \textit{partisan bias} \citep{Tufte1973}, (2) the \textit{mean-minus-median} gap \citep{Mcdonald_Best_2015_ELJ} (3) the \textit{efficiency gap} \citep{Stephanopoulos2014_UofChicagoLaw}, and most recently (4) declination \citep{Warrington2018}.
-Each of these measures can be directly calculated for the 2011 map based on district election outcomes.
-% ================================================================= 
-% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
-    \footnote{Another recently proposed measure is a simplified version of partisan bias which lacks the advantage of separately calculating bias and responsiveness (swing ratio).
-It is calculated by taking the average two-party vote for the dominant party and adjusting it uniformly downward in each district so that the two parties receive exactly 50\% of the overall vote; then we look to see how far the seat share the dominant party now receives is from 50\% \citep{Wang2016_SLR}.}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% =================================================================
-\par
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-\subsubsection*{Partisan Bias}
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-%
-    \textit{Partisan Bias} indicates asymmetry in the translation of each party’s votes into seats.
-Customarily, partisan bias is measured at the (hypothetical) point where each party’s share of the two-party vote is exactly 50\%.
-At that point, both parties should, by symmetry, receive identical seats shares.
-Thus, the difference between a party’s projected seat-share with 50\% of the vote and exactly 50\% can be taken as a measure of partisan bias.
-Note, however, that although a plan must be proportional at the ($50,50$) point, a partisan bias of zero does not imply proportional representational.
-What partisan bias taps is differential treatment of the parties?
-Thus, for example, if party R gets 56\% of the seats with 52\% of the vote, this is not a sign of bias as long as party L also can be expected to get 56\% of the seats were it to have 52\% of the vote.
-% ================================================================= 
-% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
-    \footnote{Although the basic idea goes back at least as far as \citet{Dahl1956}, the modern \textit{locus classicus} in political science for the idea of partisan bias(and the parallel concept of responsiveness/swing ratio) is \citet{Tufte1973}, with important recent work done by Gary King (see e.g., \citep{GelmanKing1994_unifiedAJPS}; and review in \citet{GrofmanKing2007}).
-A closely related line of research is found in the political geography literature (with \textit{locus classicus} \citet{Brookes1959, Brookes1960} -- with some further methodological improvements by scholars such as \citet{Johnston1994, Rossiter1997} and \citet{Johnston2002}.}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% =================================================================
-\par
-    There are a number of different ways to estimate partisan bias based on the shape of the votes to seats distribution \citep{Tufte1973, Grofman1983, Browning_King_1987_seats_votes, GelmanKing1994_unifiedAJPS, Grofman_et_al_1997_SwingRatio_Bias, Zingher2016_bias_swingratio_JEPP}.
-The method we use here is basically that of \citet{Tufte1973} with some additional complexity building in ideas of \citet{GelmanKing1994_unifiedAJPS} on simulating potential outcomes in a probabilistic fashion.
-We adapted the code from \citet{Kastellec_et_al_2008_PS}, which parallels that used by \texttt{JudgeIt}.
-It models inter-election shifts and adjusts for differences over time in vote shares among and between districts in a probabilistic fashion, using a simulation with 1000 repetitions (see more below).
-We then apply the following procedure to find the mean partisan bias and range of expected bias over the course of the decade.
-\par
-    Hypothetical elections are constructed by incrementally adding (or subtracting) one percentage point (on average) at a time to find aggregate seat outcomes under differing mean vote shares.
-The resulting ($s$, $v$) points on the votes-to-seats curve are then converted to a log odds form by using $ log(\frac{s}{1-s}) $ as the dependent variable, and $ log(\frac{v}{1-v}) $ as the independent variable.
-A regression on the transformed variables is then run, with transformed $s$ as the dependent variable, for $v$ values in some range near a 50\% vote share.
-% ================================================================= 
-% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------    
-    \footnote{Where to center the \textit{seats/votes} curve is an unresolved issue in the literature.
-In \citet{GelmanKing1994_unifiedAJPS} \texttt{JudgeIt} program, the choice is to center the election at the average Democratic vote share over recent elections.
-\citet{Kastellec_et_al_2008_PS}, similarly, simulates data for average vote shares between 45\% and 55\%.
-We checked for consistency between the methods by running the numbers in three ways: centering the simulated outcomes at an average election vote share, centering the simulated outcomes at a tied 50/50 election, and centering the simulated outcomes at the actual election results in the particular election.
-All three variations provide similar estimates, as do the results from \texttt{JudgeIt}.
-Readers should note that with only 18 districts, too much weight should not be place exact bias coefficients reported [see the discussion in \citet{Browning_King_1987_seats_votes}]; but the relative magnitude of the bias values across the different plans is fully congruent with the evaluations of probable partisan gerrymandering yielded by the other measures.
-Because of the limited number of districts in the plans, the same caution is urged for all the measures of gerrymandering reported in this essay.}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% =================================================================
-    \textit{Partisan Bias} is then calculated from the intercept of this regression using an exponential transformation (for details see \citet{Grofman1983}).
-Partisan bias is calculated from the Republican perspective, and \underline{negative} values indicates bias against Republicans.
-\par
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-        \subsubsection*{Mean-Median Gap}
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-    When districts are sorted according to two-party vote share, the \textit{Mean-Median Gap} is the difference between the average vote percentage for a party and its vote share in the median district \citep[see][]{Best2018}.
-This measure is a variant of \textit{skewness}.
-When the mean is substantially higher or lower than the median, this is taken to be indicative of partisan bias.
-\par
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-        \subsubsection*{Efficiency Gap}
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-    The \textit{Efficiency Gap} is calculated as defined in \citet{Stephanopoulos2014_UofChicagoLaw}, where all the party’s votes are wasted if they lose the district, and all the winner’s votes over 50\% are wasted.
-The difference between each party’s wasted votes is then divided by the total votes cast to produce the \textit{Efficiency Gap}, with a value of zero denoting what is regarded as ideal.
-As noted in \citet[][p.
-13]{Best2018}, this is equivalent to taking an aggregate responsiveness (swing ratio) of two as ideal.
-\par
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-        \subsubsection*{Declination}
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-    Among the more recent measures of gerrymandering is the declination.
-Declination is an astronomical term referring to the angle on a compass, with a Latin root meaning ``a bending away".
-For gerrymandering, it uses the angles created from ordering all of the districts by vote share, computing the mean vote-share for each party separately for the seats they won, and then comparing the differences in their distribution.
-To compare, a line is draw from the mean vote-share to the 0.5 line separately for each party.
-The angle formed from the different slopes of the line is called the declination.
-An angle indicates whether a plan is a gerrymandering, under the logic that vote-shares should be distributed in some uniform manner across districts.
-If partisan are \textit{packed}, the angle will be greater.
-All values range between -1 and 1.
-Though we have defined the vote-share in terms of Republican outcomes, we adjusted the values such that \underline{negative} values are favorable to the Democrats.
-Further information can be found in \citet{Warrington2018}.
-\par
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-        \subsubsection*{Measuring Gerrymandering}
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-    We emphasize that a partisan gerrymandering claim based on political consequences should also provide evidence that the (a) gerrymandering indicators that are found cannot be explained simply by the geographic distribution of each party’s electoral support (e.g., differential concentration of electoral support can lead to what has been called a ``natural gerrymander"), and (b) that the partisan asymmetry in translating votes into seats is not due to chance alone (e.g., many highly competitive seats won by very narrow margins) but can be expected to be durable.
-With only 18 districts, sensitivity to chance effects is particularly important.
-\par
-    Rebuttal to any claims that the partisan asymmetries observed in the consistent outcomes of the 2011 map can be attributed primarily to differential concentration of party voting strengths, or to chance, have been decisively refuted by the expert witness testimony in the case (see esp.
-the simulations of potential good government maps done by Professor Chen, discussed in detail in the Pennsylvania Supreme Court opinion in the case, and on which the Court placed great reliance).
-With respect to the durability of the effects of the 2011 map, we have already pointed out that this plan yielded unchanged 13R-5D results for three elections in a row.
-To help explain this over time consistency we note that the 2011 Map has primarily safe seats (13 of 18 won with greater than a 10\% margin).
-But, indicative of the pro-Republican gerrymander achieved with the 2011 map, those safe seats are distributed in a highly asymmetric fashion, with nine essentially safe seats for Republicans and only four for Democrats.
-%        ================================================================= 
-%        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% ------------------------------------------------------------------------
-        \footnote{However, we should note that some authors \citep[e.g.,][]{Grofman2019_ELJ} have argued that a partisan gerrymandering claim at a jurisdiction-wide level must also be reinforced by district specific evidence of manipulation of particular district boundaries.
-As noted earlier, we do not discuss this type of district specific evidence here, other than to point out that such evidence was presented at the trial, and that this evidence was used to support the Court’s finding that the Pennsylvania congressional map was an unconstitutional gerrymander under Pennsylvania state law.}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% =================================================================
-\par
-    We show in Table \ref{tab:congsum} the results for the 2011 plan found unconstitutional.
-Results of the four measures for the 2011 plan are based on using the average of actual congressional results for the 2012-16 period under that plan.
-\underline{Positive} values favor Republicans (add to their seat share), while \underline{negative} values favor Democrats (subtract from Republican seat share).
-\par
-    In our view, none of these four measures standing alone can be taken to be proof of gerrymandering.
-Rather we see them as potentially reinforcing indicators.
-Our own analyses have found hypotheticals where these measures can generate either a false positive or a false negative, but we have also found that consistent bias in the same direction is unlikely when we use multiple indicators (data omitted for space reasons).
-Here, however, all four indicators, like the actual election results, point clearly and overwhelmingly in the same direction vis-à-vis the 2011 Enacted plan -- namely that it is a blatant pro-Republican partisan gerrymander.
-%
-% =================================================================
-\subsection*{Evaluating Partisan Gerrymandering using Statewide elections to provide a normal vote baseline}
-% =================================================================
-%
-    In order to assess expected partisan consequences for plans that have not yet been implemented we need to develop a predictive equation, and that is most readily done with data on partisan outcomes for state-wide elections projected into the new districts.
+    In order to assess expected partisan consequences for plans that have not yet been implemented we need to develop a predictive equation, and one such way is with data on partisan outcomes for state-wide elections projected into the new districts.
 For example, Professor Jowei Chen in his testimony in the Pennsylvania case used a composite of the six state-wide elections between 2008 and 2010 to estimate the political consequences of alternative maps.
@@
 % ------------------------------------------------------------------------
-        \footnote{We compiled the composite measure using both the sum of the raw votes by party and an average of the two-party vote in the five races at the district level.
-Both ways result in the same district partisanship breakdowns in all plans.}
+    \footnote{We report the Republican share of the two-party vote ($\delta_{i}$).
+Districting plans are represented by $\mathcal D $, (e.g., $\mathcal D_{enacted}$, $\mathcal D_{remedial}$, \dots, $\mathcal D_{j}$), and each election in year $ y $ has a district vote distribution $ \Delta_{y} = [\delta_{y1}, \delta_{y2}, \dots, \delta_{yi}] $.
+To find the overall state-wide vote share, we calculate the average district vote share, $ \bar{\Delta}_{\mathcal D y} = \frac{1}{n}\sum\limits_{i=1}^{n} \delta_{yi} $.
+By averaging, we reduce the influence of turnout variation among districts.
+This district-specific average is a useful state-wide estimate of voter sentiment though it is in part effected by the particular configuration of districts being examined \citep[see e.g.][cf.
+\cite{Grofman_et_al_1997__IntedgratedPerspective_ES}]{Kastellec_et_al_2008_PS}.
+In particular, incumbency advantage effects affect our ability to accurately predict vote outcomes.
+In ancillary work not reported here, we have looked at changes in congressional incumbency advantage in Pennsylvania over multiple decades \citep[cf.][]{Jacobson2015}.
+We compiled the composite measure using both the sum of the raw votes by party and an average of the two-party vote in the five races at the district level (see discussion on alternative ways to tabulate vote-shares in \citet{Grofman_et_al_1997__IntedgratedPerspective_ES}.
+Both ways result in the same district partisanship outcomes in all plans.}
 % ----------------------------------------------------------------- 
@@
 \par
-    We use these five elections to calculate the same four previously introduced metrics but now for all four plans.
+    We need to check that our five-election composite offers a plausible basis for a prediction model that accounts for the idiosyncrasies inherent to congressional districts.
+Encouragingly, our five-election composite results is the same 13R-5D result that the 2016 congressional races delivered.
+Additionally, each district’s outcome is identical to the 2016 congressional district results, when we uniformly adjust in each district the projected vote based on the composite state-wide elections (the ``normal'' vote, without  congressional incumbency effects) by an amount equaling the difference between the statewide legislative vote and the the statewide congressional vote.
+\par
+    When we look at the five election composite for the 2018 court remedial plan, we need to (a) adjust it for the actual statewide congressional Democratic vote share in 2018, and (b) to take into account this differential incumbency advantage that still benefits Republicans.
+Without doing so, the outcomes of the Court-ordered plan might have a wide range, with anywhere from a 5R to an 11R outcome possible because of the high number of competitive seats.
+However, when we take both factors into account by uniformly shifting the state-wide composite to reflect the 2018 statewide two-party congressional vote, and by adding in congressional incumbency advantage, which we have estimated in the 2012-2016 period to be on the order of magnitude of 4.3\% (analyses omitted for space reasons), we get a 9R-9D outcome as the most likely outcome in 2018 under the court-ordered map in our predictive model.
+Thus, our confidence in the five election projection method is further buttressed when we applied this method to predict the results of the 2018 election under the court-ordered remedial map in the light of the actual statewide aggregate results in 2018, and taking incumbency effects into account.
+In the congressional elections of 2018, the raw Republican two-party vote share in 2018 was 44.8\%, which represented a Democratic advantage, but not enough to overcome the twin barriers to a Democratic majority of over-concentration of Democratic support in urban areas and continued Republican incumbency advantage.
+% ================================================================= 
+% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --- %
+% -----------------------------------------------------------------     
+        \footnote{Six Republicans and five Democrats ran for re-election after the new map was put in place.
+No incumbent lost.
+There were two Republican incumbents who won by less than 2 percentage points, along with one who won by less than 3 percentage points.
+There was just one seat won by Democrats in 2018 with a margin of 5 percentage points or less.
+Republicans therefore had four incumbents running and winning in competitive districts compared to just two for Democrats (10+ percentage point margin).}
+% ----------------------------------------------------------------- 
+% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
+% =================================================================
+    We are not claiming to predict vote shares in advance of an election.
+We also recognize that the composite of state-wide elections is intended to provide a ``normal vote'' baseline that must be ``corrected'' with incumbency effects if we wish to actually predict outcomes.
+\par
+    We must distinguish our ability to assess asymmetry in expected seats-vote relationships based on the ``normal vote'' distribution of party support across districts -- and thus to assess the degree to which a plan is a gerrymander -- from our ability to predict actual election outcomes.
+The more competitive districts there are in a map, the more results depend strongly on electoral tides, and the greater the potential for winning vote margins in some districts to be small and determined by factors idiosyncratic to a district and by incumbency advantage.
+The 2011 map has 13 non-competitive districts (10+ percentage point margin) with a clear 9-4 Republican advantage, thus virtually guaranteeing continued Republican dominance regardless of aggregate two party vote share.
+The Court drawn map, comparatively, had only 8 non-competitive districts, with a balance of 3-R and 5-D.
+Which party will control the delegation under a highly competitive plan such as the Court-ordered plan depends almost entirely on the state-wide vote share and on incumbency advantage, along with idiosyncratic features of the competitive districts that might be small in magnitude but still matter in highly competitive seats.
+\par
+    We wish to do a further robustness check on the plausibility of using our five-election composite to calculate the various metrics designed to measure potential gerrymandering.
+One way to do this is to see how well the composite (calculated at a hypothetical 50\% vote share) mimics the results of the values for the four metrics for the 2011 map that we previously obtained using the actual congressional election results for that map.
+What we are doing is comparing a hypothetical ``normal vote'' \citep{Converse1966} assumption (based on a statewide vote in which we are not taking incumbency effects into account) with the average of what we actually get in three different congressional elections.
+The congressional results for 2012-2016 and an average of the three, reported in Table \ref{tab:congsum}, are 0.11 for partisan bias, 0.16 for the efficiency gap, 0.06 for the value of the mean minus median gap, and 0.40 for \textit{declination}.
+The comparable results from Table \ref{tab:gerry} (see below) are 0.09, 0.24, 0.05, and 0.39.
+It is apparent how close these two sets of values are to one other even though one is based on actual elections with a mean Republican (two-party) vote share close to 49.1\% while the other is based on projected election results set to yield a 50-50 two-party vote distribution.
+If instead we calculate these four metrics using the composite five election statewide vote in these elections over the period 2012-2016, a Democratic vote share of 49.1\% we get virtually identical values: 0.09, 0.24, 0.05, 0.36.
 %        =================================================================
@@
 % ------------------------------------------------------------------------
-        \footnote{We report the Republican share of the two-party vote ($\delta_{i}$).
-Districting plans are represented by $\mathcal D $, (e.g., $\mathcal D_{enacted}$, $\mathcal D_{remedial}$, \dots, $\mathcal D_{j}$), and each election in year $ y $ has a district vote distribution $ \Delta_{y} = [\delta_{y1}, \delta_{y2}, \dots, \delta_{yi}] $.
-To find the overall state-wide vote, we calculate the average district vote share, $ \bar{\Delta}_{\mathcal D y} = \frac{1}{n}\sum\limits_{i=1}^{n} \delta_{yi} $.
-By averaging, we reduce the influence of turnout variation among districts.
-This district-specific average is a useful state-wide estimate of voter sentiment though it is in part effected by the particular configuration of districts being examined \citep[see e.g.][cf.
-\cite{Grofman_et_al_1997__IntedgratedPerspective_ES}]{Kastellec_et_al_2008_PS}.
-In particular, incumbency advantage effects affect our ability to accurately predict vote outcomes.
-In ancillary work not reported here, we have looked at changes in congressional incumbency advantage in Pennsylvania over multiple decades.}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% =================================================================
-    We aggregate voting district level (precincts) data to the congressional district level.
-%        ================================================================= 
-%        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% ------------------------------------------------------------------------
-		\footnote{We wrote a function in \texttt{R} to assign all VTDs to a district based on the centroid of the VTD.
-This method creates a slight difference between it and the actual vote, but averages just -0.06\% per district.
-Most of the mis-allocation is between two adjacent districts, District 1 and District 13, both overwhelmingly Democratic districts.}
-% ------------------------------------------------------------------------ 
-%        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-%        =================================================================
-
-\par
-    \begin{center}    \textbf{US House Election Summaries (PA 2004-2016)} \par \end{center}
-    \begin{center}   \textbf{[ INSERT TABLE \ref{tab:congsum} ABOUT HERE ]} \end{center}
-% ••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••• 
-\par
-    We do need to check that our five-election composite offers a plausible basis for a prediction model once we take into account electoral tides. 
-%        ================================================================= 
-%        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% ------------------------------------------------------------------------
-		\footnote{We can only validate our measure for the 2016 election and only for the 2011 enacted plan.
-For all the remedial plans we are analyzing, the composite is our ``best guess".
-We have to make some assumptions about state and district inter-election swings.
-For simplicity, we assume that all districts have a national shift equal to the state-level swing, ie, difference from the 2016 elections.
-Later in the paper, we conduct simulated elections that vary the district level variability.}
-% ------------------------------------------------------------------------ 
-%        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-%        =================================================================
-    Encouragingly, our five-election composite results is the same 13R-5D result that the 2016 congressional races delivered.  
-%        ================================================================= 
-%        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% ------------------------------------------------------------------------
-		\footnote{Whether we take the sum of the five elections or average the five elections, both methods deliver the same relative distributions.}
-% ------------------------------------------------------------------------ 
-%        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-%        ================================================================= 
-    Additionally, each district's outcome is identical to the 2016 congressional district results. 
-\par
-    But we do wish to do a further robustness check on the plausibility of using our five election composite to calculate the various metrics designed to measure potential gerrymandering.
-One way to do this is to see how well the composite (calculated at a hypothetical 50\% vote share) mimics the results of the values for the four metrics for the 2011 map that we previously obtained using the actual congressional election results for that map.
-What we are doing is comparing a hypothetical ``normal vote” \citep{Converse1966} assumption (based on a statewide vote in which we are not taking incumbency effects into account) with the average of what we actually get in three different congressional elections.
-The results in Table \ref{tab:congsum} are 0.11 for bias, 0.16 for the efficiency gap, 0.06 for the value of the mean minus median gap, and 0.40 for declination.
-The comparable results from Table \ref{tab:gerry} are 0.09, 0.25, 0.05, and 0.40.
-It is apparent how close these two sets of values are to one other even though one is based on actual elections with a mean Republican (two-party) vote share close to 53\% while the other is based on projected election results set to yield a 50-50 two-part vote distribution.
-If instead we calculate these four metrics using the average five election statewide vote in these elections over the period 2012-2016, a Democratic vote share of 49.1\% we get virtually identical values: 0.09, 0.24,  0.05, 0.36.
-%        ================================================================= 
-%        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% ------------------------------------------------------------------------
-		\footnote{If we center the simulations at the 2012-2016 average statewide Republican vote of 53.3\% of the vote, we again get remarkably consistent estimates [0.09, 0.19, 0.05, 0.46 ].}
-% ------------------------------------------------------------------------ 
-%        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-%        =================================================================    
-\par
-    Our confidence in the five election projection method is further buttressed because, when we applied this method to predict the results of the 2018 election under the court-ordered remedial map.
-Although the method had been developed prior to the 2018 election, using data no more recent than 2016, and not including any incumbency advantage, it predicted that the court-ordered remedial plan would have produced an outcome of 7R-11D.
-%        ================================================================= 
-%        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% ------------------------------------------------------------------------
-        \footnote{After uniformly shifting each district's composite such that the statewide average equalled the actual 2018 statewide average.
-The large range is indicative of the number of competitive districts that Republican candidates won.
-Had we modeled incumbency advantage instead of the `normal vote', the two Republican incumbents who won re-election by less than 2 percentage points, along with the one who won by less than 3 percentage points would have made the actual 9R-9D outcome the likely predicted outcome.}
+        \footnote{If we center the simulations at the 2016 statewide Republican vote of 53.3\% of the vote, we again get remarkably consistent estimates [0.09, 0.19, 0.05, 0.46].}
 % ------------------------------------------------------------------------
@@
 % •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
+\par
+    The findings from Table \ref{tab:gerry} are very clear; regardless of which metric we examine, the 2011 map is the most gerrymandered.
+%        =================================================================
+%        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
+% ------------------------------------------------------------------------
+        \footnote{We use the five-election composite to calculate all four plans.
+We aggregate voting district level (precincts) data for these five statewide elections and project them to the three maps in which we are comparing the 2011 enacted map.}
+% ------------------------------------------------------------------------
+%        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
+%        =================================================================
 %
+    Not surprisingly, the nature of the bias is in a pro-Republican direction.
+Given the cumulative weight of all the evidence, the 2011 congressional map was clearly an egregious pro-Republican gerrymander.
+All four measures of potential gerrymandering are statistically significant for the 2011 map.
+In contrast, as we might expect, the court-ordered plan, prepared by a respected expert whose instructions were to satisfy good government criteria, was by a considerable margin the closest to a perfectly neutral plan in partisan terms according to all four measures.
+Indeed, none of the four measures of potential gerrymandering is statistically significant for the court-ordered map.
+And when we turn to the Republican Joint Legislative plan we find that all four measures are again statistically significant.
+\par
+    Finally, we return to the plan offered by the Democratic governor, Governor Wolf.
+To our surprise, this plan appeared to have a pro-Republican tilt, nearly to the same extent as the joint legislative plan for three of the four metrics in Table \ref{tab:gerry}.
+This puzzling finding suggests that the Governor might have been under pressure from Democratic incumbents not to reduce their expected vote margins, or that he sincerely intended it to be a compromise that the Republican legislature might agree to.
+However, as shown in \ref{tab:gerry}, two of the apparent pro-Republican indicators for the plan are statistically insignificant.
+Moreover, there is one important difference between the joint plan and that of the Governor: among the seven districts likely to be won by Democrats, the Democratic percentages are cut more narrowly in the Republican plan than in the Governor’s plan in three of the districts so that, in a good Republican year, Republicans will do better under their plan than under the Democratic plan.
+\par
 % =================================================================
@@
 \par
-    The findings from Table \ref{tab:gerry} are very clear; Regardless of which metric we examine, the 2011 map is the most gerrymandered.
-Not surprisingly, the nature of the bias is in a pro-Republican direction.
-Given the cumulative weight of all the evidence, the 2011 congressional map was clearly an egregious pro-Republican gerrymander.
-In contrast, as we might expect, the court-ordered plan, prepared by a respected expert whose instructions were to satisfy good government criteria, was by a considerable margin the closest to a perfectly neutral plan in partisan terms according to all four measures.
-\par
-    Finally, we turn to the plan offered by the Democratic governor, Governor Wolf.
-To our surprise, this plan turned out also to have a pro-Republican tilt, nearly to the same extent as the joint legislative plan for three of the four metrics in Table \ref{tab:gerry}.
-This puzzling finding suggests that the Governor might have been under pressure from Democratic incumbents not to reduce their expected vote margins, or that he sincerely intended it to be a compromise that the Republican legislature might agree to.
-However, there is one important difference between the joint plan and that of the Governor: among the seven districts likely to be won by Democrats, the Democratic percentages are cut more narrowly in the Republican plan than in the Governor’s plan in three of the districts so that, in a good Republican year, Republicans will do better under their plan than under the Democratic plan.
-\par
-    Still, when we look at Table \ref{tab:gerry} we are struck by the fact that all four plans, even the Court-ordered one, still appear to have some level -- very slight in one case, extreme in others -- of pro-Republican bias.
-In the next section we consider why this might have happened, and we introduce a potentially important distinction between ``\textit{neutral plans}" and ``\textit{fair plans}."
-\par
-    In sum, looking at the evidence above, we see exactly the same ordering of plans according to their characterizability as a partisan gerrymander as we saw when we ordered plans according to the degree to which they satisfied good government criteria (see Table \ref{tab:summaries}), with the 2011 map the worst and the 2018 court-ordered map far and away the best.
+    Looking at the evidence above, we see exactly the same ordering of plans according to their characterizability as a partisan gerrymander as we saw when we ordered plans according to the degree to which they satisfied good government criteria (see Table \ref{tab:summaries}), with the 2011 map the worst and the 2018 court-ordered map far and away the best.
 This conclusion is further buttressed by the data we present in Table \ref{tab:prob}.
 Here we create a probabilistic simulation using statewide five-election composite (set to a 50-50 vote share) to estimate the likelihood that random shocks (based on past state inter-election shift data) would change party control of the congressional delegation.
+We report the 95 percentile interval of expect seat shares for each party when the vote is exactly tied 50-50.
+%        =================================================================
+%        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
+% ------------------------------------------------------------------------
+        \footnote{Simulations are conducted using a simple bootstrapping procedure where we uses the historic inter-election variability to add random noise to the composite results.
+An overview of this method can be found in \citet{Efron1979}.}
+% ------------------------------------------------------------------------ 
+%        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
+%        ================================================================= 
+    The expectation is that at a tied vote, when there is an even number of districts, that each party would receive half the seats.
+This symmetric result is equivalent to a ``fair plan''.
+In the court-remedial plan, we see a normal distribution of expected seat-shares ranging from 7.2 to 12.6 Republican seats at 50\% of the vote (Republicans are expected to gain at least 10 seats 69\% of the time).
+For the Republican joint plan and the Democratic plan submitted by Gov.
+Wolf, the Democrats are virtually never expected to have a majority of seats.
+Even in the Democratic plan, claimed by some to be a pro-Democratic gerrymander, Democrats are only expected to receive half or more of the seats 11.7\% of the time.
+In looking at the unconstitutional gerrymander plan of 2011, Republicans are, except in highly unlikely outliers, predicted to gain a majority (10+) seats, and to win as many as 14/18 at \underline{50\%} of the vote!
+In our simulations of 1,000 potential election outcomes under the enacted 2011 map, 986 resulted in at least ten Republican seats.
+The Republican remedial plan is not much better, delivering ten plus seats for the Republicans in 973 of the 1,000 simulations.
+The Democratic plan improves slightly by limiting Republican majorities to just 883/1,000 simulations, while the courts map is the most fair with only 694 simulations resulting in an outright majority for Republicans.
 \par
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
+    A number of analyses that come to conclusions similar to our own are found in \citet{Nagle2019_ELJ}, an essay apparently written simultaneously with this one.
+His work is a useful complement to this one because much of it is concerned with alternative good government maps and their properties.
+His primary concern is that voters are fairly represented and that plans are responsive to voters.
+Nagle classifies the 2011 map as a pro-Republican gerrymander and also finds the Republican remedial map to be a ``stealth'' gerrymander.
+He additionally agrees with our conclusion that, if there is any bias in the court plan it is in a pro-Republican direction.
+\par    
+    In the next section we introduce a potentially important distinction between ``\textit{neutral plans}'' and ``\textit{fair plans}'', and we consider some important aspects of Pennsylvania's electoral geography.
+\par
+%
+% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
+% ▀▄▀▄▀▄ SUBSECTION -- SUBSECTION -- SUBSECTION -- SUBSECTION -- ▄▀▄▀▄▀
+% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
         \subsection*{Natural Gerrymanders}
@@
     In 2018, even under a court-drawn plan, in a year when Democrats had won a clear majority of the aggregate two-party congressional vote, Democrats only succeeded in winning exactly 50\% of the seats.
-Trying to understand why all four of the plans we considered had some level of pro-Republican bias requires us to look in more detail at the partisan electoral geography of Pennsylvania.
+Trying to understand why evan plans not drawn by Republicans had some level of pro-Republican bias requires us to look in more detail at the partisan electoral geography of Pennsylvania.
 In Pennsylvania, Philadelphia is overwhelmingly Democratic in voting.
@@
 There are no other large, concentrated pockets of equivalently overwhelmingly Republican voting strength.
-Thus, more Democratic votes will \textit{naturally} be ``wasted" in Philadelphia than Republican votes will be ``wasted" elsewhere in the state.
+Thus, more Democratic votes will \textit{naturally} be ``wasted'' in Philadelphia than Republican votes will be ``wasted'' elsewhere in the state.
 Additionally, the city of Philadelphia is landlocked, and its position on the state's border reduces the ability of line drawers who favor Democrats to distribute these voters into multiple districts so as to minimize the extent to which Democratic votes in the county are wasted.
@@
 % ------------------------------------------------------------------------
-        \footnote{Expert Witness Report, Jowei Chen (Pg. 15, 16, Figure2).}
+        \footnote{Expert Witness Report, Jowei Chen (Pg. 15, 16, Figure 2).}
 % ------------------------------------------------------------------------ 
@@
 \par
-    Recognizing the potential for so-called ``natural gerrymandering", we wish to highlight a potentially important distinction between \textit{neutral} plans and \textit{fair} plans -- each of which reflects a different approach to defining the baseline for what constitutes gerrymandering.
+    Recognizing the potential for so-called ``natural gerrymandering'', we wish to highlight a potentially important distinction between \textit{neutral} plans and \textit{fair} plans -- each of which reflects a different approach to defining the baseline for what constitutes gerrymandering.
 Neutral plans refer to those that are drawn entirely with respect to traditional good government criteria, with no attention paid to partisan considerations.
@@
 % ------------------------------------------------------------------------
-        \footnote{Similarly, \citet{Grofman2018_ELJ} asserts that ``Gerrymandering occurs when a districting plan creates a disparate treatment of the vote share of the minority and majority voting blocs in a way that penalizes the minority in its ability to translate its voting support into seats compared to what we might expect from a plan drawn on the basis of neutral principles."}
+        \footnote{Similarly, \citet{Grofman2018_ELJ} asserts that ``Gerrymandering occurs when a districting plan creates a disparate treatment of the vote share of the minority and majority voting blocs in a way that penalizes the minority in its ability to translate its voting support into seats compared to what we might expect from a plan drawn on the basis of neutral principles.''}
 % ------------------------------------------------------------------------ 
@@
 \par
-    In contrast, we may wish to use the term ``fair" as plans that seek to compensate for the difference in level of partisan concentration.
+    In contrast, we may wish to use the term ``fair'' as plans that compensate for the difference in level of partisan concentration.
 Fair plans do not seek to achieve proportional representation.
-Rather, they seek to achieve partisan symmetry.
-A fair plan would yield a proportional result when the vote shares are evenly split 50-50, but at points beyond 50-50, such as when party R wins 70\% of the districts with only 55\% of the vote, party L can similarly win 70\% of the districts if it had received 55\% of the vote.
+Rather, they aim to achieve partisan symmetry.
+A fair plan would yield a proportional result when the vote shares are evenly split 50-50, but at points beyond 50-50, such as when party R wins 70\% of the districts with only 56\% of the vote, party L can similarly win 70\% of the districts if it had received 56\% of the vote.
+%        ================================================================= 
+%        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
+% ------------------------------------------------------------------------  
+    \footnote{\citet{Nagle2019_ELJ} views ``fair plans'' as the ideal, even if drawing a plan requires some degree of relaxing traditional and neutral criteria to accomplish symmetric maps.}
+% ------------------------------------------------------------------------ 
+%        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
+%        =================================================================
+%
 \par
-    One clear finding of our analysis of the court-ordered map is that it was drawn as a ``neutral map", though not necessarily a fair map in the sense we are using the term above.
-In order to create a truly ``fair map", in our view, it would have been necessary to break up Philadelphia County in more than three pieces so as to avoid the ``natural" packing of Democratic voters.
+    One clear finding of our analysis of the court-ordered map is that it was drawn as a ``neutral map'', though not necessarily a ``fair map'' in the sense we are using the term above.
+In order to create a truly ``fair map'', in our view, it would have been necessary to break up Philadelphia County in more than three pieces so as to avoid the natural packing of Democratic voters.
 %        ================================================================= 
@@
 %        ================================================================= 
-    But the Pennsylvania Supreme Court opted not to do that, but to instead preserve county lines and draw a good government map. We therefore identify the Court remedial plan as a ``neutral" plan.
+    But the Pennsylvania Supreme Court opted not to do that, but to instead preserve county lines and draw a good government map. We therefore identify the Court remedial plan as a ``neutral'' plan.
 \par
@@
 % =================================================================
-% -- SECTION -- SECTION -- SECTION -- SECTION -- SECTION -- SECTI %
+% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
+% █▀▀ █▀▀ █▀▀ ▀▀█▀▀ ░▀░ █▀▀█ █▀▀▄
+% ▀▀█ █▀▀ █░░ ░░█░░ ▀█▀ █░░█ █░░█
+% ▀▀▀ ▀▀▀ ▀▀▀ ░░▀░░ ▀▀▀ ▀▀▀▀ ▀░░▀
 % •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
@@
     While the Pennsylvania court opinion is limited to Pennsylvania, and thus it might have seemed of importance only in Pennsylvania, in footnote 71 of the Opinion (slip op.
-pp.116-117), the court took what we regard as a rather unusual step.
-It issued what can only be called an invitation to other state courts to use the same logic it used to invalidate partisan gerrymanders in their own state.
-As we noted earlier, the Court pointed out that there are twelve additional states whose constitutions contain election clauses identical to the Pennsylvania charter, requiring elections to be ``free and equal”: Arizona, Arkansas, Delaware, Illinois, Indiana, Kentucky, Oklahoma, South Dakota, Oregon, Tennessee, Washington, and Wyoming.
+pp.
+116-117), the court took what we regard as a rather unusual step.
+It issued what can only be regarded as an invitation to other state courts to use the same logic it used to invalidate partisan gerrymanders in their own state.
+As we noted earlier, the Court pointed out that there are twelve additional states whose constitutions contain election clauses identical to the Pennsylvania charter, requiring elections to be ``free and equal'': Arizona, Arkansas, Delaware, Illinois, Indiana, Kentucky, Oklahoma, Oregon, South Dakota, Tennessee, Washington, and Wyoming.
 % ================================================================= 
@@
 % =================================================================
-    However, only a handful of these states are ripe for partisan gerrymandering challenges -- in that (ca.
-2017) three lack unified party control of the state, two are single district states, and some already have a commission drawing plans and in others, \textit{indicia} of gerrymandering are missing.
+    However, only a handful of these states are ripe for partisan gerrymandering challenges -- two are single district states, several may not have unified control over the districting process, some already have a commission drawing plans, and in others, \textit{indicia} of gerrymandering are missing.
 \par
     In looking to the future, we should also note that some states that are generally regarded as among the most pernicious partisan gerrymanders, Michigan, Ohio, North Carolina, Wisconsin are not included among the twelve, nor is Maryland.
-On the other hand, we should also note that a ``''free and equal” elections clause is not the only avenue state courts might use to attack partisan gerrymanders in the future.
+On the other hand, we should also note that a ``free and equal'' elections clause is not the only avenue state courts might use to attack partisan gerrymanders in the future.
 As University of Kentucky College of Law Professor Joshua Douglas has pointed out, virtually every state constitution protects voting rights more explicitly than the U.S.
 Constitution does \citep{Douglas2014_RightToVote}.
-In addition to the thirteen states that require elections to be ``free and equal”, an additional thirteen have state constitutional provisions that require elections to be ``free" or ``free and open”, and these clauses could, in principle, be used in exactly the same way as the ``free and equal” clause.
+In addition to the thirteen states that require elections to be ``free and equal'', an additional thirteen have state constitutional provisions that require elections to be ``free'' or ``free and open'', and these clauses could, in principle, be used in exactly the same way as the ``free and equal'' clause.
 % ================================================================= 
@@
         \footnote{We are indebted to Jonathan Lai of the Philadelphia Inquirer (personal communication, April 2018) for calling this information to our attention.
+These states include Colorado, Maryland, Massachusetts, Missouri, Montana, Nebraska, New Hampshire, New Mexico, North Carolina, South Carolina, Utah, Vermont, and Virginia \citep[][footnote 86]{Douglas2014_RightToVote}.
 See a more detailed discussion in \citet{Elmendorf2018}.}
@@
 % =================================================================
-    Such clauses are found in a number of the states widely regarded as having the worst congressional gerrymandering. 
+    Such clauses are found in some of the states widely regarded as having the worst congressional plans, such as Maryland and North Carolina.
+\par
+    Accordingly, some scholars have suggested turning away from the federal judiciary altogether and instead focusing on state-level litigation \citep[e.g.
+][]{Wang_et_al_2019_Labortories_UPJCL}.
+We believe that the \textit{LWV} decision and the further comparative analyses of alternative plans presented in the previous sections of this paper will help other state courts navigate their way to decisions that strike down some partisan gerrymanders as unconstitutional, while allowing others to remain in place on the grounds that either they are not that severe, or they are unlikely to be lasting in that they have sufficiently many competitive seats as to be responsive to realistic changes in voting patterns.
+\par    
+    Now that litigation has been successful in state courts on the grounds of egregious deviations from good government districting criteria, it seems plausible that in the next wave of redistricting, mapmakers may create covert gerrymanders like we've described in this essay.
+If so, there remains the issue of whether state courts have the legal tools to deal with these ``\textit{stealth gerrymander}''.
 % ================================================================= 
@@
 % -----------------------------------------------------------------
-        \footnote{These states include Arizona, Arkansas, Colorado, Delaware, Illinois, Indiana, Kentucky, Maryland, Massachusetts, Missouri, Montana, Nebraska, New Hampshire, New Mexico, North Carolina, Oklahoma, Oregon, Pennsylvania, South Carolina, South Dakota, Tennessee, Utah, Vermont, Virginia, Washington, and Wyoming \citep[][footnote 86]{Douglas2014_RightToVote}.}
+        \footnote{Commenting on the potential for the North Carolina district court to strike down their maps as unconstitutional (which ultimately they did), David Daley, author of ``Ratf**ked: Why Your Vote Doesn't Count'' \cite{Daley2016_ratf**ked}, stated about a possible remedial plan by the map's original consultant that, ``[Thomas B.
+Hofeller] was really good at maps that look more normal, that break up fewer towns and counties but are just as partisan and just as advantageous for Republicans.
+It wouldn’t surprise me at all if on a hard drive somewhere in Raleigh Tom Hofeller has another set of gifts for legislators.'' (as cited in \citet{Barnett2018_NC_gerry}).
+As we have pondered in this essay about Pennsylvania, the North Carolina court would have had to grapple with a plan which, according to traditional criteria, doesn't meet the standards of a gerrymander, but when looking at the partisan outcomes, clearly does.
+That a plan which was created which accomplished the partisan goals of the districters but also shielded itself from agitators by hiding the obvious indicators of gerrymandering was not enacted over one that makes obvious the intentions and thus potentially vulnerable to litigation is beyond the current authors' imagination.}
 % ----------------------------------------------------------------- 
@@
 \par
-    Accordingly, some scholars have suggested turning to turn away from the federal judiciary altogether and instead focusing on state-level litigation \citep{Wang_et_al_2019_Labortories_UPJCL}.
-We believe that the \textit{LWV} decision and the further comparative analyses of alternative plans presented in the previous sections of this paper will help other state courts navigate their way to decisions that strike down some partisan gerrymanders as unconstitutional, while allowing others to remain in place on the grounds that either they are not that severe, or they are unlikely to be lasting in that they have sufficiently many competitive seats as to be responsive to realistic changes in voting patterns.
-Now that litigation has been successful in state courts on the grounds of egregious deviations from good government districting criteria, it seems plausible that in the next wave of redistricting, mapmakers may create covert gerrymanders like we've described in this essay.
-If so, there remains the issue of whether state courts have the legal tools to deal with these ``stealth gerrymander”.
+    According to all four measures, the joint legislative plan offered by Republican leaders is nearly as much a gerrymandered map -- in a pro-Republican direction -- as is the 2011 map.
+As noted above, we believe it appropriate to label this plan as a \textit{stealth gerrymander} since it creates partisan asymmetries while (mostly) providing the visual appearance of a good government map (see Figure \ref{fig:maps}).
+However, having labeled the joint legislative plan a \textit{stealth gerrymander}, it is worth reminding readers of two important points.
 \par
-    According to all four measures, the joint legislative plan offered by Republican leaders is nearly as much a gerrymandered map -- in a pro-Republican direction -- as is the 2011 map.
-As noted above, we believe it appropriate to label this plan as a \textit{stealth gerrymander} since it creates partisan assymetries while (mostly) providing the visual appearance of a good government map (see Figure \ref{fig:maps}).
-However, having labeled the joint legislative plan a \textit{stealth gerrymander}, it is worth reminding readers of two important points.
     First, had this plan been the original plan adopted by Republicans, it is very much an open question as to whether it would have been rejected by the Pennsylvania Supreme Court.
 Even though, in partisan terms, it is nearly identical to the 2011 plan, it is considerably more consistent with good government criteria (see Table \ref{tab:summaries}) -- not that much worse than the Court Remedial plan -- that it would have required detailed political election analyses like those we have done here to demonstrate its \textit{stealth gerrymander} features, plus a court willing to rule on political grounds, rather than good government grounds that it was unconstitutional.
-Because the joint legislative plan was not passed by the State of Pennsylvania's legislature, the Court felt no need for particular deference to it (or to the Governor's plan, for that matter).
-Thus, the Court was not bound by the normal supposition with respect to districting that a plan authorized by the state need not be the ``best possible", but only unconstitutional.
-That ``deference" to legislative judgments and the criteria they reflect, might well have tipped the balance toward acceptance of the Republican joint legislative plan were it offered as having been fully sanctioned by the State of Pennsylvania (the duly elected legislature and Governor).
+Second, because the joint legislative plan was not passed by the State of Pennsylvania's legislature, the Court felt no need for particular deference to it (or to the Governor's plan, for that matter).
+Thus, the Court was not bound by the normal supposition with respect to districting that a plan authorized by the state need not be the ``best possible'', but only constitutional.
+That ``deference'' to legislative judgments and the criteria they reflect, might well have tipped the balance toward acceptance of the Republican joint legislative plan were it offered as having been fully sanctioned by the State of Pennsylvania (the duly elected legislature and Governor).
 % ================================================================= 
@@
         \footnote{The Court picked the plan among those before it that most closely satisfied good government criteria -- which happily, thanks to Professor Persily's expertise as consultant to the Court, turned out to be its own plan, and thus a plan which the Court could know with certainty was not intended as a partisan gerrymander for either party.
-Whether it would have still sought to maximize compliance with good government criteria had the plan which did so been a ``stealth gerrymander" is not a question we can answer.}
+Whether it would have still sought to maximize compliance with good government criteria had the plan which did so been a ``stealth gerrymander'' is not a question we can answer.}
 % ----------------------------------------------------------------- 
@@
 % =================================================================
+\par    
     On the other hand, in the presence of this \textit{stealth gerrymander} rather than the blatant gerrymander that came before it, the Pennsylvania Court might well have relied on partisan impact evidence to reach a conclusion of unconstitutionality.
@@
 I, § 5.
-On this record, it is clear that the 2011 Plan violates Article I, Section 5, since a diluted vote is not an equal vote" (slip op.
+On this record, it is clear that the 2011 Plan violates Article I, Section 5, since a diluted vote is not an equal vote'' (slip op.
 p.
@@
 Citizen-based reforms such as referendum and initiatives may also help with reducing egregious gerrymandering in states where the passage of such measures is feasible.
+% ================================================================= 
+% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
+% -----------------------------------------------------------------
+        \footnote{Since the U.S.
+Supreme Court's remand of the \textit{Gill} case in June 2018, Colorado, Michigan, Missouri, and Utah all passed ballot initiatives taking control of redistricting away from legislatures and given them to independent commissions.
+Six other states had, in the past several decades, adopted independent commissions: Alaska, Arizona, California, Idaho, Montana, and Washington.
+While some reformers see this as an excellent alternative to litigation, University of California Irvine Law Professor Rick Hasen (\href{https://blog.harvardlawreview.org/the-next-threat-to-redistricting-reform/}{Harvard Law Review -- The Next Threat to Redistricting Reform}) argues that the U.S.
+Supreme Court might, in the near future, disallow such commissions as unconstitutional when they involve congressional districting.
+Hasen's argument turns on the explicit language of the Constitution (''The Times, Places and Manner of holding Elections for Senators and Representatives, shall be prescribed in each State by the Legislature thereof; but the Congress may at any time by Law make or alter such Regulations, except as to the Places of chusing [sic] Senators.''), and the views of Justice Roberts in the Arizona Independent Commission case.
+On the other hand, forbidding redistricting commissions would call into question popular sovereignty and the right of the people to place checks on the legislature.
+H.R.1, introduced on the first day of the new legislative session after Democrats took control of the House of Representatives in what is widely viewed as a rebuke of President Trump's first two years in office, takes advantage of the constitutional provision allowing for Congress to make or alter the manner of elections by explicitly allowing for independent commissions in apparent anticipation of future litigation.
+It has passed the U.S.
+House of Representatives, but action has not been taken by a Republican majority U.S.
+Senate (circa June 2019).}
+% ----------------------------------------------------------------- 
+% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
+% ================================================================= 
     Ultimately, there is no substitute for a federal standard that guarantees each citizen in the United States \textit{equal protection under the law}.
 Absent the United States Supreme Court setting this standard, state courts offer one promising pathway towards ending undemocratic partisan gerrymandering.
-
-% 
-% 
-% 
-% 
-%
-
-% % ================================================================= 
-% % -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% % -----------------------------------------------------------------
-%         \footnote{Because of its high number of competitive seats, a point prediction such as 9R-9D is somewhat misleading for the court-ordered map.
-Using the five-election model as our predictor, there are fewer safe Democratic seats (5) than safe Republican seats (7) in the Court plan, using 45\%-55\% definition of a competitive seat.
-This is one fewer safe seat for Republicans and one additional safe Democratic seat than is found in the 2011 plan, but the Court plan also has five districts balanced on a knife edge, with a projected winner’s margin under two percentage points, making it highly responsive to shifting voter sentiment.
-This is realized in our simulations where the mean seat share, though near 9R-9D, is either lower or higher depending on where the state-wide vote is centered.}
-% % ----------------------------------------------------------------- 
-% % -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% % =================================================================
-
-% 	Past congressional results alone often are departures from the underlying district partisanship several reasons.
-The vote share of an incumbent has historically been greater than it is for a non-incumbent or two non-incumbents running for an open seat.
-
-% %        ================================================================= 
-% %        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% % ------------------------------------------------------------------------
-% 		\footnote{A long literature in political science has evaluated and confirmed the presence and increased incumbency advantage \citep{Abramowitz1991,Levitt1997,Ansolabehere2002b,Gelman_Huang_2008}.
-Recent scholarship, however, has shown a decrease in the advantage towards zero \citep{Jacobson2015_JOP}.}
-% % ------------------------------------------------------------------------ 
-% %        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% %        =================================================================
-
-%     A well-financed incumbent tends to suppress quality competition \citep{Jacobson2015_JOP}, vulnerable incumbents who can sense the political winds opt against running for re-election \citep{Jacobson_Kernell_1981,Ansolabehere_Snyder_2004_LSQ}, and electoral tides and varying turnout can distort the state-wide average.
-Midterm elections often have vastly different electorates than presidential elections \citep{Abramowitz_et_al_1986_AJPS}, while the midterm pattern of ``surge and decline" (what goes up must come down principle) reduces the sitting Presidents' party's vote share \citep{CampbellA1960}.
-%  \par
-% 	What we are after is a measure of the ``normal vote''.
-The normal vote is the underlying partisanship of a district \citep{Converse1966}.
-It is sort of a mystical number since it is uncovered and unobserved, but powerful theoretically.
-Map builders implicitly assume a normal vote when constructing a district plan.
-We believe that state-wide elections offer the most useful approximation of the normal vote.
-% %        ================================================================= 
-% %        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% % ------------------------------------------------------------------------
-% 		\footnote{We particularly believe this is true when the votes between elections are highly correlated, and the district-level variation is low, as it was in Pennsylvania from 2012 to 2016.
-The more stable the electorate is, the more confidence we have that state-wide elections are revealing the normal vote \citep{Ansolabehere_et_al_2000_AJPS}.}
-% % ------------------------------------------------------------------------ 
-% %        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% %        =================================================================
-% 	 All districts vote on the same candidates; hence outcome affects all voters equally.
-The effects of incumbency can distort the vote in congressional races in a way that doesn't resemble the actual partisan preferences at the district-level.
-State-wide elections eliminate the need to impute or otherwise deal with uncontested districts.
-Taking an average of the state-wide elections provides additional reason to believe it's capturing a normal vote due to a regression to the mean effects.
-Justices in \textit{Vieth} offered skepticism in using elections that have contrasting results; we do not share this view, instead seeing it as an important source of variation that helps to recognize the underlying partisanship in which \textit{should} be used in measuring bias in districting plans.
-Additionally, averaging over multiple races and years helps to reduce the bias imposed by candidate effects.
-% \par
-% 	We construct the five-election composite of state-wide elections by aggregating voting district level (precincts) data to the congressional district level.
-
-% 	Additionally, each district's outcome is identical to the 2016 congressional district results.
-Indeed, after accounting for incumbency, our composite measure is remarkably accurate in its predictions of the actual congressional results validating our measure.
-The average difference between our composite and the congressional results was just 2.16\%.
-
-% %        ================================================================= 
-% %        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% % ------------------------------------------------------------------------	
-% 	We calculate this average only for districts contested in 2016, since uncontested results have significant deviations.
-Including the uncontested districts increases this mean difference to just 3.93\%.
-We believe our composite measure offers better representation of the true partisanship in these districts.
-Individually, only the Auditor's race correctly predicts all the districts correctly, but by either summing or by taking the average of the five races, both methods correctly predict each district's outcome.
-% % ------------------------------------------------------------------------ 
-% %        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% %        =================================================================
-
-%     We imputed the values 25\% and 75\% for uncontested or essentially uncontested races.
-Rather than taking the projected outcome we arrive at from this method as a single value, we looked at the sensitivity of this projection to the empirical range of inter-election swing, simulating outcomes 1,000 times.
-To arrive at our estimate for inter-election variation that can't be attributed to the normal vote or incumbency advantage, we estimate the equation using congressional elections in Pennsylvania in the 2004-16 time period:
-%         \begin{align*}
-%         \mathrm{Vote}_{ty} = \alpha + \beta_{1} \mathrm{Vote}_{yt-1} + \beta_{2} \mathrm{Incumbency}_{yt} + \beta_{4} \mathrm{Uncontestedness}_y{t} + \epsilon
-%         \end{align*}
-%     We are interested in $\epsilon$, which is the unexplained variance, which includes inter-election swing.
-We run this equation separately for each of the non-redistricting years and then average the coefficients.
-We estimate incumbency advantage to average 4.27\% for this time period, while inter-election variance is 4.46\%.
-When simulating elections, we use the composite as the normal vote, but add in for each district a random draw from a normal distribution centered at the district vote with a standard deviation of $\epsilon$.
-We can then calculate statistics for each of these simulated elections to find a distribution for each measure of gerrymandering.

### 2019-06-18 R/Tables.R
Table title changed

@@ -178,7 +178,7 @@ gerry.table.gen <-
 		"Declination",
 		"CI4")
 
-gerry.caption = "Measures of Gerrymandering for the Four Considered Plans"
+gerry.caption = "Measures of Gerrymandering for the Four Considered Plans at 50\\% Vote Share"
 gerry.label = "tab:gerry"
 gerry.footnote = "$^{*}$p $<$ 0.05; $^{**}$p $<$ .01; $^{***}$p $<$ 0.001. Measures are averages of 1,000 simulations for each map using the 2016 composite. Brackets numbers are the 95\\% range."
  

### 2019-06-18 Latex/_maintext.tex
File Updates

--- 2019-06-06 Latex/_maintext.tex
@@
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-% -- SECTION -- SECTION -- SECTION -- SECTION -- SECTION -- SECTI %
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-\section{Introduction}
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-    \lettrine[lines=3, lraise=0.1, nindent=0em]{\color{lightgray}I}{n} \textit{League of Women Voters v.
-Pennsylvania}, decided January 22, 2018 (henceforth abbreviated \textit{LWV}), the Pennsylvania Supreme Court offered a new and distinctive test for unconstitutional partisan gerrymandering under which it found the state’s congressional districts to be a ``severe and durable" partisan gerrymander that violated a long-standing provision of the state’s constitution.
-The congressional map that was struck down was drawn in 2011 by a Republican-controlled legislature and signed into law by the then Republican governor.
-It was used in elections from 2012 to 2016.
-Prior to the \textit{LWV} decision, many academics had identified this map as a partisan gerrymander \citep[see e.g.,][]{Royden2017}, with some even identifying it as the most egregious congressional districting in the U.S.
-\citep[][ esp.
-Figure 3]{Wang2016_SLR}.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % =================================================================
-% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
-        \footnote{See also \citet{Mcgann_et_al_2015_ELJ, McGann_et_al_2016_gerrymandering}}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
+\title{Tools for Identifying Partisan Gerrymandering}
+\author{J. Cervas \& B. Grofman}
+% 
+%
+% Jonathan R. Cervas - jcervas@uci.edu - University of California Irvine
+% Bernard Grofman - bgrofman@uci.edu - University of California Irvine
+%
+% 
+% First Submitted: Sept 12, 2018
+% Revise and Resubmit: January 21, 2019
+% Accepted: 
 % =================================================================
-\par
-    The Pennsylvania Supreme Court held that ``the fact that the 2011 Plan cannot, as a statistical matter, be a plan directed at complying with traditional redistricting requirements is \underline{sufficient} to establish that it violates the Free and Equal Elections Clause" [of the Pennsylvania State Constitution] (slip op.
-at p.128, emphasis added).
-% ================================================================= 
-% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
-        \footnote{``Elections shall be free and equal; and no power, civil or military, shall at any time interfere to prevent the free exercise of the right of suffrage." Pa. Const. art. I, § 5.}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% =================================================================
-    By \textit{traditional} principles (also sometime called \textit{neutral} principles, or \textit{good government} principles), the Court meant drawing contiguous districts, satisfying the \textit{one person, one vote} standard, avoiding diluting the voting strength of protected racial or ethnic minority groups, minimizing unnecessary political sub-unit splits, and providing reasonably compact districts.
-% ================================================================= 
-% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
-        \footnote{Avoiding fragmentation of communities of interest is also sometimes included in this list.}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% =================================================================
-    However, the Court also considered a variety of other indicators involving the partisan aspects of the plan in coming to the conclusion that the map created a ``severe and durable" partisan gerrymander.
-% ================================================================= 
-% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
-        \footnote{The trial court considered a variety of factors offered in expert witness testimony, including evidence about the map‘s political impact.
-The trial court’s findings of fact reveal that the 2011 map exhibited clearly a multiplicity of indicia of partisan gerrymandering.
-In his review, the trial judge found, in essence, that the legislative process through which the plan had been chosen demonstrated the indicia of a partisan gerrymander, that the weird shapes of a number of the districts demonstrated the indicia of a partisan gerrymander, that the dismemberment of municipalities, townships and counties in the state demonstrated the indicia of a partisan gerrymander, that the cracking and packing of Democrat-leaning voters demonstrated the indicia of a partisan gerrymander, and that the frozen 13R-5D results over the course of three elections in a state that is highly politically competitive statewide demonstrated the indicia of a partisan gerrymander.
-And the trial judge also found that none of these features of the map could be explained as necessitated by the electoral geography and demography of the state, though he suggested the possibility that consideration such as incumbency protection might have mattered.
-However, the trial court then, rather regretfully, concluded that it did not have a legal basis to declare the plan unconstitutional under either the Pennsylvania State Constitution or under then existing federal law precedents.
-Instead, it expressed a pious hope that the U.S.
-Supreme Court, in its review of the appeals of lower court cases, such as the 2017 case of \href{https://www.brennancenter.org/sites/default/files/legal-work/16-1161_Opinion.pdf}{\textit{Gill v.
-Whitford 585 U.
-S.
-\_\_\_ (2018)}.}, which had overturned as a partisan gerrymander a Wisconsin legislative map, would finally adopt a (manageable) standard for identifying unconstitutional partisan gerrymandering that all courts could apply.
-The Pennsylvania Supreme Court, for all practical purposes, adopted in \textit{toto} the factual finding of the trial court magistrate, Judge Brobson, but came to a diametrically opposite legal conclusion about the constitutionality of the congressional plan and the legal grounding for that finding.}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% =================================================================
-\par
-    Having invalidated the \textit{2011 Enacted} map, the Pennsylvania Supreme Court offered deferrence to the State to draw a remedy plan.
-The Pennsylvania legislature failed to enact a lawful remedy plan of its own due mostly to divided government that led to disagreement as to how to proceed between the newly elected Democratic governor and the still Republican-controlled legislature.
-Had such a plan been passed (and signed), the Court would have reviewed its constitutionality.
-In the absence of a legislatively enacted remedial plan, the Court invited submissions of plans from the parties to the lawsuit, as well the public at large, to submit remedy plans; and it provided guidance as to the features of the plan in which it was most interested.
-The Court Order in February 2018 inviting submissions of feasible maps mentioned only good government criteria: minimizing the number of county, township and municipality splits, and maximizing district compactness under several different measures of compactness.
-It made no mention of incumbent protection, nor voting rights act protections, nor of protecting communities of interests, nor did it mention political outcome projections.
-\par
-    While a remedy plan was offered by Republican leaders from both chambers (the \textit{Joint legislative} plan) and by the Democratic Governor (\textit{Gov.
-Wolf}), as well as by others, the Court rejected all such plans.
-Instead, the Court ordered into place for the 2018 election a map of its own drawn for it by a court-appointed consultant.
-In a split decision, the Court map was endorsed only by judges with Democratic affiliations, though some of the disagreement on the Court had less to do with the map than with the timing of its implementation.
-% ================================================================= 
-% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
-        \footnote{Pennsylvania Supreme Court Judges are elected or retained to ten-year terms in elections with party labels on the ballot.
-Judges with Democratic affiliations now constitute the majority of the Court In 2011 this was not the case.}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% =================================================================
-    The initial ruling that the 2011 plan was unconstitutional was sharply attacked by Republicans as judicial overreach, and the map the court majority adopted was rather viciously attacked by Republicans as simply a pro-Democratic gerrymander under the guise of a court-drawn map.
-% ================================================================= 
-% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
-        \footnote{On March 20, 2018, Rep.
-Cris Dush, of Jefferson County introduced a measure that would impeach the four Democratic members of the court that voted to replace the map.
-His argument is that the members of the court overstepped their rights and violated the Separation of Powers principle of the state.
-Others saw this as simply sour grapes: \\ \href{https://wapo.st/2LlEVvE}{``Pennsylvania Republicans lost the redistricting battle.
-Now, they’re declaring war on the courts", https://wapo.st/2LlEVvE (Christopher Ingraham, February 22, 2018).;} \\ \href{https://www.philly.com/philly/news/politics/will-republicans-impeach-pennsylvania-supreme-court-justices-20180222.html}{``Pa.
-Republicans are talking about impeaching state Supreme Court justices.
-Do they have an argument?", https://www.philly.com/philly/news/politics/will-republicans-impeach-pennsylvania-supreme-court-justices-20180222.html (Andrew Seidman \& Jonathan Lai, February 22, 2018).}}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% =================================================================
-    However, the U.S.
-Supreme Court has declined reviewing either the state court opinion rejecting the 2011 plan or the state court’s chosen map -- presumably because these decision are based entirely on state law.
-\par
-    In contrast to the \textit{LWV} finding that the 2011 Pennsylvania congressional map was unconstitutional under state law, a 2017 challenge to the Pennsylvania congressional map under the Elections Clause of the U.S.
-Constitution was rejected by a three-judge federal court by a vote of 2-1.
-Because the \textit{LWV} decision mooted the federal litigation, the federal challenge was not appealed to the U.S.
-Supreme Court.
-However, a 2018 remand of a Wisconsin partisan gerrymandering case, \textit{Gill v.
-Whitford} has, to a limited extent, provided clues about the current [ca.
-May 2019] state of federal partisan gerrymandering jurisprudence than anything before it.
-% ================================================================= 
-% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
-        \footnote{The \textit{Gill} remand decision still leaves open the question that has haunted opponents of partisan gerrymandering for the last thirty plus years, namely `When, if ever, will the U.S.
-Supreme Court actually declare a redistricting map unconstitutional?'.
-The potential for such a finding of unconstitutionality is laid out in \textit{Davis v.
-Bandemer, 478 U.S.
-109 (1986)}, which declared partisan gerrymandering justiciable, but under which the Supreme Court has never yet (as of May 2019) struck down a plan as an unconstitutional partisan gerrymander.
-Despite the 7-2 decision on the Wisconsin remand, which rejected the view that the Court should simply bow out of the process of reviewing maps as possible unconstitutional partisan gerrymanders, the retirement of Justice Kennedy and his replacement by a more conservative Justice creates even greater uncertainty about whether, in the lifetime of the present authors, there will ever be a U.S.
-Supreme Court majority that rejects some challenged district as an unconstitutional partisan gerrymander.
-We will, relatively soon, have the answer to that question due to the inevitable appeals of partisan gerrymandering cases from North Carolina and Maryland that have already had oral argument on March 26, 2019 before the U.S.
-Supreme Court (\textit{Rucho, et.
-al}.
-v.
-\textit{Common Cause, et.
-al} [1:16-CV-1026]) and \textit{Lamone, et.
-al.} v.
-\textit{Benisek, et.
-al.} [1:13-cv-03233-JKB]), with decisions expected in June 2019.
-In addition, there is a partisan gerrymandering challenge in Michigan, \textit{League of Women Voters of Michigan} v.
-\textit{Johnson} that will certainly be appealed to the U S.
-Supreme Court, as will the next lower court decision in \textit{Gill} itself.}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% =================================================================
-    In that remand decision (\textit{Gill v.
-Whitford, 585 U.S.
-\_\_\_ (2018)}), the Supreme Court indicated that any standard for partisan gerrymandering must be a challenge to individual districts which demonstrates that the voters in the particular district have had their votes either ``packed" or ``cracked."
-\par
-    Districts that have been \textit{packed} are those which have had voters from one partisan leaning grouped together so as to waste votes of the group because they win the district by an overwhelming majority, thus preventing those  excess votes from being used more efficiently in other districts.
-Districts that have been \textit{cracked} are ones where voters of one partisan leaning have had their share of the district’s voting strength reduced so as to make it impossible (or unlikely) that members of this group will be able to carry the district.
-The standard for unconstitutional gerrymandering at the federal level -- if there is ever is one -- can thus be expected to be very different from that offered by the Pennsylvania Supreme Court, since \textit{LWV} relies on data about the features and effects of the plan as a whole, while the US Supreme Court appears to be insisting on a district-specific test of packing and cracking.
-\par
-    We can identify five key features of the \textit{LWV} ruling: (a) though some districts might be particularly egregious, evidence of unconstitutional gerrymandering can be found in a plan as a whole; (b) evidence of purposeful discrimination was not necessary in order to find unconstitutionality; (c) evidence of discriminatory partisan effects was not necessary to show a violation; (d) evidence that a plan complied with good government criteria is almost certainly not sufficient to rule out a finding of unconstitutionality, even though a finding that a plan failed to comply with such criteria was sufficient to invalidate the plan; and (e) statistical evidence about the likelihood that  observed features of a plan could not be explained either by chance or by electoral geography played a pivotal role in the decision.
-\par
-    Although only directly applicable in Pennsylvania, \textit{LWV} is, we believe, nonetheless the most important partisan gerrymandering case yet decided in a final and definitive fashion this decade.
-% ================================================================= 
-% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
-        \footnote{Indeed, if the Supreme Court continues to dodge any responsibility for checking egregious partisan gerrymandering, \textit{LWV} may well prove to be the most important partisan gerrymandering decision ever \citep{Grofman_Cervas_2018_ELJ}.}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% =================================================================
-    Because the US Constitution is the minimum rights provided to all citizens, states offer additional protections.
-Language virtually identical to the provision of the state constitution that the Pennsylvania Supreme Court relied on is found in the constitutions of another dozen states, and another thirteen have very similar language \citep{Douglas2014_RightToVote}.
-% ================================================================= 
-% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
-        \footnote{See also \citet{Elmendorf2018} and further discussion in the conclusions section of this paper.}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% =================================================================
-    That language could be readily adapted to control gerrymandering in those states in much the same way as did the Pennsylvania court.
-In fact, in its opinion, the Pennsylvania Supreme Court offers what we read as direct encouragement to state courts in other states to follow the lead of Pennsylvania in crafting a way to overturn partisan gerrymandering as a violation of a \underline{state constitution}.
-% ================================================================= 
-% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
-        \footnote{Footnote 71 of the Opinion (slip op. pp.116-117)}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% =================================================================
-\par
-    However, as we show later, some states in which possible state law challenges could be made will be insulated from those challenges by factors such as the nature of institutional control over the redistricting process in the state, or political gridlock in the state courts.
-Nonetheless, if a state court does follow the Pennsylvania Supreme Court lead, we view the expected effects on limiting partisan gerrymandering to be potentially greater than we might expect under federal law, since they would involve the overturning of whole plans, not just individual districts -- although, in practice, a finding of overall unconstitutionality is likely to lead to changes only in the most egregiously gerrymandered districts and some of the districts adjacent to them.
-Unlike all federal courts that have ruled on partisan gerrymandering to date, the Pennsylvania opinion does not require proof of intent to discriminate, which could make proving a violation easier under state law than under any likely federal standard.
-Moreover, although the Pennsylvania Supreme Court opinion does not require evidence of partisan effects, and emphasizes failure to comply with good government criteria, we believe that the \textit{LWV} opinion can be ``tweaked" to address what \citet{GrofmanCervas2018_WashPo} in their on-line essay in the \textit{Washington Post} refer to as \textit{stealth gerrymanders}, i.e., plans that satisfy good government redistricting criteria to a reasonable degree but are nonetheless carefully crafted invidious partisan gerrymanders in terms of their expected political impact.
-The reason for our optimism in this respect are the findings in the opinion that do address partisan effects of the Pennsylvania congressional map.
-\par
-    In the remainder of this essay we will not seek to review in any more detail the various legal and expert witness issues raised in the Pennsylvania litigation itself.
-% ================================================================= 
-% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
-        \footnote{\citet{Grofman_Cervas_2018_ELJ} provide a review of the various expert witness testimony in the Pennsylvania case.}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% =================================================================
-    Rather, our primary aim is the straightforward one of comparing and contrasting the 2011 and 2018 Pennsylvania congressional maps in terms of a variety of proposed metrics for detecting partisan gerrymandering.
-In particular, we examine the evidence for the claim that the 2011 map was a pro-Republican gerrymander, and the claim that the 2018 map was a pro-Democratic gerrymander.
-The former claim is one virtually universally acknowledged \citep{Fang2014, Mcgann_et_al_2015_ELJ, McGann_et_al_2016_gerrymandering, Wang2016_SLR, Wang2016_ELJ}, while the latter claim has been offered not only as partisan griping by Republicans on the losing side of the litigation, but by at least one independent journalist.
-% ================================================================= 
-% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
-        \footnote{See NYT Upshot: \href{https://nyti.ms/2w8b19k}{``Democrats Didn’t Even Dream of This Pennsylvania Map. How Did It Happen?", https://nyti.ms/2w8b19k (Nate Cohn, Feb. 21, 2018).;} \\
-        \href{https://nyti.ms/2wkpEFT}{``Hundreds of Simulated Maps Show How Well Democrats Fared in Pennsylvania", https://nyti.ms/2wkpEFT (Nate Cohn, Feb. 26, 2018).}}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% =================================================================
-    However, the U.S.
-Supreme Court has declined reviewing either the state court opinion rejecting the 2011 plan or the state court’s chosen map -- presumably because these decision are based entirely on state law.
-\par
-    In addition to the rejected 2011 and the court-ordered 2018 map, we also address the partisan and other features of two maps offered to the Pennsylvania Court as a proposed remedy for constitutional violation found in the 2011 map: the \textit{joint legislative} plan offered by Republican legislators from both chambers, and the congressional map proposed by the Democratic governor (\textit{Gov.
-Wolf}).
-Here we focus on the question of whether either of these plans was a \textit{stealth gerrymander} done to aid the party drawing the map.
-\par
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-% -- SECTION -- SECTION -- SECTION -- SECTION -- SECTION -- SECTI %
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-\section{Evaluating the Various Plans -- Methods and Theory}
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-    As noted above, to contrast the Court-ordered Remedial plan with the 2011 Enacted plan, and to make further comparisons to the two most politically important remedy plans offered to the Court, the joint plan proposed by the Republicans leaders of the two legislative chambers, and the Governor’s proposed plan as a Democratic alternative, we will look at a number of features of each of these four plans as a whole.
-We do not, however, discuss district-specific criteria for identifying gerrymandering in these maps, since, although some district-specific evidence was presented about the 2011 map in the trial testimony by an expert who was highly knowledgeable about communities of interest in the state, the Pennsylvania Supreme Court opinion itself offers a \underline{statewide} standard for a finding of unconstitutional partisan gerrymandering.
-   % ================================================================= 
-% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
-        \footnote{As noted earlier, given the language in the U.S.
-Supreme Court’s remand of the \textit{Gill} case, it appears that any federal standard for \underline{partisan gerrymandering} would need to be district specific.}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% ================================================================= 
-    All the plans create contiguous districts.
-Similarly, all four plans appear to satisfy voting rights requirements.
-Thus, we do not discuss these two ``good government” criteria further.
-Consequently, we limit ourselves to metrics that report results for a jurisdiction as whole.
-\par
-    We begin our analyses describing map features distinct from projected (or actual) partisan outcomes or proposed measures of partisan effects.
-Table \ref{tab:summaries} offers a comparison of two key good government features of the map.
-\begin{itemize}
-    \item the number of county splits, since the Pennsylvania Supreme Court specifically sought to limit the number of counties spanning multiple districts
-    \item compactness measures, since compactness is also a criterion specifically referenced by the Pennsylvania Supreme Court
-\end{itemize}
-\par
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-% -- SUBSECTION -- SUBSECTION -- SUBSECTION -- SUBSECTION -- SUBS %
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-\subsection*{Measurement and Data}
-\subsubsection*{County Splits}
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-    Avoiding county splits is incorporated into the language of many state constitutions \citep{Grofman1985} and can be seen as a way of guarding against gerrymandering by making it more difficult to draw partisan maps.
-There are a total of 67 counties in PA, and the state is allocated 18 congressional districts.
-Some county splits are necessitated by the size of the county population relative to the ideal population of a congressional district (705,687).
-Allegheny and Montgomery must each be split at least twice, while Philadelphia must be split at least three times.
-If no set of counties exactly equals the ideal congressional population, every districting plan requires a minimum of $n-1$ splits (where $n$ equals the number of congressional districts).
-The Pennsylvania Supreme Court found that an unusually and egregiously large number of counties were split in the 2011 congressional map, and they found this fact deeply troubling.
-\par
-% =================================================================
-% -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE  % 
-% -----------------------------------------------------------------
-\input{Tables/_tab_summaries.tex}
-\begin{center} INSERT TABLE \ref{tab:summaries} ABOUT HERE \end{center}
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-\par
-    The data provided in Table \ref{tab:summaries} shows how the high number of county splits in the 2011 map was completely unnecessary.
-Both the Republican and Democrat proposed remedy plans cut the number of county splits from 41 to 19.
-% ================================================================= 
-% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
-       \footnote{There is a debate about how to count split counties, which was evident when different parties in the case counted different total county splits for the same plan.
-One way to count splits to count the total number of counties that are in at least two districts.
-The other way is to count the number of times a county is split, i.e., the number of separate pieces into which the county is divided.
-In the first counting method, a county can have portions of itself found in 18 different districts, but have this counted as only one county spit.
-Under the second method, this wholesale splitting of the county would be counted as 18 splits.
-While both methods provide useful information, we regard the second as more informative and it is the one reported in Table 1.
-However, we must also be alert for divisions of a county into two or more discontiguous pieces found within the same district.
-We call this “fracking,” to provide a term similar sounding to ‘packing’ and ‘cracking.’ We find “fracking” a particularly egregious problem since there is no need for counties to be fracked, and fracking may be used as a gerrymandering tool.
-Instances where counties  have multiple discontiguous pieces located in a single district are not discussed in the Pennsylvania Supreme Court ruling on the unconstitutionality of the 2011 Pennsylvania congressional plan.
-However, our analysis of the 2011 plan reveals several cases of fracking: two of the most egregious are districts 14 and 18 in Allegheny County (Pittsburgh).
-This concept has also been addressed by the US District Court in North Carolina (\textit{Common Cause v.
-Rucho} No 1:16-CV-1026 (U.S.
-District Court, Middle District of North Carolina, 2018, slip op at p.
-105 [p.
-194])).
-}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% =================================================================
-    The Court plan goes even further and gets the number of split counties down to 17. 
-\par
-    Which counties get split and how many ways can impact a plan’s election outcome, which can affect both national politics and a county's relationship with the federal government \citep{Ansolabehere2002a}.
-That the 2011 map is an eyesore vis-a-vis unnecessary splits of county boundaries is revealed clearly in Figure \ref{fig:maps}.
+\documentclass[12pt, a4paper, twoside]{article}
+\RequirePackage{cervasanon}
+% _________________________________________________________________
+%
+% _________________________________________________________________
 %
 % =================================================================
-% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
-    \footnote{We find counties that are split into multiple pieces particularly egregious.
-``\textit{Fracking}" occurs when one district is split into two or more non-contiguous pieces of a single county.
-It was not mentioned in the \textit{LWV} decision, but too can be used as an indicia of partisan (and racial) manipulation.
-We call this ``\textit{fracking}" to parallel the terminology of other districting terms of ``\textit{cracking}" and ``\textit{packing}".
-Examples of ``fracking" in the 2011 enacted map include districts 14 \& 18 in Allegheny County (Pittsburgh).}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
 % =================================================================
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-\subsubsection*{Compactness}
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-    The measurement of compactness has long been a central issue in the districting literature \citep[see e.g.,][]{Reock1961, Niemi1990, Polsby1991, Webster2013, Kaufman2018}.
-Indeed, the original gerrymander was assumed to be evidence of political manipulation because of its irregular shape \citep{Martis2008}.
-Some minimal level of compactness can make it harder to gerrymander \citep[][cf.
-\citealt{Webster2013}]{Reock1961, Altman1998}.
-But compactness is not a magic wand that rules out the possibility of political manipulation.
-Moreover, as many authors have shown, compactness has multiple dimensions, and these different dimensions need not move in parallel with one another; thus a plan appears compact on one dimension might appear ill-compact on another.
-The two most common ideas of compactness refer, on the one hand, to how close a legislative district’s boundaries are to its geographic center and, on the other, to how ``regular" in shape a district appears to be \citep{Niemi1990, Kaufman2018}.
-In Table \ref{tab:summaries} we report two well-known measures which tap respectively each of these two dimensions.
-The \textit{Polsby-Popper} measure looks at perimeter irregularity by examining the area of the district compared to that of a circle with similar perimeter, while the \textit{Reock} measure compares the area of a district with that of the district’s circumscribing circle \citep{Reock1961, Polsby1991}.
+%                       - Paper Information -
 % =================================================================
-% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
-        \footnote{$ A_D $ = area of district, $P_D$= perimeter of district, $Circle_D$= minimum circumscribing circle; \\
-        \textit{POLSBYPOPPER} = $\frac{4 \pi A_D}{{P_{D}}^{2}}$ \quad
-        \textit{REOCK} = $\frac{A_D}{A(Circle_D)}$}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% =================================================================
-\par
-    The 2011 Enacted map looks awful by both compactness measures compared to the Court-ordered map, or even compared to the two proposed remedial maps; while both the Republican and the Democratic proposed remedial maps look reasonable by both compactness measures, even though the Democratic plan is superior to the Republican plan and the court-ordered map is superior to both.
-However, until we examine the likely partisan effects of these four plans we are not in a position to conclusively rule out the possibility that one or more of the remedial plans is a ``\textit{stealth gerrymander}".
-As is now well recognized, gerrymandering can be found even in plans with compact appearing districts, or in a plan with districts that preserve most county borders.
-\par
-    Each of the three measures in Table \ref{tab:summaries} indicates a very dramatic deviation from good government standards in the 2011 Enacted map.
-And, each of three metrics order the plans in good government terms from worst to best in exactly the same way, with the worst plan being the 2011 Enacted map, the next worst being the Republican remedy plan, followed by the the Democratic remedy plan, and the very best being the court-ordered map.
-Moreover, as we will see later, the same ordering of plans is reflected when we examine the partisan implications of the various maps, with the map ordered into place as a remedy by the Pennsylvania Supreme Court close to perfectly neutral in its expected partisan effects once we recognize the implications of the partisan geography in Pennsylvania in ``naturally” tilting outcomes slightly in a pro-Republican direction because of differences in the degree of geographic concentration of each party’s supporters.
-(see below) Moreover, we show that the proposed Republican remedy plan can be characterized as a ``disguised” pro-Republican partisan gerrymander -- what we call a ``stealth gerrymander” -- which is nearly as pernicious in its expected pro-Republican partisan consequences as the 2011 plan whose effects it claims to ``remedy".
-\par
-    To give a more intuitive feel for how egregiously ill-compact the 2011 map is compared to the other alternatives we are considering, we show all four maps in the same scale, in a gray and white format.
-What is visually apparent from these maps is how aesthetically ``ugly" the 2011 map is compared to all of its alternatives.
-But it also clear that three remedy maps that are similar in having relatively few county cuts nonetheless can look quite different in terms of how each configures districts.
-% ================================================================= 
-% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
-    \footnote{This latter point is reinforced if we examine good government maps prepared by \href{https://www.dailykos.com/stories/2018/2/8/1739648/-Pennsylvania-will-soon-redraw-its-House-map-to-end-GOP-gerrymandering-How-would-you-draw-the-lines}{DailyKos} and by the present authors (available upon request).}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
 % =================================================================
 %
+\begin{document}
+    \authorone{Jonathan R. Cervas}
+        \institone{University of California Irvine}
+    \authortwo{Bernard Grofman}
+        \instittwo{University of California Irvine}
+    \titlerunning{Tools for Identifying Partisan Gerrymandering}
+    \subtitle{With an Application to Congressional Districting in Pennsylvania}
+    \status{Revise and Resubmit, Political Geography}
+    \data{https://bit.ly/2NaYLOY
+    \par
+    \centering{\textbf{Key Words}:  gerrymandering, Pennsylvania Supreme Court, redistricting, League of Women Voters, good government}}
+    \abstract{In League of Women Voters v.
+Commonwealth of Pennsylvania (2018) the Pennsylvania Supreme Court struck down as a ``severe and durable” partisan gerrymander the congressional map drawn by Republicans in 2011 and used in elections from 2012-2016.
+It did so entirely on state law grounds after a three-judge federal court had rejected issuing a preliminary injunction against the plan.
+After Pennsylvania failed to enact a lawful remedy plan of its own (due to total disagreement as to how to proceed between the newly elected Democratic governor and the still Republican-controlled legislature), the Court then ordered into place for the 2018 election a map of its own drawn for it by a court-appointed consultant.
+In a split court, the Court map was endorsed only by judges with Democratic affiliations.
+Here we compare the 2011 and 2018 congressional maps in terms of a variety of proposed metrics for detecting partisan gerrymandering.
+We also examine the remedy map proposed by a group of Republican legislators and that proposed by the Democratic governor.
+We conclude that the 2011 map was a blatant and undisguised pro-Republican gerrymander.
+Moreover, the remedy map proposed by Republican legislators was a covert pro-Republican gerrymander (what we refer to as a ``stealth gerrymander”).
+The Democratic governor's proposed plan cannot be classified as a pro-Democratic gerrymander and indeed has, if anything, a slight pro-Republican tilt.
+The 2018 court-drawn remedial map, by all measures, was not a gerrymander.}
+
+    \thanks{\href{https://www.jonathancervas.com}{Jonathan Cervas} is a Ph.D.
+candidate in political science at the University of California, Irvine with an interest in elections and skills in geographic information systems.
+When Grofman was the Special Master in service to the U.S.
+District Court of Utah, Cervas prepared under his direction remedial maps for the County Commission and the School Board in San Juan County (Navajo Nation et al v.
+San Juan County et al [12/21/2017]) When Grofman was the Special Master in service to the US District Court for Eastern Virginia, Cervas prepared under his direction remedial maps for Virginia House of Delegates districts (Bethune-Hill et al v.
+State Board of Election [02/14/2019]).
+Cervas’s work on this article was supported by the Peltason Chair at UCI and the UCI School of Social Sciences, with supplementary funding from the Peltason Center for the Study of Democracy at UCI.
+
+    \href{http://www.socsci.uci.edu/~bgrofman/}{Bernard Grofman} is the Jack W.
+Peltason (Bren Foundation Endowed) Chair of Democracy Studies and Distinguished Professor of Political Science at the University of California, Irvine.
+He has drawn remedial redistricting plans for federal courts and has testified as an expert in numerous redistricting cases over four decades.
+Most recently.
+he is the co-author of an Amicus Brief in two 2019 partisan gerrymandering cases that were heard by the U.S.
+Supreme Court, one from North Carolina and one from Maryland.
+Grofman’s work on this article was supported by the Peltason Chair at UCI, with supplementary funding from the Peltason Center for the Study of Democracy at UCI.
+\\
+
+    Questions or comments should be directed to \url{jcervas@uci.edu}.
+The authors would like to thank \href{https://www.linkedin.com/in/esther-bailey-9289391a/}{Esther Elizabeth Bailey}, \href{https://polisci.wisc.edu/people/faculty/david-canon}{David Canon}, \href{http://www.philly.com/archive/jonathan_lai/}{Jonathan Lai}, \href{https://github.com/modestbayes}{Lingge Li}, \href{mailto:psafarza@uci.edu}{Pooya Safarzadeh}, \href{https://www.faculty.uci.edu/profile.cfm?faculty_id=5443}{Charles Anthony Smith}, \href{https://twitter.com/aruhnv?lang=en}{Aruhn Venkat}, and \href{https://www.dailykos.com/user/Stephen Wolf}{Stephen Wolf} for their helpful suggestions.}
 % ================================================================= 
-% -- FIGURE -- FIGURE -- FIGURE -- FIGURE -- FIGURE -- FIGURE --   % 
-% ----------------------------------------------------------------- 
-% Maps Of The Proposed Pennsylvania Congressional Districts
- \begin{center} INSERT FIGURE \ref{fig:maps} ABOUT HERE \end{center}
-        \
-% -----------------------------------------------------------------
-% -- END FIGURE -- END FIGURE -- END FIGURE -- END FIGURE -- END FI %
+    % ------- TITLE PAGE ------------------------------------------
+        \maketitle
+        %\input{anon_title.tex}
 % =================================================================
-\par
-    Now we turn to a consideration of probable partisan consequences of the four maps. We begin with a discussion of the 2011 map.
-\par
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-\subsubsection*{Partisan results in actual elections held under the 2011 map}
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-%
-    The best evidence about a plan’s partisan consequences is, of course, evidence derived from actual elections under the plan -- if there are any -- though we do need to check if there are highly anomalous aspects to one or more of these elections which would make results from them hard to generalize.
-Also, the existence of incumbency advantage can have a distorting effect on the translation of votes into seats, since incumbents may deter strong challengers, thus reducing the vote-share of the minority party \citep{Jacobson_Kernell_1981, Abramowitz1991}.
-Nonetheless, simply from examining outcomes in the three elections held under the plan (2012, 2014, 2016), the claim that the 2011 Pennsylvania congressional plan is an egregious partisan gerrymander is very hard to refute since (a) regardless of voter preferences the plan returned 13 Republicans and 5 Democrats, with no district changing hands over the course of the three elections, and (b) in those three elections, Democrats won only 28\% of the seats (5 out of 18), despite winning a substantial share of the vote in all three elections.
-In 2012, the Democratic share of the average two-party vote in the congressional elections in Pennsylvania was 48.0\%, in 2014 it was 44.5\%, and in 2016 it was 45.8\%.
+\input{_maintext.tex}
 % =================================================================
-% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
-    \footnote{For consistency, we report the average vote share after imputing 0.25 and 0.75 for uncontested districts.
-In raw votes, Democrats won more votes than Republicans in 2012, but by taking the imputed average, it shows they won a minority of votes due to disproportionate uncontested districts.
-The Republican raw vote shares are 49.2\% in 2012, 55.5\% in 2014, and 54.1\% in 2016.}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
+% - Bibliography
+%     \input{Tables/_tab_summaries.tex}
+% \clearpage
+%     \begin{landscape}
+%         \input{Tables/_tab_congsum.tex}
+%     \end{landscape}
+% \clearpage
+%     \begin{landscape}
+%             \input{Tables/_tab_gerry.tex}
+%     \end{landscape}
+% \clearpage
+%     \begin{landscape}
+%         \input{Tables/_tab_prob.tex}
+%     \end{landscape}
+% % \clearpage
+%     \begin{landscape}
+%         \input{Tables/_tab_states_compare.tex}
+%     \end{landscape}
+\clearpage
+
+    % \input{Figures/_maps.tex}
+
+    \nolinenumbers
+    \newgeometry{
+        top=1in,
+        bottom=1in,
+        inner=0.85in,
+        outer=0.85in,
+        ignorehead,
+        ignorefoot,
+        nomarginpar,
+    }
+
+\begin{multicols}{2}
+    \footnotesize{\bibliography{library}}   % Bibliography linked to bib.bib
+\end{multicols}
 % =================================================================
-    Thus, the discrepancy between Republican vote share and Republican seat share was huge, and yet the seat share appeared to be immutable.
-\par
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-%        
-\subsection*{Asymmetry measures for partisan outcomes in the 2011 Plan based on statewide election outcomes}
-%
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-%
-    There are other indicators of statewide partisan gerrymandering effects that also provide a compelling picture of the 2011 map as an (enduring) partisan gerrymander.
-There are several metrics that have been proposed to look at the degree to which there is asymmetry in seats-votes relationships that might be indicative of partisan gerrymandering: (1) \textit{partisan bias} \citep{Tufte1973}, (2) the \textit{mean-minus-median} gap \citep{Mcdonald_Best_2015_ELJ} (3) the \textit{efficiency gap} \citep{Stephanopoulos2014_UofChicagoLaw}, and most recently (4) declination \citep{Warrington2018}.
-Each of these measures can be directly calculated for the 2011 map based on district election outcomes.
 % =================================================================
-% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
-    \footnote{Another recently proposed measure is a simplified version of partisan bias which lacks the advantage of separately calculating bias and responsiveness (swing ratio).
-It is calculated by taking the average two-party vote for the dominant party and adjusting it uniformly downward in each district so that the two parties receive exactly 50\% of the overall vote; then we look to see how far the seat share the dominant party now receives is from 50\% \citep{Wang2016_SLR}.}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
+% APPENDIX
 % =================================================================
-\par
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-\subsubsection*{Partisan Bias}
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-%
-    \textit{Partisan Bias} indicates asymmetry in the translation of each party’s votes into seats.
-Customarily, partisan bias is measured at the (hypothetical) point where each party’s share of the two-party vote is exactly 50\%.
-At that point, both parties should, by symmetry, receive identical seats shares.
-Thus, the difference between a party’s projected seat-share with 50\% of the vote and exactly 50\% can be taken as a measure of partisan bias.
-Note, however, that although a plan must be proportional at the ($50,50$) point, a partisan bias of zero does not imply proportional representational.
-What partisan bias taps is differential treatment of the parties?
-Thus, for example, if party R gets 56\% of the seats with 52\% of the vote, this is not a sign of bias as long as party L also can be expected to get 56\% of the seats were it to have 52\% of the vote.
-% ================================================================= 
-% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
-    \footnote{Although the basic idea goes back at least as far as \citet{Dahl1956}, the modern \textit{locus classicus} in political science for the idea of partisan bias(and the parallel concept of responsiveness/swing ratio) is \citet{Tufte1973}, with important recent work done by Gary King (see e.g., \citep{GelmanKing1994_unifiedAJPS}; and review in \citet{GrofmanKing2007}).
-A closely related line of research is found in the political geography literature (with \textit{locus classicus} \citet{Brookes1959, Brookes1960} -- with some further methodological improvements by scholars such as \citet{Johnston1994, Rossiter1997} and \citet{Johnston2002}.}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% =================================================================
-\par
-    There are a number of different ways to estimate partisan bias based on the shape of the votes to seats distribution \citep{Tufte1973, Grofman1983, Browning_King_1987_seats_votes, GelmanKing1994_unifiedAJPS, Grofman_et_al_1997_SwingRatio_Bias, Zingher2016_bias_swingratio_JEPP}.
-The method we use here is basically that of \citet{Tufte1973} with some additional complexity building in ideas of \citet{GelmanKing1994_unifiedAJPS} on simulating potential outcomes in a probabilistic fashion.
-We adapted the code from \citet{Kastellec_et_al_2008_PS}, which parallels that used by \texttt{JudgeIt}.
-It models inter-election shifts and adjusts for differences over time in vote shares among and between districts in a probabilistic fashion, using a simulation with 1000 repetitions (see more below).
-We then apply the following procedure to find the mean partisan bias and range of expected bias over the course of the decade.
-\par
-    Hypothetical elections are constructed by incrementally adding (or subtracting) one percentage point (on average) at a time to find aggregate seat outcomes under differing mean vote shares.
-The resulting ($s$, $v$) points on the votes-to-seats curve are then converted to a log odds form by using $ log(\frac{s}{1-s}) $ as the dependent variable, and $ log(\frac{v}{1-v}) $ as the independent variable.
-A regression on the transformed variables is then run, with transformed $s$ as the dependent variable, for $v$ values in some range near a 50\% vote share.
-% ================================================================= 
-% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------    
-    \footnote{Where to center the \textit{seats/votes} curve is an unresolved issue in the literature.
-In \citet{GelmanKing1994_unifiedAJPS} \texttt{JudgeIt} program, the choice is to center the election at the average Democratic vote share over recent elections.
-\citet{Kastellec_et_al_2008_PS}, similarly, simulates data for average vote shares between 45\% and 55\%.
-We checked for consistency between the methods by running the numbers in three ways: centering the simulated outcomes at an average election vote share, centering the simulated outcomes at a tied 50/50 election, and centering the simulated outcomes at the actual election results in the particular election.
-All three variations provide similar estimates, as do the results from \texttt{JudgeIt}.
-Readers should note that with only 18 districts, too much weight should not be place exact bias coefficients reported [see the discussion in \citet{Browning_King_1987_seats_votes}]; but the relative magnitude of the bias values across the different plans is fully congruent with the evaluations of probable partisan gerrymandering yielded by the other measures.
-Because of the limited number of districts in the plans, the same caution is urged for all the measures of gerrymandering reported in this essay.}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% =================================================================
-    \textit{Partisan Bias} is then calculated from the intercept of this regression using an exponential transformation (for details see \citet{Grofman1983}).
-Partisan bias is calculated from the Republican perspective, and \underline{negative} values indicates bias against Republicans.
-\par
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-        \subsubsection*{Mean-Median Gap}
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-    When districts are sorted according to two-party vote share, the \textit{Mean-Median Gap} is the difference between the average vote percentage for a party and its vote share in the median district \citep[see][]{Best2018}.
-This measure is a variant of \textit{skewness}.
-When the mean is substantially higher or lower than the median, this is taken to be indicative of partisan bias.
-\par
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-        \subsubsection*{Efficiency Gap}
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-    The \textit{Efficiency Gap} is calculated as defined in \citet{Stephanopoulos2014_UofChicagoLaw}, where all the party’s votes are wasted if they lose the district, and all the winner’s votes over 50\% are wasted.
-The difference between each party’s wasted votes is then divided by the total votes cast to produce the \textit{Efficiency Gap}, with a value of zero denoting what is regarded as ideal.
-As noted in \citet[][p.
-13]{Best2018}, this is equivalent to taking an aggregate responsiveness (swing ratio) of two as ideal.
-\par
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-        \subsubsection*{Declination}
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-    Among the more recent measures of gerrymandering is the declination.
-Declination is an astronomical term referring to the angle on a compass, with a Latin root meaning ``a bending away".
-For gerrymandering, it uses the angles created from ordering all of the districts by vote share, computing the mean vote-share for each party separately for the seats they won, and then comparing the differences in their distribution.
-To compare, a line is draw from the mean vote-share to the 0.5 line separately for each party.
-The angle formed from the different slopes of the line is called the declination.
-An angle indicates whether a plan is a gerrymandering, under the logic that vote-shares should be distributed in some uniform manner across districts.
-If partisan are \textit{packed}, the angle will be greater.
-All values range between -1 and 1.
-Though we have defined the vote-share in terms of Republican outcomes, we adjusted the values such that \underline{negative} values are favorable to the Democrats.
-Further information can be found in \citet{Warrington2018}.
-\par
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-        \subsubsection*{Measuring Gerrymandering}
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-    We emphasize that a partisan gerrymandering claim based on political consequences should also provide evidence that the (a) gerrymandering indicators that are found cannot be explained simply by the geographic distribution of each party’s electoral support (e.g., differential concentration of electoral support can lead to what has been called a ``natural gerrymander"), and (b) that the partisan asymmetry in translating votes into seats is not due to chance alone (e.g., many highly competitive seats won by very narrow margins) but can be expected to be durable.
-With only 18 districts, sensitivity to chance effects is particularly important.
-\par
-    Rebuttal to any claims that the partisan asymmetries observed in the consistent outcomes of the 2011 map can be attributed primarily to differential concentration of party voting strengths, or to chance, have been decisively refuted by the expert witness testimony in the case (see esp.
-the simulations of potential good government maps done by Professor Chen, discussed in detail in the Pennsylvania Supreme Court opinion in the case, and on which the Court placed great reliance).
-With respect to the durability of the effects of the 2011 map, we have already pointed out that this plan yielded unchanged 13R-5D results for three elections in a row.
-To help explain this over time consistency we note that the 2011 Map has primarily safe seats (13 of 18 won with greater than a 10\% margin).
-But, indicative of the pro-Republican gerrymander achieved with the 2011 map, those safe seats are distributed in a highly asymmetric fashion, with nine essentially safe seats for Republicans and only four for Democrats.
-%        ================================================================= 
-%        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% ------------------------------------------------------------------------
-        \footnote{However, we should note that some authors \citep[e.g.,][]{Grofman2019_ELJ} have argued that a partisan gerrymandering claim at a jurisdiction-wide level must also be reinforced by district specific evidence of manipulation of particular district boundaries.
-As noted earlier, we do not discuss this type of district specific evidence here, other than to point out that such evidence was presented at the trial, and that this evidence was used to support the Court’s finding that the Pennsylvania congressional map was an unconstitutional gerrymander under Pennsylvania state law.}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% =================================================================
-\par
-    We show in Table \ref{tab:congsum} the results for the 2011 plan found unconstitutional.
-Results of the four measures for the 2011 plan are based on using the average of actual congressional results for the 2012-16 period under that plan.
-\underline{Positive} values favor Republicans (add to their seat share), while \underline{negative} values favor Democrats (subtract from Republican seat share).
-\par
-    In our view, none of these four measures standing alone can be taken to be proof of gerrymandering.
-Rather we see them as potentially reinforcing indicators.
-Our own analyses have found hypotheticals where these measures can generate either a false positive or a false negative, but we have also found that consistent bias in the same direction is unlikely when we use multiple indicators (data omitted for space reasons).
-Here, however, all four indicators, like the actual election results, point clearly and overwhelmingly in the same direction vis-à-vis the 2011 Enacted plan -- namely that it is a blatant pro-Republican partisan gerrymander.
-%
-% =================================================================
-\subsection*{Evaluating Partisan Gerrymandering using Statewide elections to provide a normal vote baseline}
-% =================================================================
-%
-    In order to assess expected partisan consequences for plans that have not yet been implemented we need to develop a predictive equation, and that is most readily done with data on partisan outcomes for state-wide elections projected into the new districts.
-For example, Professor Jowei Chen in his testimony in the Pennsylvania case used a composite of the six state-wide elections between 2008 and 2010 to estimate the political consequences of alternative maps.
-After extensive investigation of alternative prediction models and their accuracy in reproducing the partisan outcomes of the 2012, 2014, and 2016 congressional elections at the district level (data omitted for space reasons), we have opted for a model that takes the average of five state-wide 2016 elections that reflect a balance of Republican and Democratic success: President, U.S.
-Senate, PA Attorney General, PA Treasurer, and PA Auditor.
-Republicans won the statewide vote in two of them (President, U.S.
-Senate), while the Democrats won the statewide vote in three of them (Auditor, Treasurer, and Attorney General).
-The average vote for the Republicans across these five elections was 49.1\%.
-%        ================================================================= 
-%        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% ------------------------------------------------------------------------
-        \footnote{We compiled the composite measure using both the sum of the raw votes by party and an average of the two-party vote in the five races at the district level.
-Both ways result in the same district partisanship breakdowns in all plans.}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% =================================================================
-\par
-    We use these five elections to calculate the same four previously introduced metrics but now for all four plans.
-%        ================================================================= 
-%        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% ------------------------------------------------------------------------
-        \footnote{We report the Republican share of the two-party vote ($\delta_{i}$).
-Districting plans are represented by $\mathcal D $, (e.g., $\mathcal D_{enacted}$, $\mathcal D_{remedial}$, \dots, $\mathcal D_{j}$), and each election in year $ y $ has a district vote distribution $ \Delta_{y} = [\delta_{y1}, \delta_{y2}, \dots, \delta_{yi}] $.
-To find the overall state-wide vote, we calculate the average district vote share, $ \bar{\Delta}_{\mathcal D y} = \frac{1}{n}\sum\limits_{i=1}^{n} \delta_{yi} $.
-By averaging, we reduce the influence of turnout variation among districts.
-This district-specific average is a useful state-wide estimate of voter sentiment though it is in part effected by the particular configuration of districts being examined \citep[see e.g.][cf.
-\cite{Grofman_et_al_1997__IntedgratedPerspective_ES}]{Kastellec_et_al_2008_PS}.
-In particular, incumbency advantage effects affect our ability to accurately predict vote outcomes.
-In ancillary work not reported here, we have looked at changes in congressional incumbency advantage in Pennsylvania over multiple decades.}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% =================================================================
-    We aggregate voting district level (precincts) data to the congressional district level.
-%        ================================================================= 
-%        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% ------------------------------------------------------------------------
-		\footnote{We wrote a function in \texttt{R} to assign all VTDs to a district based on the centroid of the VTD.
-This method creates a slight difference between it and the actual vote, but averages just -0.06\% per district.
-Most of the mis-allocation is between two adjacent districts, District 1 and District 13, both overwhelmingly Democratic districts.}
-% ------------------------------------------------------------------------ 
-%        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-%        =================================================================
+\setcounter{section}{0} \renewcommand{\thesection}{Appendix \Alph{section}}
 
-\par
-    \begin{center}    \textbf{US House Election Summaries (PA 2004-2016)} \par \end{center}
-    \begin{center}   \textbf{[ INSERT TABLE \ref{tab:congsum} ABOUT HERE ]} \end{center}
-% ••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••• 
-\par
-    We do need to check that our five-election composite offers a plausible basis for a prediction model once we take into account electoral tides. 
-%        ================================================================= 
-%        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% ------------------------------------------------------------------------
-		\footnote{We can only validate our measure for the 2016 election and only for the 2011 enacted plan.
-For all the remedial plans we are analyzing, the composite is our ``best guess".
-We have to make some assumptions about state and district inter-election swings.
-For simplicity, we assume that all districts have a national shift equal to the state-level swing, ie, difference from the 2016 elections.
-Later in the paper, we conduct simulated elections that vary the district level variability.}
-% ------------------------------------------------------------------------ 
-%        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-%        =================================================================
-    Encouragingly, our five-election composite results is the same 13R-5D result that the 2016 congressional races delivered.  
-%        ================================================================= 
-%        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% ------------------------------------------------------------------------
-		\footnote{Whether we take the sum of the five elections or average the five elections, both methods deliver the same relative distributions.}
-% ------------------------------------------------------------------------ 
-%        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-%        ================================================================= 
-    Additionally, each district's outcome is identical to the 2016 congressional district results. 
-\par
-    But we do wish to do a further robustness check on the plausibility of using our five election composite to calculate the various metrics designed to measure potential gerrymandering.
-One way to do this is to see how well the composite (calculated at a hypothetical 50\% vote share) mimics the results of the values for the four metrics for the 2011 map that we previously obtained using the actual congressional election results for that map.
-What we are doing is comparing a hypothetical ``normal vote” \citep{Converse1966} assumption (based on a statewide vote in which we are not taking incumbency effects into account) with the average of what we actually get in three different congressional elections.
-The results in Table \ref{tab:congsum} are 0.11 for bias, 0.16 for the efficiency gap, 0.06 for the value of the mean minus median gap, and 0.40 for declination.
-The comparable results from Table \ref{tab:gerry} are 0.09, 0.25, 0.05, and 0.40.
-It is apparent how close these two sets of values are to one other even though one is based on actual elections with a mean Republican (two-party) vote share close to 53\% while the other is based on projected election results set to yield a 50-50 two-part vote distribution.
-If instead we calculate these four metrics using the average five election statewide vote in these elections over the period 2012-2016, a Democratic vote share of 49.1\% we get virtually identical values: 0.09, 0.24,  0.05, 0.36.
-%        ================================================================= 
-%        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% ------------------------------------------------------------------------
-		\footnote{If we center the simulations at the 2012-2016 average statewide Republican vote of 53.3\% of the vote, we again get remarkably consistent estimates [0.09, 0.19, 0.05, 0.46 ].}
-% ------------------------------------------------------------------------ 
-%        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-%        =================================================================    
-\par
-    Our confidence in the five election projection method is further buttressed because, when we applied this method to predict the results of the 2018 election under the court-ordered remedial map.
-Although the method had been developed prior to the 2018 election, using data no more recent than 2016, and not including any incumbency advantage, it predicted that the court-ordered remedial plan would have produced an outcome of 7R-11D.
-%        ================================================================= 
-%        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% ------------------------------------------------------------------------
-        \footnote{After uniformly shifting each district's composite such that the statewide average equalled the actual 2018 statewide average.
-The large range is indicative of the number of competitive districts that Republican candidates won.
-Had we modeled incumbency advantage instead of the `normal vote', the two Republican incumbents who won re-election by less than 2 percentage points, along with the one who won by less than 3 percentage points would have made the actual 9R-9D outcome the likely predicted outcome.}
-% ------------------------------------------------------------------------ 
-%        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-%        =================================================================
-%
-% =================================================================
-% -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE  % 
-% -----------------------------------------------------------------
-\input{Tables/_tab_gerry.tex}
-\begin{center}\textbf{INSERT TABLE \ref{tab:gerry} ABOUT HERE} \end{center}
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-%
-% =================================================================
-% -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE  % 
-% -----------------------------------------------------------------
-\input{Tables/_tab_prob.tex}
-\begin{center}\textbf{INSERT TABLE \ref{tab:prob} ABOUT HERE} \end{center}
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-%
-\par
-    The findings from Table \ref{tab:gerry} are very clear; Regardless of which metric we examine, the 2011 map is the most gerrymandered.
-Not surprisingly, the nature of the bias is in a pro-Republican direction.
-Given the cumulative weight of all the evidence, the 2011 congressional map was clearly an egregious pro-Republican gerrymander.
-In contrast, as we might expect, the court-ordered plan, prepared by a respected expert whose instructions were to satisfy good government criteria, was by a considerable margin the closest to a perfectly neutral plan in partisan terms according to all four measures.
-\par
-    Finally, we turn to the plan offered by the Democratic governor, Governor Wolf.
-To our surprise, this plan turned out also to have a pro-Republican tilt, nearly to the same extent as the joint legislative plan for three of the four metrics in Table \ref{tab:gerry}.
-This puzzling finding suggests that the Governor might have been under pressure from Democratic incumbents not to reduce their expected vote margins, or that he sincerely intended it to be a compromise that the Republican legislature might agree to.
-However, there is one important difference between the joint plan and that of the Governor: among the seven districts likely to be won by Democrats, the Democratic percentages are cut more narrowly in the Republican plan than in the Governor’s plan in three of the districts so that, in a good Republican year, Republicans will do better under their plan than under the Democratic plan.
-\par
-    Still, when we look at Table \ref{tab:gerry} we are struck by the fact that all four plans, even the Court-ordered one, still appear to have some level -- very slight in one case, extreme in others -- of pro-Republican bias.
-In the next section we consider why this might have happened, and we introduce a potentially important distinction between ``\textit{neutral plans}" and ``\textit{fair plans}."
-\par
-    In sum, looking at the evidence above, we see exactly the same ordering of plans according to their characterizability as a partisan gerrymander as we saw when we ordered plans according to the degree to which they satisfied good government criteria (see Table \ref{tab:summaries}), with the 2011 map the worst and the 2018 court-ordered map far and away the best.
-This conclusion is further buttressed by the data we present in Table \ref{tab:prob}.
-Here we create a probabilistic simulation using statewide five-election composite (set to a 50-50 vote share) to estimate the likelihood that random shocks (based on past state inter-election shift data) would change party control of the congressional delegation.
-\par
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-    \subsection*{Natural Gerrymanders}
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-    In 2018, even under a court-drawn plan, in a year when Democrats had won a clear majority of the aggregate two-party congressional vote, Democrats only succeeded in winning exactly 50\% of the seats.
-Trying to understand why all four of the plans we considered had some level of pro-Republican bias requires us to look in more detail at the partisan electoral geography of Pennsylvania.
-In Pennsylvania, Philadelphia is overwhelmingly Democratic in voting.
-In particular, if you draw two congressional districts entirely within Philadelphia County, one of them is very likely to give Democratic candidates around 90\% of the vote, and for sure, as long as both are wholly within the County, the average vote in the two will be around 80\% Democratic no matter how you draw the two districts.
-There are no other large, concentrated pockets of equivalently overwhelmingly Republican voting strength.
-Thus, more Democratic votes will \textit{naturally} be ``wasted" in Philadelphia than Republican votes will be ``wasted" elsewhere in the state.
-Additionally, the city of Philadelphia is landlocked, and its position on the state's border reduces the ability of line drawers who favor Democrats to distribute these voters into multiple districts so as to minimize the extent to which Democratic votes in the county are wasted.
-Additional Democratic wasted votes come in heavily Democratic Allegheny County, the county in which the city of Pittsburgh is located.
-However, saying that urban concentration of Democrats means that it is harder for Democrats to turn their votes into seats than is the case for Republicans does not mean that we cannot distinguish the effects of such \textit{natural} gerrymanders from \textit{intentional} political gerrymandering, since the latter generates perverse consequences for Democratic voters to a far greater extent than the likely one district or so penalty created by the nature of Pennsylvania’s present electoral geography.
-%        ================================================================= 
-%        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% ------------------------------------------------------------------------
-        \footnote{\citet{Chen2013} estimate this bias as about 1.45 seats (8\%) but Chen’s expert witness testimony in the Pennsylvania cases generates a lower estimate of natural bias until one builds in the need to protect minority opportunity districts.
-Here, we need to be careful since the minority population in such districts in the enacted map may be larger than actually needed to assure effective minority representation, and the act of maintaining or even increasing high levels of minority population in the district may be a disguised way of packing Democratic voters so as to advantage the interests of the Republican party.}
-% ------------------------------------------------------------------------ 
-%        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-%        ================================================================= 
-    For example, in the first set of 500 map simulations in his expert witness report, based only on traditional redistricting criteria, but not including voting rights protections, Professor Chen found that none resulted in the 13R-5D split as happened in 2012 and subsequent years.
-The mode was 9R-9D.
-%        ================================================================= 
-%        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% ------------------------------------------------------------------------
-        \footnote{Expert Witness Report, Jowei Chen (Pg. 15, 16, Figure2).}
-% ------------------------------------------------------------------------ 
-%        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-%        =================================================================
-\par
-    Recognizing the potential for so-called ``natural gerrymandering", we wish to highlight a potentially important distinction between \textit{neutral} plans and \textit{fair} plans -- each of which reflects a different approach to defining the baseline for what constitutes gerrymandering.
-Neutral plans refer to those that are drawn entirely with respect to traditional good government criteria, with no attention paid to partisan considerations.
-One way to define partisan gerrymandering is with respect to a baseline defined by the set of feasible neutral plans -- and this was how one of the plaintiff’s experts in the Pennsylvania case, Professor Jowei Chen \citep[cf.][]{Chen2015_JOP} shaped his testimony in the Pennsylvania case, and in other cases in which he has testified.
-%        ================================================================= 
-%        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% ------------------------------------------------------------------------
-        \footnote{Similarly, \citet{Grofman2018_ELJ} asserts that ``Gerrymandering occurs when a districting plan creates a disparate treatment of the vote share of the minority and majority voting blocs in a way that penalizes the minority in its ability to translate its voting support into seats compared to what we might expect from a plan drawn on the basis of neutral principles."}
-% ------------------------------------------------------------------------ 
-%        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-%        ================================================================= 
-\par
-    In contrast, we may wish to use the term ``fair" as plans that seek to compensate for the difference in level of partisan concentration.
-Fair plans do not seek to achieve proportional representation.
-Rather, they seek to achieve partisan symmetry.
-A fair plan would yield a proportional result when the vote shares are evenly split 50-50, but at points beyond 50-50, such as when party R wins 70\% of the districts with only 55\% of the vote, party L can similarly win 70\% of the districts if it had received 55\% of the vote.
-\par
-    One clear finding of our analysis of the court-ordered map is that it was drawn as a ``neutral map", though not necessarily a fair map in the sense we are using the term above.
-In order to create a truly ``fair map", in our view, it would have been necessary to break up Philadelphia County in more than three pieces so as to avoid the ``natural" packing of Democratic voters.
-%        ================================================================= 
-%        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% ------------------------------------------------------------------------
-    \footnote{Note that we certainly not advocating that such a three way split of Philadelphia should have been done.}
-% ------------------------------------------------------------------------ 
-%        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-%        ================================================================= 
-    But the Pennsylvania Supreme Court opted not to do that, but to instead preserve county lines and draw a good government map. We therefore identify the Court remedial plan as a ``neutral" plan.
-\par
-%
-% =================================================================
-% -- SECTION -- SECTION -- SECTION -- SECTION -- SECTION -- SECTI %
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-\section{Conclusions and Lessons for the future; The Potential Impact of LWV}
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-    While the Pennsylvania court opinion is limited to Pennsylvania, and thus it might have seemed of importance only in Pennsylvania, in footnote 71 of the Opinion (slip op.
-pp.116-117), the court took what we regard as a rather unusual step.
-It issued what can only be called an invitation to other state courts to use the same logic it used to invalidate partisan gerrymanders in their own state.
-As we noted earlier, the Court pointed out that there are twelve additional states whose constitutions contain election clauses identical to the Pennsylvania charter, requiring elections to be ``free and equal”: Arizona, Arkansas, Delaware, Illinois, Indiana, Kentucky, Oklahoma, South Dakota, Oregon, Tennessee, Washington, and Wyoming.
-% ================================================================= 
-% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
-        \footnote{In its Opinion, the Court provided specific citations to each of these provisions which we have not reproduced.}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% =================================================================
-    However, only a handful of these states are ripe for partisan gerrymandering challenges -- in that (ca.
-2017) three lack unified party control of the state, two are single district states, and some already have a commission drawing plans and in others, \textit{indicia} of gerrymandering are missing.
-\par
-    In looking to the future, we should also note that some states that are generally regarded as among the most pernicious partisan gerrymanders, Michigan, Ohio, North Carolina, Wisconsin are not included among the twelve, nor is Maryland.
-On the other hand, we should also note that a ``''free and equal” elections clause is not the only avenue state courts might use to attack partisan gerrymanders in the future.
-As University of Kentucky College of Law Professor Joshua Douglas has pointed out, virtually every state constitution protects voting rights more explicitly than the U.S.
-Constitution does \citep{Douglas2014_RightToVote}.
-In addition to the thirteen states that require elections to be ``free and equal”, an additional thirteen have state constitutional provisions that require elections to be ``free" or ``free and open”, and these clauses could, in principle, be used in exactly the same way as the ``free and equal” clause.
-% ================================================================= 
-% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
-        \footnote{We are indebted to Jonathan Lai of the Philadelphia Inquirer (personal communication, April 2018) for calling this information to our attention.
-See a more detailed discussion in \citet{Elmendorf2018}.}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% =================================================================
-    Such clauses are found in a number of the states widely regarded as having the worst congressional gerrymandering. 
-% ================================================================= 
-% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
-        \footnote{These states include Arizona, Arkansas, Colorado, Delaware, Illinois, Indiana, Kentucky, Maryland, Massachusetts, Missouri, Montana, Nebraska, New Hampshire, New Mexico, North Carolina, Oklahoma, Oregon, Pennsylvania, South Carolina, South Dakota, Tennessee, Utah, Vermont, Virginia, Washington, and Wyoming \citep[][footnote 86]{Douglas2014_RightToVote}.}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% =================================================================
-\par
-    Accordingly, some scholars have suggested turning to turn away from the federal judiciary altogether and instead focusing on state-level litigation \citep{Wang_et_al_2019_Labortories_UPJCL}.
-We believe that the \textit{LWV} decision and the further comparative analyses of alternative plans presented in the previous sections of this paper will help other state courts navigate their way to decisions that strike down some partisan gerrymanders as unconstitutional, while allowing others to remain in place on the grounds that either they are not that severe, or they are unlikely to be lasting in that they have sufficiently many competitive seats as to be responsive to realistic changes in voting patterns.
-Now that litigation has been successful in state courts on the grounds of egregious deviations from good government districting criteria, it seems plausible that in the next wave of redistricting, mapmakers may create covert gerrymanders like we've described in this essay.
-If so, there remains the issue of whether state courts have the legal tools to deal with these ``stealth gerrymander”.
-\par
-    According to all four measures, the joint legislative plan offered by Republican leaders is nearly as much a gerrymandered map -- in a pro-Republican direction -- as is the 2011 map.
-As noted above, we believe it appropriate to label this plan as a \textit{stealth gerrymander} since it creates partisan assymetries while (mostly) providing the visual appearance of a good government map (see Figure \ref{fig:maps}).
-However, having labeled the joint legislative plan a \textit{stealth gerrymander}, it is worth reminding readers of two important points.
-First, had this plan been the original plan adopted by Republicans, it is very much an open question as to whether it would have been rejected by the Pennsylvania Supreme Court.
-Even though, in partisan terms, it is nearly identical to the 2011 plan, it is considerably more consistent with good government criteria (see Table \ref{tab:summaries}) -- not that much worse than the Court Remedial plan -- that it would have required detailed political election analyses like those we have done here to demonstrate its \textit{stealth gerrymander} features, plus a court willing to rule on political grounds, rather than good government grounds that it was unconstitutional.
-Because the joint legislative plan was not passed by the State of Pennsylvania's legislature, the Court felt no need for particular deference to it (or to the Governor's plan, for that matter).
-Thus, the Court was not bound by the normal supposition with respect to districting that a plan authorized by the state need not be the ``best possible", but only unconstitutional.
-That ``deference" to legislative judgments and the criteria they reflect, might well have tipped the balance toward acceptance of the Republican joint legislative plan were it offered as having been fully sanctioned by the State of Pennsylvania (the duly elected legislature and Governor).
-% ================================================================= 
-% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% -----------------------------------------------------------------
-        \footnote{The Court picked the plan among those before it that most closely satisfied good government criteria -- which happily, thanks to Professor Persily's expertise as consultant to the Court, turned out to be its own plan, and thus a plan which the Court could know with certainty was not intended as a partisan gerrymander for either party.
-Whether it would have still sought to maximize compliance with good government criteria had the plan which did so been a ``stealth gerrymander" is not a question we can answer.}
-% ----------------------------------------------------------------- 
-% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% =================================================================
-    On the other hand, in the presence of this \textit{stealth gerrymander} rather than the blatant gerrymander that came before it, the Pennsylvania Court might well have relied on partisan impact evidence to reach a conclusion of unconstitutionality.
-The language of the Final Order implementing the Court's own plan strongly suggests this possibility.
-There, the Court said about the 2011 plan that it
-% =====================================================================
-    % ------- QUOTE ---------------------------------------------------
-    \begin{quote}
-        ``was designed to dilute the votes of those who in prior elections voted for the party not in power in order to give the party in power a lasting electoral advantage.
-In stark contrast, Article I, Section 5 of the Pennsylvania Constitution provides: ‘Elections shall be free and equal; and no power, civil or military, shall at any time interfere to prevent the free exercise of the right of suffrage.’ Pa.
-Const.
-art.
-I, § 5.
-On this record, it is clear that the 2011 Plan violates Article I, Section 5, since a diluted vote is not an equal vote" (slip op.
-p.
-2).
-    \end{quote}
-    % ------- END QUOTE -----------------------------------------------
-% =====================================================================
-\par
-	Thus, we have some optimism that state courts can provide a venue to check partisan gerrymandering, even when it comes in the form of a \textit{stealth gerrymander}.
-But we have also placed a number of caveats on any claim that state courts can compensate fully for any continued failure of the federal courts to act.
-Citizen-based reforms such as referendum and initiatives may also help with reducing egregious gerrymandering in states where the passage of such measures is feasible.
-Ultimately, there is no substitute for a federal standard that guarantees each citizen in the United States \textit{equal protection under the law}.
-Absent the United States Supreme Court setting this standard, state courts offer one promising pathway towards ending undemocratic partisan gerrymandering.
+    %     \clearpage
+    % \input{Tables/a_tab_districtvotes.tex}
+    %      \clearpage
+    % \input{Tables/a_tab_congress_predict.tex}
+    %     \clearpage
 
-% 
-% 
-% 
-% 
-%
-
-% % ================================================================= 
-% % -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% % -----------------------------------------------------------------
-%         \footnote{Because of its high number of competitive seats, a point prediction such as 9R-9D is somewhat misleading for the court-ordered map.
-Using the five-election model as our predictor, there are fewer safe Democratic seats (5) than safe Republican seats (7) in the Court plan, using 45\%-55\% definition of a competitive seat.
-This is one fewer safe seat for Republicans and one additional safe Democratic seat than is found in the 2011 plan, but the Court plan also has five districts balanced on a knife edge, with a projected winner’s margin under two percentage points, making it highly responsive to shifting voter sentiment.
-This is realized in our simulations where the mean seat share, though near 9R-9D, is either lower or higher depending on where the state-wide vote is centered.}
-% % ----------------------------------------------------------------- 
-% % -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% % =================================================================
-
-% 	Past congressional results alone often are departures from the underlying district partisanship several reasons.
-The vote share of an incumbent has historically been greater than it is for a non-incumbent or two non-incumbents running for an open seat.
-
-% %        ================================================================= 
-% %        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% % ------------------------------------------------------------------------
-% 		\footnote{A long literature in political science has evaluated and confirmed the presence and increased incumbency advantage \citep{Abramowitz1991,Levitt1997,Ansolabehere2002b,Gelman_Huang_2008}.
-Recent scholarship, however, has shown a decrease in the advantage towards zero \citep{Jacobson2015_JOP}.}
-% % ------------------------------------------------------------------------ 
-% %        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% %        =================================================================
-
-%     A well-financed incumbent tends to suppress quality competition \citep{Jacobson2015_JOP}, vulnerable incumbents who can sense the political winds opt against running for re-election \citep{Jacobson_Kernell_1981,Ansolabehere_Snyder_2004_LSQ}, and electoral tides and varying turnout can distort the state-wide average.
-Midterm elections often have vastly different electorates than presidential elections \citep{Abramowitz_et_al_1986_AJPS}, while the midterm pattern of ``surge and decline" (what goes up must come down principle) reduces the sitting Presidents' party's vote share \citep{CampbellA1960}.
-%  \par
-% 	What we are after is a measure of the ``normal vote''.
-The normal vote is the underlying partisanship of a district \citep{Converse1966}.
-It is sort of a mystical number since it is uncovered and unobserved, but powerful theoretically.
-Map builders implicitly assume a normal vote when constructing a district plan.
-We believe that state-wide elections offer the most useful approximation of the normal vote.
-% %        ================================================================= 
-% %        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% % ------------------------------------------------------------------------
-% 		\footnote{We particularly believe this is true when the votes between elections are highly correlated, and the district-level variation is low, as it was in Pennsylvania from 2012 to 2016.
-The more stable the electorate is, the more confidence we have that state-wide elections are revealing the normal vote \citep{Ansolabehere_et_al_2000_AJPS}.}
-% % ------------------------------------------------------------------------ 
-% %        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% %        =================================================================
-% 	 All districts vote on the same candidates; hence outcome affects all voters equally.
-The effects of incumbency can distort the vote in congressional races in a way that doesn't resemble the actual partisan preferences at the district-level.
-State-wide elections eliminate the need to impute or otherwise deal with uncontested districts.
-Taking an average of the state-wide elections provides additional reason to believe it's capturing a normal vote due to a regression to the mean effects.
-Justices in \textit{Vieth} offered skepticism in using elections that have contrasting results; we do not share this view, instead seeing it as an important source of variation that helps to recognize the underlying partisanship in which \textit{should} be used in measuring bias in districting plans.
-Additionally, averaging over multiple races and years helps to reduce the bias imposed by candidate effects.
-% \par
-% 	We construct the five-election composite of state-wide elections by aggregating voting district level (precincts) data to the congressional district level.
-
-% 	Additionally, each district's outcome is identical to the 2016 congressional district results.
-Indeed, after accounting for incumbency, our composite measure is remarkably accurate in its predictions of the actual congressional results validating our measure.
-The average difference between our composite and the congressional results was just 2.16\%.
-
-% %        ================================================================= 
-% %        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% % ------------------------------------------------------------------------	
-% 	We calculate this average only for districts contested in 2016, since uncontested results have significant deviations.
-Including the uncontested districts increases this mean difference to just 3.93\%.
-We believe our composite measure offers better representation of the true partisanship in these districts.
-Individually, only the Auditor's race correctly predicts all the districts correctly, but by either summing or by taking the average of the five races, both methods correctly predict each district's outcome.
-% % ------------------------------------------------------------------------ 
-% %        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-% %        =================================================================
-
-%     We imputed the values 25\% and 75\% for uncontested or essentially uncontested races.
-Rather than taking the projected outcome we arrive at from this method as a single value, we looked at the sensitivity of this projection to the empirical range of inter-election swing, simulating outcomes 1,000 times.
-To arrive at our estimate for inter-election variation that can't be attributed to the normal vote or incumbency advantage, we estimate the equation using congressional elections in Pennsylvania in the 2004-16 time period:
-%         \begin{align*}
-%         \mathrm{Vote}_{ty} = \alpha + \beta_{1} \mathrm{Vote}_{yt-1} + \beta_{2} \mathrm{Incumbency}_{yt} + \beta_{4} \mathrm{Uncontestedness}_y{t} + \epsilon
-%         \end{align*}
-%     We are interested in $\epsilon$, which is the unexplained variance, which includes inter-election swing.
-We run this equation separately for each of the non-redistricting years and then average the coefficients.
-We estimate incumbency advantage to average 4.27\% for this time period, while inter-election variance is 4.46\%.
-When simulating elections, we use the composite as the normal vote, but add in for each district a random draw from a normal distribution centered at the district vote with a standard deviation of $\epsilon$.
-We can then calculate statistics for each of these simulated elections to find a distribution for each measure of gerrymandering.
+\end{document}

### 2019-06-18 Latex/_maintext.tex
Update _maintext.tex

--- 2019-06-18 Latex/_maintext.tex
@@
 % •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-    To facilitate comparisons to the two most politically important remedy plans offered to the Court, the joint plan proposed by the Republicans leaders of the two legislative chambers, and the Governor’s proposed plan as a Democratic alternative, we look at a number of features in each of the four plans as a whole.
-We do not, however, discuss district-specific criteria for identifying gerrymandering in these maps, since, although some district-specific evidence was presented about the 2011 map in the trial testimony by an expert who was highly knowledgeable about communities of interest in the state, the Pennsylvania Supreme Court opinion itself offers a \underline{statewide} standard for a finding of unconstitutional partisan gerrymandering.
+    To facilitate comparisons of the 2011 enacted and 2018 court plans to the two most politically important remedy plans offered to the Court, the joint plan proposed by the Republicans leaders of the two legislative chambers, and the Governor’s proposed plan as a Democratic alternative, we look at a number of features in each of the four plans as a whole.
+We do not, however, discuss district-specific criteria for identifying gerrymandering in these maps, since, although some district-specific evidence was presented about the 2011 map in the trial testimony by an expert who was highly knowledgeable about communities of interest in the state, the Pennsylvania Supreme Court opinion itself offers a \underline{statewide standard} for a finding of unconstitutional partisan gerrymandering.
    % ================================================================= 
@@
         \footnote{As noted earlier, given the language in the U.S.
-Supreme Court’s remand of the \textit{Gill} case, it appears that any federal standard for \underline{partisan gerrymandering} would need to be district specific.}
+Supreme Court’s remand of the \textit{Gill} case, it appears that any \underline{federal} standard for \underline{partisan gerrymandering} would need to be district specific.}
 % ----------------------------------------------------------------- 
@@
 the simulations of potential good government maps done by Professor Chen, discussed in detail in the Pennsylvania Supreme Court opinion in the case, and on which the Court placed great reliance).
-With respect to the durability of the effects of the 2011 map, we have already pointed out that this plan yielded unchanged 13R-5D results for three elections in a row.
+With respect to the durability of the effects of the 2011 map, this plan yielded unchanged 13R-5D results in all three elections where the map was used.
 To help explain this over time consistency we note that the 2011 Map has primarily safe seats (13 of 18 won with greater than a 10 percentage point margin).
@@
 \par
-    There are several metrics that have been proposed to look at the degree to which there is asymmetry in seats-votes relationships that might be indicative of partisan gerrymandering: (1) \textit{partisan bias} \citep{Tufte1973}, (2) the \textit{mean-minus-median} gap \citep{Mcdonald_Best_2015_ELJ} (3) the \textit{efficiency gap} \citep{Stephanopoulos2014_UofChicagoLaw}, and most recently (4) \textit{declination} \citep{Warrington2018} (details below).
+    There are several metrics that have been proposed to look at the degree to which there is asymmetry in seats-votes relationships that might be indicative of partisan gerrymandering: (1) \textit{partisan bias} \citep{Tufte1973}, (2) the \textit{mean-minus-median} gap \citep{Mcdonald_Best_2015_ELJ} (3) the \textit{efficiency gap} \citep{Stephanopoulos2014_UofChicagoLaw}, and most recently (4) \textit{declination} \citep{Warrington2018}.
 Each of these measures can be directly calculated for the 2011 map based on district election outcomes in 2012, 2014, and 2016.
@@
     \textit{Partisan Bias} is then calculated from the intercept of this regression using an exponential transformation (for details see \citet{Grofman1983}).
-Partisan bias is calculated from the Republican perspective, and \underline{negative} values indicates bias against Republicans.
 \par
@@
 Senate, PA Attorney General, PA Treasurer, and PA Auditor.
+%        =================================================================
+%        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
+% ------------------------------------------------------------------------
+        \footnote{We use the five-election composite to calculate all four plans.
+We aggregate voting district level (precincts) data for these five statewide elections and project them to the three maps in which we are comparing the 2011 enacted map.}
+% ------------------------------------------------------------------------
+%        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
+%        =================================================================
+%    
     Republicans won the statewide vote in two of them (President, U.S.
@@
     We are not claiming to predict vote shares in advance of an election.
-We also recognize that the composite of state-wide elections is intended to provide a ``normal vote'' baseline that must be ``corrected'' with incumbency effects if we wish to actually predict outcomes.
+We also recognize that the composite of state-wide elections is intended to provide a ``normal vote'' baseline that must be ``corrected'' with incumbency effects if we do wish to actually predict outcomes.
 \par
@@
 The congressional results for 2012-2016 and an average of the three, reported in Table \ref{tab:congsum}, are 0.11 for partisan bias, 0.16 for the efficiency gap, 0.06 for the value of the mean minus median gap, and 0.40 for \textit{declination}.
-The comparable results from Table \ref{tab:gerry} (see below) are 0.09, 0.24, 0.05, and 0.39.
-It is apparent how close these two sets of values are to one other even though one is based on actual elections with a mean Republican (two-party) vote share close to 49.1\% while the other is based on projected election results set to yield a 50-50 two-party vote distribution.
-If instead we calculate these four metrics using the composite five election statewide vote in these elections over the period 2012-2016, a Democratic vote share of 49.1\% we get virtually identical values: 0.09, 0.24, 0.05, 0.36.
+The comparable results from Table \ref{tab:gerry} for the 2011 enacted map at 50\% vote-share are 0.09, 0.24, 0.05, and 0.39.
 %        =================================================================
@@
     The findings from Table \ref{tab:gerry} are very clear; regardless of which metric we examine, the 2011 map is the most gerrymandered.
-%        =================================================================
-%        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
-% ------------------------------------------------------------------------
-        \footnote{We use the five-election composite to calculate all four plans.
-We aggregate voting district level (precincts) data for these five statewide elections and project them to the three maps in which we are comparing the 2011 enacted map.}
-% ------------------------------------------------------------------------
-%        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
-%        =================================================================
-%
 Not surprisingly, the nature of the bias is in a pro-Republican direction.
@@
     In 2018, even under a court-drawn plan, in a year when Democrats had won a clear majority of the aggregate two-party congressional vote, Democrats only succeeded in winning exactly 50\% of the seats.
-Trying to understand why evan plans not drawn by Republicans had some level of pro-Republican bias requires us to look in more detail at the partisan electoral geography of Pennsylvania.
+Trying to understand why evn plans not drawn by Republicans had some level of pro-Republican bias requires us to look in more detail at the partisan electoral geography of Pennsylvania.
 In Pennsylvania, Philadelphia is overwhelmingly Democratic in voting.

### 2019-06-19 R/GERRYfunctions.R
LaTeX files

@@ -5,7 +5,7 @@ cat(
 •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••\n")
 `Table` <- function(x, path=NULL, caption="", label="", footnote="", landscape=FALSE, out=NULL)
   {
-    x <- append(x,
+t <- 
 paste0(
 "\n% =====================================================================
 % ▀▄▀▄▀▄ T̟A̟B̟L̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ T̟A̟B̟L̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ T̟A̟B̟L̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ T̟A̟B̟L̟E̟
@@ -15,8 +15,10 @@ paste0("\\begin{center} \\textbf{", caption, "} \\end{center}"),
 paste0("\\input{", path, "}")),
 
 "\n \\begin{center}\\textbf{INSERT TABLE \\ref{", label, "} ABOUT HERE} \\end{center}
-% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••\n
-\n% =====================================================================
+% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••\n")
+
+    x <- append(x,
+paste0("\n% =====================================================================
 % ▀▄▀▄▀▄ T̟A̟B̟L̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ T̟A̟B̟L̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ T̟A̟B̟L̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ T̟A̟B̟L̟E̟
 % ---------------------------------------------------------------------",
 ifelse(landscape==T, "\n\\begin{landscape}", "")),
@@ -32,7 +34,7 @@ ifelse(landscape==T, "\n\\end{landscape}", ""),
 % ▀▄▀▄▀▄ E͎N͎D͎ T͎A͎B͎L͎E͎ ▄▀▄▀▄▀▀▄▀▄▀▄ E͎N͎D͎ T͎A͎B͎L͎E͎ ▄▀▄▀▄▀▀▄▀▄▀▄ E͎N͎D͎ T͎A͎B͎L͎E͎ ▄▀▄▀▄▀
 % ===================================================================== \n \n"),
 after = length(x))
[…]

### 2019-06-19 R/GIS.R

@@ -32,7 +32,7 @@ govwolf.compressed <- readOGR("shapefiles/topojson/shapefiles3.json", "govwolf")
 	proj4string(govwolf.compressed) <- CRS("+init=epsg:4269")
 
 enacted2011 <- spTransform(readOGR("shapefiles/enacted/enacted.shp"), CRS(projection))
-court <- spTransform(readOGR("shapefiles/remedial/remedial.shp"), CRS(projection))
+court <- spTransform(readOGR("shapefiles/court/remedial.shp"), CRS(projection))
 joint <- spTransform(readOGR("shapefiles/joint/joint.shp"), CRS(projection))
 govwolf <- spTransform(readOGR("shapefiles/govwolf/govwolf.shp"), CRS(projection))
 

### 2019-06-19 Latex/_maintext.tex
LATEX

@@
     Some states in which possible state law challenges could be made will be insulated from those challenges by factors such as the nature of institutional control over the redistricting process in the state, or political gridlock in the state courts.
-Nonetheless, if a state court does follow the Pennsylvania Supreme Court lead, we view the expected effects on limiting partisan gerrymandering to be potentially greater than we might expect under federal law, since they would involve the overturning of whole plans, not just individual districts -- although, in practice, a finding of overall unconstitutionality is likely to lead to changes only in the most egregiously gerrymandered districts and some of the districts adjacent to them.
+Nonetheless, if a state court does follow the Pennsylvania Supreme Court lead, we view the expected effects on limiting partisan gerrymandering to be potentially greater than we might expect under federal law, since they would involve the overturning of whole plans, not just individual districts.
 \par
@@
 % •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
-        \subsection*{Institutional Measures}
+        \subsection*{Geographical Indicators}
 % ------ ••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
@@
 \par
-    We now define these four measures of gerrymandering in which we will evaluate the four competing maps.
+    We now define these four measures of gerrymandering with which we will evaluate the four competing maps.
 %
@@
 \par
+\par
     In our view, none of these four measures standing alone can be taken to be proof of gerrymandering.
@@
 We will return to a discussion of the results later in this essay.
-Here we point out, however, all four indicators point clearly and overwhelmingly in the same direction vis-à-vis the 2011 Enacted plan -- namely that it is a blatant pro-Republican partisan gerrymander.
+Here we point out, however, all four indicators point clearly and overwhelmingly in the same direction vis-à-vis the 2011 Enacted plan, as Table \ref{tab:congsum} distinctly shows -- namely that it is a blatant pro-Republican partisan gerrymander.
 \par
@@
 Encouragingly, our five-election composite results is the same 13R-5D result that the 2016 congressional races delivered.
-Additionally, each district’s outcome is identical to the 2016 congressional district results, when we uniformly adjust in each district the projected vote based on the composite state-wide elections (the ``normal'' vote, without  congressional incumbency effects) by an amount equaling the difference between the statewide legislative vote and the the statewide congressional vote.
+Additionally, each district’s outcome is identical to the 2016 congressional district results when we uniformly adjust in each district the projected vote based on the composite state-wide elections (the ``normal'' vote, without  congressional incumbency effects) by an amount equaling the difference between the statewide legislative vote and the the statewide congressional vote.
 \par
@@
 Without doing so, the outcomes of the Court-ordered plan might have a wide range, with anywhere from a 5R to an 11R outcome possible because of the high number of competitive seats.
-However, when we take both factors into account by uniformly shifting the state-wide composite to reflect the 2018 statewide two-party congressional vote, and by adding in congressional incumbency advantage, which we have estimated in the 2012-2016 period to be on the order of magnitude of 4.3\% (analyses omitted for space reasons), we get a 9R-9D outcome as the most likely outcome in 2018 under the court-ordered map in our predictive model.
+However, when we take both factors into account by uniformly shifting the state-wide composite to reflect the 2018 statewide two-party congressional vote, and by adding in congressional incumbency advantage, which we have estimated in the 2012-2016 period to be on the order of magnitude of 4.3\% (analyses omitted for space reasons, \citet[see ][]{GelmenKing1990_inc_AJPS}), we get a 9R-9D outcome as the most likely outcome in 2018 under the court-ordered map in our predictive model.
 Thus, our confidence in the five election projection method is further buttressed when we applied this method to predict the results of the 2018 election under the court-ordered remedial map in the light of the actual statewide aggregate results in 2018, and taking incumbency effects into account.
@@
 Given the cumulative weight of all the evidence, the 2011 congressional map was clearly an egregious pro-Republican gerrymander.
-All four measures of potential gerrymandering are statistically significant for the 2011 map.
+All four measures of potential gerrymandering are statistically significant for the 2011 map using the composite simulations.
 In contrast, as we might expect, the court-ordered plan, prepared by a respected expert whose instructions were to satisfy good government criteria, was by a considerable margin the closest to a perfectly neutral plan in partisan terms according to all four measures.
@@
     Finally, we return to the plan offered by the Democratic governor, Governor Wolf.
-To our surprise, this plan appeared to have a pro-Republican tilt, nearly to the same extent as the joint legislative plan for three of the four metrics in Table \ref{tab:gerry}.
-This puzzling finding suggests that the Governor might have been under pressure from Democratic incumbents not to reduce their expected vote margins, or that he sincerely intended it to be a compromise that the Republican legislature might agree to.
+To our surprise, this plan appeared to have a pro-Republican tilt, though not to the same extent as either the 2011 enacted plan or the Republican alternative.
+This somewhat puzzling finding suggests that the Governor might have been under pressure from Democratic incumbents not to reduce their expected vote margins, or that he sincerely intended it to be a compromise that the Republican legislature might agree to.
 However, as shown in \ref{tab:gerry}, two of the apparent pro-Republican indicators for the plan are statistically insignificant.

### 2019-06-19 Response to Reviewers.txt
Create Response to Reviewers.txt

We apologize for the delay in returning this manuscript. The reviewer comments were invaluable, and we decided to makes extensive changes in the manuscript to be able to respond
 fully to them. That took considerably more time than we had originally thought. We believe that we have fully addressed all of the reviewer’s concerns and created a much better paper
 as a result. Since the reviewers were basically positive, we are optimistic that this much improved and very extensively rewritten version can be published in Political Geography.

There were several key changes in both substance and style: First, we have simplified all tables and figures to concentrate on only four plans rather than eight. 
The four we have picked are the ones most relevant to the case: the 2011 Enacted map, the Court ordered map used in 2018, and the two most important partisan remedial plans offered:
 one by Republican legislators, one by the Democratic governor. Second, we have completely redone all the tables, and used only one composite measure of state-wide election outcomes 
rather than a plethora of measures. This was requested by the reviewers.  Furthermore, we have added robustness checks for the composite we have chosen. 
Third, we have added in additional measures of partisan asymmetry, INCLUDING THE EFFICIENCY GAP, as requested. 
Fourth we have added in lots of new language in the second half of the paper explaining 
much more carefully exactly what we have done and what we trying to do, and most importantly, why it might matter.

REVIEWER 1

Reviewer 1 had three main criticisms. 
	(1) we should either pick one metric for assessing gerrymandering or use multiple metrics and show that they gave the same results –- and, with respect to the latter approach, 
we picked an unnecessary fight by omitting calculations for the efficiency gap. 
	(2) We needed more justification for projecting outcomes and our methodology was to helter-skelter 
since we offered multiple models without a clear reason to prefer one to the other. 
Moreover, this reviewer suggested that 
	(3)  our pre-2018 projections for the 2018 congressional elections were erroneous in the light of the actual election results, which was a significantly more democratic year
 than anything earlier in the decade.

The first point was easy to deal with –- the reviewer was quite right -- We now provide calculations for four different measures of (statewide) asymmetry in votes
 to seats relationships 
as a way to measure gerrymandering, with the efficiency gap one of these four, and we have omitted our critique of the efficiency gap, simply presenting results. 
The other three metrics 
are partisan bias, the mean-median gap, and a new recently offered measure called declination. 
In addition, in a more clearly organized table, we provide comparisons of the four plans
 with respect to the degree to which EACH preserves county boundaries and offers compact districts (two measures; one perimeter based, one area-based) 
 We are pleased to report that all seven measures give exactly the same ordering of plans with respect to the claim that they are gerrymanders. 
The 2011 map is a clear gerrymander; the Republican proposed remedial plan can appropriately be called a “stealth gerrymander,” 
the 2018 court-ordered plan is a politically neutral map, and the Democratic Governor’s plan is hard to characterize but the one thing that can be said for sure
 is that it is not a pro-Democratic gerrymander (indeed, it has, if anything, a slight pro-Republican tilt to it).

The second point was very well taken, and caused us to do lots more work in evaluating different projection models involving state-wide elections 
to find the best model that best fits the observed 2012-16 congressional data for the 2011 map and also works for the court-ordered map in 2018. 
That has taken us a while -- settling on a composite involving five statewide elections, with outcomes split 3-2 between the parties. We also realized that we 
first needed to clarify a really important methodological issue that was very poorly written up in the previous draft, and that led the reviewer to 
mistake what we were trying to do. We failed to make clear that there is a critical difference between the methods used to assess the degree to which
 a plan is a partisan gerrymander over a wide range of hypothetical election outcomes that are feasible given the historical level of inter-election shift,
 on the one hand, and predicting specific outcomes at the congressional district level in future election before those elections take place, on the other. 
 We have never tried to predict specific district outcomes in advance of an election. Rather what we want to do is show feasible ranges given empirically 
supported inter-election swings.  When this range of outcomes is sufficiently skewed in favor of one party, then we have evidence of gerrymandering.

One source of the confusion is that both assessment of asymmetry and attempts to project outcomes make use of the same data: actual results of congressional elections
 that have already taken place and/or the normal statewide vote from statewide elections that have already taken place. What we now do is to show that our normal vote
 data from statewide elections gives plausible estimates of partisan control/district outcomes for the 2011 Enacted map, since for that map we have 
actual election outcomes for three different election years to compare against. Using the actual statewide two-party vote for a composite of five 2016 
statewide elections (the best normal vote prediction model we have found), projected into congressional districts, we show that the state-wide election based composite 
“normal vote” can be used to reliably project outcomes in the 2011 plan, once we take into account the actual inter-election electoral tides over the 2012-16 period.
 That result gave us confidence that the same method could be used to assess the level of gerrymandering in a plan had not yet been implemented at the time of our 
original submission, the Court-ordered map that was (to be) used in 2018, as well as for two other maps that will never be used, the Republican legislative proposed
 remedial map and the Democratic Governor’s proposed remedial map.  THAT CONFIDENCE IS FURTHER CONFIRMED BY OUR SUCCESS IN USING 2016 STATEWIDE ELECTION RESULTS PROJECTED
 INTO 2018 DISTRICTS TO ASSESS CONGRESSIONAL DISTRICT OUTCOMES, BUT IN ORDER TO GET AN ACCURATE PREDICTION WE FURTHER NEED TO INCORPORATE THE  CONGRESIONAL INCUMBENCY EFFECTS 
WHICH WE HAVE SEPARATELY MODELED.  YOU NEED TO UNBOLD THIS

The new language addressing this point reads as follows.

	“We must distinguish our ability to assess asymmetry in expected seats-vote relationships based on the ``normal vote'' distribution of party support across 
districts -- and thus to assess the degree to which a plan is a gerrymander -- from our ability to predict actual election outcomes. The more competitive districts 
there are in a map, the more results depend strongly on electoral tides, and the greater the potential for winning vote margins in some districts to be small 
and determined by factors idiosyncratic to a district and by incumbency advantage. The 2011 map has 13 non-competitive districts (10+ percentage point margin) 
with a clear 9-4 Republican advantage, thus virtually guaranteeing continued Republican dominance regardless of aggregate two party vote share. The Court drawn map, comparatively, had only 8 non-competitive districts, with a balance of 3-R and 5-D. Which party will control the delegation under a highly competitive plan such as the Court-ordered plan depends almost entirely on the state-wide vote share and on incumbency advantage, along with idiosyncratic features of the competitive districts that might be small in magnitude but still matter in highly competitive seats."

With respect to the third point about projection for the court ordered map in 2018, once we incorporated the aggregate statewide share of the two-party congressional 
vote in 2018, and data on which districts had Dem or Rep incumbents in that election, our projection of the composite “normal vote” state-wide election data into the
 new 2018 congressional districts also turns out to work quite well. While there is a considerable range of feasible outcomes in the 2018 map based on the potential inter-election electoral shifts -— exactly as the reviewer pointed out -- once the election happens, we can check to seek to project results at the district level by adjusting the 2016 state-wide composite to the actual state-wide share of the two-party vote, and add in information about incumbency, since incumbency advantage also factors into election outcomes in highly competitive districts. When we combine all this information into one model -– information not available in advance --  we project the most likely congressional outcome in 2018 to be 9R-9D, but we also emphasize that the high number of highly competitive districts in the 2018 map made many other outcomes possible.

 The new language addressing this point reads as follows.

	In the congressional elections of 2018, the raw Republican two-party vote share in 2018 was 44.8\%, which represented a Democratic advantage, but not enough to overcome the twin barriers to a Democratic majority of over-concentration of Democratic support in urban areas and continued Republican incumbency advantage.

	% ================================================================= 
	% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
	% -----------------------------------------------------------------		
		Six Republicans and five Democrats ran for re-election after the new map was put in place. No incumbent lost. There were two Republican incumbents who won by less than 2 percentage points, along with one who won by less than 3 percentage points. There was just one seat won by Democrats in 2018 with a margin of 5 percentage points or less. Republicans therefore had four incumbents running and winning in competitive districts compared to just two for Democrats (10+ percentage point margin).
	% ----------------------------------------------------------------- 
	% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
	% =================================================================

    We are not claiming to predict vote shares in advance of an election. We also recognize that the composite of state-wide elections is intended to provide a 
``normal vote'' baseline that must be ``corrected'' with incumbency effects if we do wish to actually predict outcomes.

	When we look at the five election composite we need to (a) adjust it for the actual statewide congressional Democratic vote share in 2018, and 
(b) to take into account this differential incumbency advantage that still benefits Republicans. Without doing so, the outcomes of the Court-ordered plan 
have a wide range, with anywhere from a 5R to an 11R outcome possible because of the high number of competitive seats. However, when we take both factors into account by uniformly shifting the state-wide composite to reflect the 2018 statewide two-party congressional vote, and by adding in congressional incumbency advantage, which we have estimated in the 2012-2016 period to be on the order of magnitude of 4.3\% (analyses omitted for space reasons), we get a 9R-9D outcome as the most likely outcome in 2018 under the court-ordered map in our predictive model. Thus, our confidence in the five election projection method is further buttressed when we applied this method to predict the results of the 2018 election under the court-ordered remedial map in the light of the actual statewide aggregate results in 2018.”

Finally, the reviewer's interest in the comparison between a "fair" plan and a "neutral" plan led us to expand 
the disscussion, especially in light of the natural geographical features in Pennsylvania. It now is featured as it's own sub-section.

______________________________

REVIEWER 2
Reviewer 2 had one main criticism which overlapped that of Reviewer 1. Reviewer 2 notes that we picked an unnecessary fight by omitting calculations 
for the efficiency gap. Here, as noted in our response to Reviewer 1, we have fully complied with the reviewer’s suggestion putting that measure 
into the paper (without seeking to critique it, or any of the other measures, but simply reporting results), 
and it gives us exactly the same results for the four plans as the other asymmetry measures.

The second point made by Reviewer 2 is that there is inevitable arbitrariness in saying whether a plan is or is not a partisan gerrymander 
since we need to specify a threshold.

In our view it is self-evident as significant the difference between the 43 county splits in the 2011 map and the 17 county splits in the court-ordered map used in 2018,
 and judging as  lacking  IN clear significance the comparison of the 17 county splits in the court-ordered map versus the 19 county splits in the 
Republican and Democratic plans (see Table 1). On the other hand, though there is no red-line test yet adopted by the courts for defining a gerrymander, we have in this
 new submission added tests for  statistical significance for the asymmetry measures.  In particular, as reported in Table 3, all four gerrymandering metrics yield 
results that are statistically significant for both the 2011 map and the Republican proposed remedial map, but none are statistically significant for the court-ordered
 map. This material on statistical significance is new to this revised draft.

 Reviewers were correct in pointing out that our composite measure is somewhat of an arbitrary choice and our justification for using it should be derived from 
it's ability to accurately predict elections. Following this logic, we constructed simulations of our measure and then used these simulations to
 show how closely our measure relates to the actual 2012, 2014, and 2016 House elections, while also adding in data from 2018 now that the election has happened since
 our submission. This data shows that   our measure is remarkably accurate at the district level for all these elections once we take into account election -specific statewide results and also take incumbency effects into account.

The third point made by the reviewer has to do with typos.  We are grateful for the care with which this reviewer read the paper! We have, we believe, cleaned up 
all the mistakes the reviewer identified, though we have now have so much new text that, despite our best efforts, we may have inadvertently created new ones.

Responses to the reviewers concerns appropriately added some lengthy discussions to the text. We offset the increase in page count by eliminating non-essential data
 and streamlining the essay to more succinctly follow the paper's title and objective. We feel this more narrow   design helps to explicate our
 contribution as both 
a historical account of the most important partisan gerrymandering court case to date, but also to help structure a framework with which the partisan effects of 
districting plans can be measured and compared.

### 2019-06-19 Response to Reviewers.txt
Update Response to Reviewers.txt

@@
-We apologize for the delay in returning this manuscript. The reviewer comments were invaluable, and we decided to makes extensive changes in the manuscript to be able to respond
- fully to them. That took considerably more time than we had originally thought. We believe that we have fully addressed all of the reviewer’s concerns and created a much better paper
- as a result. Since the reviewers were basically positive, we are optimistic that this much improved and very extensively rewritten version can be published in Political Geography.
+We apologize for the delay in returning this manuscript.
+The reviewer comments were invaluable, and we decided to make extensive changes in the manuscript to be able to respond fully to them.
+That took considerably more time than we had originally thought.
+We believe that we have fully addressed all of the reviewer’s concerns and created a much better paper as a result.
+Since the reviewers were basically positive, we are optimistic that this much improved and very extensively rewritten version can be published in Political Geography.
 
 There were several key changes in both substance and style: First, we have simplified all tables and figures to concentrate on only four plans rather than eight.
-The four we have picked are the ones most relevant to the case: the 2011 Enacted map, the Court ordered map used in 2018, and the two most important partisan remedial plans offered:
- one by Republican legislators, one by the Democratic governor. Second, we have completely redone all the tables, and used only one composite measure of state-wide election outcomes 
-rather than a plethora of measures. This was requested by the reviewers.  Furthermore, we have added robustness checks for the composite we have chosen. 
-Third, we have added in additional measures of partisan asymmetry, INCLUDING THE EFFICIENCY GAP, as requested. 
-Fourth we have added in lots of new language in the second half of the paper explaining 
-much more carefully exactly what we have done and what we trying to do, and most importantly, why it might matter.
+The four we have picked are the ones most relevant to the case: the 2011 Enacted map, the Court ordered map used in 2018, and the two most important partisan remedial plans offered: one by Republican legislators, one by the Democratic governor.
+Second, we have completely redone all the tables, and used only one composite measure of state-wide election outcomes rather than a plethora of measures.
+This was requested by the reviewers.
+Furthermore, we have added robustness checks for the composite we have chosen.
+Third, we have added in additional measures of partisan asymmetry, including the efficiency gap, as requested.
+Fourth we have added in lots of new language in the second half of the paper explaining much more carefully exactly what we have done and what we trying to do, and most importantly, why it might matter.
 
@@
 Reviewer 1 had three main criticisms. 
-	(1) we should either pick one metric for assessing gerrymandering or use multiple metrics and show that they gave the same results –- and, with respect to the latter approach, 
-we picked an unnecessary fight by omitting calculations for the efficiency gap. 
-	(2) We needed more justification for projecting outcomes and our methodology was to helter-skelter 
-since we offered multiple models without a clear reason to prefer one to the other. 
+	(1) we should either pick one metric for assessing gerrymandering or use multiple metrics and show that they gave the same results –- and, with respect to the latter approach, we picked an unnecessary fight by omitting calculations for the efficiency gap.
+	(2) We needed more justification for projecting outcomes and our methodology was to helter-skelter since we offered multiple models without a clear reason to prefer one to the other. 
 Moreover, this reviewer suggested that 
-	(3)  our pre-2018 projections for the 2018 congressional elections were erroneous in the light of the actual election results, which was a significantly more democratic year
- than anything earlier in the decade.
+	(3) our pre-2018 projections for the 2018 congressional elections were erroneous in the light of the actual election results, which was a significantly more democratic year than anything earlier in the decade.
 
-The first point was easy to deal with –- the reviewer was quite right -- We now provide calculations for four different measures of (statewide) asymmetry in votes
- to seats relationships 
-as a way to measure gerrymandering, with the efficiency gap one of these four, and we have omitted our critique of the efficiency gap, simply presenting results. 
-The other three metrics 
-are partisan bias, the mean-median gap, and a new recently offered measure called declination. 
-In addition, in a more clearly organized table, we provide comparisons of the four plans
- with respect to the degree to which EACH preserves county boundaries and offers compact districts (two measures; one perimeter based, one area-based) 
+The first point was easy to deal with –- the reviewer was quite right -- We now provide calculations for four different measures of (statewide) asymmetry in votes to seats relationships as a way to measure gerrymandering, with the efficiency gap one of these four, and we have omitted our critique of the efficiency gap, simply presenting results.
+The other three metrics are partisan bias, the mean-median gap, and a new recently offered measure called declination.
+In addition, in a more clearly organized table, we provide comparisons of the four plans with respect to the degree to which each preserve county boundaries and offers compact districts (two measures; one perimeter based, one area-based).
 We are pleased to report that all seven measures give exactly the same ordering of plans with respect to the claim that they are gerrymanders.
-The 2011 map is a clear gerrymander; the Republican proposed remedial plan can appropriately be called a “stealth gerrymander,” 
-the 2018 court-ordered plan is a politically neutral map, and the Democratic Governor’s plan is hard to characterize but the one thing that can be said for sure
- is that it is not a pro-Democratic gerrymander (indeed, it has, if anything, a slight pro-Republican tilt to it).
+The 2011 map is a clear gerrymander; the Republican proposed remedial plan can appropriately be called a “stealth gerrymander,” the 2018 court-ordered plan is a politically neutral map, and the Democratic Governor’s plan is hard to characterize but the one thing that can be said for sure is that it is not a pro-Democratic gerrymander (indeed, it has, if anything, a slight pro-Republican tilt to it).
 
-The second point was very well taken, and caused us to do lots more work in evaluating different projection models involving state-wide elections 
-to find the best model that best fits the observed 2012-16 congressional data for the 2011 map and also works for the court-ordered map in 2018. 
-That has taken us a while -- settling on a composite involving five statewide elections, with outcomes split 3-2 between the parties. We also realized that we 
-first needed to clarify a really important methodological issue that was very poorly written up in the previous draft, and that led the reviewer to 
-mistake what we were trying to do. We failed to make clear that there is a critical difference between the methods used to assess the degree to which
- a plan is a partisan gerrymander over a wide range of hypothetical election outcomes that are feasible given the historical level of inter-election shift,
- on the one hand, and predicting specific outcomes at the congressional district level in future election before those elections take place, on the other. 
- We have never tried to predict specific district outcomes in advance of an election. Rather what we want to do is show feasible ranges given empirically 
-supported inter-election swings.  When this range of outcomes is sufficiently skewed in favor of one party, then we have evidence of gerrymandering.
+The second point was very well taken and caused us to do lots more work in evaluating different projection models involving state-wide elections to find the best model that best fits the observed 2012-16 congressional data for the 2011 map and also works for the court-ordered map in 2018.
+That has taken us a while -- settling on a composite involving five statewide elections, with outcomes split 3-2 between the parties.
+We also realized that we first needed to clarify a really important methodological issue that was very poorly written up in the previous draft, and that led the reviewer to mistake what we were trying to do.
+We failed to make clear that there is a critical difference between the methods used to assess the degree to which a plan is a partisan gerrymander over a wide range of hypothetical election outcomes that are feasible given the historical level of inter-election shift, on the one hand, and predicting specific outcomes at the congressional district level in future election before those elections take place, on the other.
+We have never tried to predict specific district outcomes in advance of an election.
+Rather what we want to do is show feasible ranges given empirically supported inter-election swings.
+When this range of outcomes is sufficiently skewed in favor of one party, then we have evidence of gerrymandering.
 
-One source of the confusion is that both assessment of asymmetry and attempts to project outcomes make use of the same data: actual results of congressional elections
- that have already taken place and/or the normal statewide vote from statewide elections that have already taken place. What we now do is to show that our normal vote
- data from statewide elections gives plausible estimates of partisan control/district outcomes for the 2011 Enacted map, since for that map we have 
-actual election outcomes for three different election years to compare against. Using the actual statewide two-party vote for a composite of five 2016 
-statewide elections (the best normal vote prediction model we have found), projected into congressional districts, we show that the state-wide election based composite 
-“normal vote” can be used to reliably project outcomes in the 2011 plan, once we take into account the actual inter-election electoral tides over the 2012-16 period.
- That result gave us confidence that the same method could be used to assess the level of gerrymandering in a plan had not yet been implemented at the time of our 
-original submission, the Court-ordered map that was (to be) used in 2018, as well as for two other maps that will never be used, the Republican legislative proposed
- remedial map and the Democratic Governor’s proposed remedial map.  THAT CONFIDENCE IS FURTHER CONFIRMED BY OUR SUCCESS IN USING 2016 STATEWIDE ELECTION RESULTS PROJECTED
- INTO 2018 DISTRICTS TO ASSESS CONGRESSIONAL DISTRICT OUTCOMES, BUT IN ORDER TO GET AN ACCURATE PREDICTION WE FURTHER NEED TO INCORPORATE THE  CONGRESIONAL INCUMBENCY EFFECTS 
-WHICH WE HAVE SEPARATELY MODELED.  YOU NEED TO UNBOLD THIS
+One source of the confusion is that both assessment of asymmetry and attempts to project outcomes make use of the same data: actual results of congressional elections that have already taken place and/or the normal statewide vote from statewide elections that have already taken place.
+What we now do is to show that our normal vote data from statewide elections gives plausible estimates of partisan control/district outcomes for the 2011 Enacted map, since for that map we have actual election outcomes for three different election years to compare against.
+Using the actual statewide two-party vote for a composite of five 2016 statewide elections (the best normal vote prediction model we have found), projected into congressional districts, we show that the state-wide election based composite “normal vote” can be used to reliably project outcomes in the 2011 plan, once we take into account the actual inter-election electoral tides over the 2012-16 period.
+That result gave us confidence that the same method could be used to assess the level of gerrymandering in a plan had not yet been implemented at the time of our original submission, the Court-ordered map that was (to be) used in 2018, as well as for two other maps that will never be used, the Republican legislative proposed remedial map and the Democratic Governor’s proposed remedial map.
+That confidence is further confirmed by our success in using 2016 statewide election results projected into 2018 districts to assess congressional district outcomes, but in order to get an accurate prediction we further need to incorporate the congresional incumbency effects which we have separately modeled.
 
@@
 
-	“We must distinguish our ability to assess asymmetry in expected seats-vote relationships based on the ``normal vote'' distribution of party support across 
-districts -- and thus to assess the degree to which a plan is a gerrymander -- from our ability to predict actual election outcomes. The more competitive districts 
-there are in a map, the more results depend strongly on electoral tides, and the greater the potential for winning vote margins in some districts to be small 
-and determined by factors idiosyncratic to a district and by incumbency advantage. The 2011 map has 13 non-competitive districts (10+ percentage point margin) 
-with a clear 9-4 Republican advantage, thus virtually guaranteeing continued Republican dominance regardless of aggregate two party vote share.
+	“We must distinguish our ability to assess asymmetry in expected seats-vote relationships based on the ``normal vote'' distribution of party support across districts -- and thus to assess the degree to which a plan is a gerrymander -- from our ability to predict actual election outcomes.
+The more competitive districts there are in a map, the more results depend strongly on electoral tides, and the greater the potential for winning vote margins in some districts to be small and determined by factors idiosyncratic to a district and by incumbency advantage.
+The 2011 map has 13 non-competitive districts (10+ percentage point margin) with a clear 9-4 Republican advantage, thus virtually guaranteeing continued Republican dominance regardless of aggregate two party vote share.
 The Court drawn map, comparatively, had only 8 non-competitive districts, with a balance of 3-R and 5-D.
@@
 
-With respect to the third point about projection for the court ordered map in 2018, once we incorporated the aggregate statewide share of the two-party congressional 
-vote in 2018, and data on which districts had Dem or Rep incumbents in that election, our projection of the composite “normal vote” state-wide election data into the
- new 2018 congressional districts also turns out to work quite well.
+With respect to the third point about projection for the court ordered map in 2018, once we incorporated the aggregate statewide share of the two-party congressional vote in 2018, and data on which districts had Dem or Rep incumbents in that election, our projection of the composite “normal vote” state-wide election data into the new 2018 congressional districts also turns out to work quite well.
 While there is a considerable range of feasible outcomes in the 2018 map based on the potential inter-election electoral shifts -— exactly as the reviewer pointed out -- once the election happens, we can check to seek to project results at the district level by adjusting the 2016 state-wide composite to the actual state-wide share of the two-party vote, and add in information about incumbency, since incumbency advantage also factors into election outcomes in highly competitive districts.
@@
 
-    We are not claiming to predict vote shares in advance of an election. We also recognize that the composite of state-wide elections is intended to provide a 
-``normal vote'' baseline that must be ``corrected'' with incumbency effects if we do wish to actually predict outcomes.
+	We are not claiming to predict vote shares in advance of an election.
+We also recognize that the composite of state-wide elections is intended to provide a ``normal vote'' baseline that must be ``corrected'' with incumbency effects if we do wish to actually predict outcomes.
 
-	When we look at the five election composite we need to (a) adjust it for the actual statewide congressional Democratic vote share in 2018, and 
-(b) to take into account this differential incumbency advantage that still benefits Republicans. Without doing so, the outcomes of the Court-ordered plan 
-have a wide range, with anywhere from a 5R to an 11R outcome possible because of the high number of competitive seats.
-However, when we take both factors into account by uniformly shifting the state-wide composite to reflect the 2018 statewide two-party congressional vote, and by adding in congressional incumbency advantage, which we have estimated in the 2012-2016 period to be on the order of magnitude of 4.3\% (analyses omitted for space reasons), we get a 9R-9D outcome as the most likely outcome in 2018 under the court-ordered map in our predictive model.
-Thus, our confidence in the five election projection method is further buttressed when we applied this method to predict the results of the 2018 election under the court-ordered remedial map in the light of the actual statewide aggregate results in 2018.”
+When we look at the five-election composite we need to (a) adjust it for the actual statewide congressional Democratic vote share in 2018, and (b) to take into account this differential incumbency advantage that still benefits Republicans.
+Without doing so, the outcomes of the Court-ordered plan have a wide range, with anywhere from a 5R to an 11R outcome possible because of the high number of competitive seats.
+However, when we take both factors into account by uniformly shifting the state-wide composite to reflect the 2018 statewide two-party congressional vote, and by adding in congressional incumbency advantage, which we have estimated in the 2012-2016 period to be on the order of magnitude of 4.3% (analyses omitted for space reasons), we get a 9R-9D outcome as the most likely outcome in 2018 under the court-ordered map in our predictive model.
+Thus, our confidence in the five-election projection method is further buttressed when we applied this method to predict the results of the 2018 election under the court-ordered remedial map in the light of the actual statewide aggregate results in 2018.”
 
-Finally, the reviewer's interest in the comparison between a "fair" plan and a "neutral" plan led us to expand 
-the disscussion, especially in light of the natural geographical features in Pennsylvania. It now is featured as it's own sub-section.
+Finally, the reviewer's interest in the comparison between a "fair" plan and a "neutral" plan led us to expand the discussion, especially in light of the natural geographical features in Pennsylvania.
+It now is featured as its own sub-section.
 
@@
 REVIEWER 2
-Reviewer 2 had one main criticism which overlapped that of Reviewer 1. Reviewer 2 notes that we picked an unnecessary fight by omitting calculations 
-for the efficiency gap. Here, as noted in our response to Reviewer 1, we have fully complied with the reviewer’s suggestion putting that measure 
-into the paper (without seeking to critique it, or any of the other measures, but simply reporting results), 
-and it gives us exactly the same results for the four plans as the other asymmetry measures.
 
-The second point made by Reviewer 2 is that there is inevitable arbitrariness in saying whether a plan is or is not a partisan gerrymander 
-since we need to specify a threshold.
+Reviewer 2 had one main criticism which overlapped that of Reviewer 1.
+Reviewer 2 notes that we picked an unnecessary fight by omitting calculations for the efficiency gap.
+Here, as noted in our response to Reviewer 1, we have fully complied with the reviewer’s suggestion putting that measure into the paper (without seeking to critique it, or any of the other measures, but simply reporting results), and it gives us exactly the same results for the four plans as the other asymmetry measures.
 
-In our view it is self-evident as significant the difference between the 43 county splits in the 2011 map and the 17 county splits in the court-ordered map used in 2018,
- and judging as  lacking  IN clear significance the comparison of the 17 county splits in the court-ordered map versus the 19 county splits in the 
-Republican and Democratic plans (see Table 1). On the other hand, though there is no red-line test yet adopted by the courts for defining a gerrymander, we have in this
- new submission added tests for  statistical significance for the asymmetry measures.  In particular, as reported in Table 3, all four gerrymandering metrics yield 
-results that are statistically significant for both the 2011 map and the Republican proposed remedial map, but none are statistically significant for the court-ordered
- map. This material on statistical significance is new to this revised draft.
+The second point made by Reviewer 2 is that there is inevitable arbitrariness in saying whether a plan is or is not a partisan gerrymander since we need to specify a threshold.
 
- Reviewers were correct in pointing out that our composite measure is somewhat of an arbitrary choice and our justification for using it should be derived from 
-it's ability to accurately predict elections. Following this logic, we constructed simulations of our measure and then used these simulations to
- show how closely our measure relates to the actual 2012, 2014, and 2016 House elections, while also adding in data from 2018 now that the election has happened since
- our submission.
-This data shows that   our measure is remarkably accurate at the district level for all these elections once we take into account election -specific statewide results and also take incumbency effects into account.
+In our view it is self-evident as significant the difference between the 43 county splits in the 2011 map and the 17 county splits in the court-ordered map used in 2018, and judging as lacking in clear significance the comparison of the 17 county splits in the court-ordered map versus the 19 county splits in the Republican and Democratic plans (see Table 1).
+On the other hand, though there is no bright-line test yet adopted by the courts for defining a gerrymander, we have in this new submission added tests for statistical significance for the asymmetry measures.
+In particular, as reported in Table 3, all four gerrymandering metrics yield results that are statistically significant for both the 2011 map and the Republican proposed remedial map, but none are statistically significant for the court-ordered map.
+This material on statistical significance is new to this revised draft.
 
-The third point made by the reviewer has to do with typos.  We are grateful for the care with which this reviewer read the paper! We have, we believe, cleaned up 
-all the mistakes the reviewer identified, though we have now have so much new text that, despite our best efforts, we may have inadvertently created new ones.
+Reviewers were correct in pointing out that our composite measure is somewhat of an arbitrary choice and our justification for using it should be derived from it's ability to accurately predict elections.
+Following this logic, we constructed simulations of our measure and then used these simulations to show how closely our measure relates to the actual 2012, 2014, and 2016 House elections, while also adding in data from 2018 now that the election has happened since our submission.
+This shows that our measure is remarkably accurate for all these elections once we take election specific statewide factors and our additional modeling work also shows the usefulness of a more probabilistic approach to estimating gerrymandering.
 
-Responses to the reviewers concerns appropriately added some lengthy discussions to the text. We offset the increase in page count by eliminating non-essential data
- and streamlining the essay to more succinctly follow the paper's title and objective. We feel this more narrow   design helps to explicate our
- contribution as both 
-a historical account of the most important partisan gerrymandering court case to date, but also to help structure a framework with which the partisan effects of 
-districting plans can be measured and compared.
+The third point made by the reviewer has to do with typos.
+We are grateful for the care with which this reviewer read the paper!
+We have, we believe, cleaned up all the mistake the reviewer identified, though we have now had so much new text that, despite our best efforts, we may have inadvertently created new ones.
+
+Responses to the reviewers concerns appropriately added some lengthy discussions to the text.
+We offset the increase in page count by eliminating non-essential data and streamlining the essay to more succinctly follow the paper's title and objective.
+We feel this narrower design helps to explicate our contribution as both a historical account of the most important partisan gerrymandering court case to date, but also to help structure a framework with which the partisan effects of districting plans can be measured and compared.

### 2019-06-19 Latex/anon.tex
Updated Submitted RandR Files

@@
 % =================================================================
-\title{Tools for Identifying Partisan Gerrymandering (PA)}
+\title{Tools for Identifying Partisan Gerrymandering}
 \author{Identifying Information Removed}
@@
 \documentclass[12pt, a4paper, twoside]{article}
-\RequirePackage{cervasanon}
+\RequirePackage{anon}
 % _________________________________________________________________
@@
 \begin{document}
-    \authorone{Identifying Information Removed}
-        \institone{Identifying Information Removed}
-    \authortwo{}
+    \authorone{}
+        \institone{}
         \instittwo{}
     \titlerunning{Tools for Identifying Partisan Gerrymandering}
-    \subtitle{}
-    \status{Under Review}
+    \subtitle{With an Application to Congressional Districting in Pennsylvania}
+    \status{Revised}
     \data{REDACTED}
     \abstract{In League of Women Voters v.
-Commonwealth of Pennsylvania (2018) the Pennsylvania Supreme Court struck down as a “severe and durable” partisan gerrymander the congressional map drawn by Republicans in 2011 and used in elections from 2012-2016.
+Commonwealth of Pennsylvania (2018) the Pennsylvania Supreme Court struck down as a ``severe and durable” partisan gerrymander the congressional map drawn by Republicans in 2011 and used in elections from 2012-2016.
 It did so entirely on state law grounds after a three-judge federal court had rejected issuing a preliminary injunction against the plan.
@@
 In a split court, the Court map was endorsed only by judges with Democratic affiliations.
-Here we compare and contrast the 2011 and 2018 maps in terms of a variety of proposed metrics for detecting partisan gerrymandering.
-For further comparison purposes, we examine six proposed remedy plans.
-These include the remedy map proposed by the Republican legislators and that proposed by the Democratic governor, along with four other maps drawn conforming either entirely or substantially to what are often referred to as ``good government” standards, and with no election outcome or party registration data used in the creation of three of them.
-We argue that the 2011 map was a blatant and undisguised pro-Republican gerrymander, while the 2018 remedy map proposed by Republican legislators was a covert gerrymander (what we refer to as a stealth gerrymander); but that, as we would expect, the 2018 court drawn map can not be classified as a gerrymander.}
+Here we compare the 2011 and 2018 congressional maps in terms of a variety of proposed metrics for detecting partisan gerrymandering.
+We also examine the remedy map proposed by a group of Republican legislators and that proposed by the Democratic governor.
+We conclude that the 2011 map was a blatant and undisguised pro-Republican gerrymander.
+Moreover, the remedy map proposed by Republican legislators was a covert pro-Republican gerrymander (what we refer to as a ``stealth gerrymander”).
+The Democratic governor's proposed plan cannot be classified as a pro-Democratic gerrymander and indeed has, if anything, a slight pro-Republican tilt.
+The 2018 court-drawn remedial map, by all measures, was not a gerrymander.}
 
@@
 %
-\input{maintext.tex}
+\input{_maintext.tex}
 
-\singlespacing
-\iffalse
-\clearpage
-\input{Tables/table1.tex}
-\clearpage
-\input{Tables/table2.tex}
-\clearpage
-\input{Tables/table3.tex}
-\clearpage
-\input{Figures/figure1.tex}
-\clearpage
-\input{Figures/figure2.tex}
-\fi
-% - Bibliography
-\clearpage
     \nolinenumbers
@@
 	}
+
 \begin{multicols}{2}
-    \footnotesize{\bibliography{bib}}   % Bibliography linked to bib.bib
+    \footnotesize{\bibliography{library}}   % Bibliography linked to bib.bib
 \end{multicols}
@@
 
-%\section{\ref{app:A} -- Methodological Appendix}
-    %\linenumbers
-    %\doublespacing
-    %\vspace{.5in}
-        %\input{appendixA.tex}
-    %\clearpage
-        %\input{Tables/tablea1.tex}
-    %\clearpage
-        %\input{Figures/figureA1.tex}
-    %\clearpage
-
 \end{document}
